---
title: "rFIA_mill_distance"
author: "Alyssa N Jain"
date: "6/1/2025"
format: html
editor: visual
---

```{r setup, include=FALSE}
#| echo: false
library(rFIA)
library(ggplot2)
library(dplyr)
library(stats)
library(knitr)
library(tidyverse)
library(tidyr)
library(stringr)
library(readr)
library(sf)
library(multcomp)
library(mediation)
library(lme4)
library(broom.mixed)
library(ggeffects)
library(purrr)       # functional helpers
library(survival)    # Cox model, counting-process Surv(start,stop,event)
library(lubridate)   # date helpers if needed
library(lfe)
library(broom)
library(nnet)
```

## How does distance to mill affect FIA plot characteristics?

**Questions:** We aim to answer the following questions:

*Q1) Does distance from the nearest mill affect the probability of fire?*

*Q2) Does distance increase the probability of fire because there’s less management happening farther from mills?*

*Q3) Does tree mortality increase as distance from mills increases?*

We separate analyses by resource regions

**Code Setup**

FIA data and GRM

```{r}
#| echo: false
# read in FIA data
ca_data <- readFIA("/Volumes/Alyssa_Backup/CA_CSV") 
or_data <- readFIA("/Volumes/Alyssa_Backup/OR_CSV")
wa_data <- readFIA("/Volumes/Alyssa_Backup/WA_CSV")

west_data <- readFIA("/Volumes/Alyssa_Backup/WEST_CSV", states = c("OR", "WA", "CA"))

west_cond <- west_data$COND #cond table
west_plot <- west_data$PLOT #plot table
west_tree <- west_data$TREE #tree table

# get most recent
# most recent 
west_data <- clipFIA(west_data, mostRecent = TRUE)

# get grm by plot
west_grm_plot <- growMort(west_data, byPlot = TRUE, stateVar = "CARB_AG")

# join
west_grm_table <- west_grm_plot %>%
  # Join tables
  left_join(west_cond, by = 'PLT_CN') # The foreign key for the COND table is PLT_CN. There is always a match of the PLT_CN value to the CN value in the PLOT table.

# make variables factors
west_grm_table$OWNGRPCD <- as.factor(west_grm_table$OWNGRPCD)
west_grm_table$FORTYPCD <- as.factor(west_grm_table$FORTYPCD)
west_grm_table$TRTCD1 <- as.factor(west_grm_table$TRTCD1)
west_grm_table$SITECLCD <- as.factor(west_grm_table$SITECLCD)
west_grm_table$PHYSCLCD <- as.factor(west_grm_table$PHYSCLCD)
west_grm_table$RESERVCD <- as.factor(west_grm_table$RESERVCD)
west_grm_table$YEAR <- as.factor(west_grm_table$YEAR)
west_grm_table$DSTRBCD1 <- as.factor(west_grm_table$DSTRBCD1)
west_grm_table$DSTRBCD2 <- as.factor(west_grm_table$DSTRBCD2)
west_grm_table$DSTRBCD3 <- as.factor(west_grm_table$DSTRBCD3)
```

Mill Travel Time

```{r}
#| echo: false
# read in mill travel time data
min_travel_time <- read_csv("min_travel_time.csv")

# filter by West
ca_min_travel_time <- min_travel_time[min_travel_time$STATE %in% c("OR", "WA", "CA"),] #this leaves some plots with FIPS code 41 (OR), even though STATE = CA
ca_min_travel_time <- min_travel_time[min_travel_time$STATECD %in% c("41", "53", "6"),]

# the last 5 digits of pltID in ca_grm_table are PLOT IDs from ca_min_travel time

# make new column in ca_grm_with last 5 digits of pltID
ca_grm_table1 <- west_grm_table

ca_grm_table1$PLOT_CD <- substring(west_grm_table$pltID, nchar(west_grm_table$pltID) - 4)

# now pivot so that time to chips and merchantable facilities are separate columns
ca_min_travel_time_clean <- ca_min_travel_time %>%
  dplyr::select(PLOT, BIOCD_DEF, ONE_WAY_HOURS) %>%
  group_by(PLOT, BIOCD_DEF) %>%
  # summarise(ONE_WAY_HOURS = min(ONE_WAY_HOURS), .groups = "drop") %>%
  pivot_wider(
    names_from = BIOCD_DEF,
    values_from = ONE_WAY_HOURS,
    names_prefix = "OWH_"
  )

# rename PLOT to match with ca_grm PLOT_CD
names(ca_min_travel_time_clean)[names(ca_min_travel_time_clean) == "PLOT"] <- "PLOT_CD"

# make PLOT_CD a character
ca_min_travel_time_clean$PLOT_CD <- as.character(ca_min_travel_time_clean$PLOT_CD)

# join with ca_grm
ca_grm_travel <- ca_grm_table1 %>%
  full_join(ca_min_travel_time_clean, by = 'PLOT_CD')
```

Fire Spatial Data

```{r}
#| echo: false
# read in spatial fire perimeters
fire_path <- "/Users/alyssajain/Downloads/mtbs_perimeter_data/mtbs_perims_DD.shp"
burns <- st_read(fire_path)

#check crs
st_crs(burns)

# filter for wildfire
wildfire_all <- burns[burns$Incid_Type == "Wildfire", ]

# wildfire_tbl <- as_tibble(wildfire_all)

wildfire_tbl <- wildfire_all %>%
  as_tibble() %>%
  slice_head(n = 10)



# read in State Shapefiles to clip boundaries
states <- st_read("/Users/alyssajain/Downloads/cb_2018_us_state_5m/cb_2018_us_state_5m.shp")
v
#check crs
st_crs(states) # matches

# filter states to CA, WA, OR
western_states <- states[states$STUSPS %in% c("CA", "WA", "OR"), ]
ca <- states[states$STUSPS %in% c( "WA", "OR"), ]

# Replace invalid geometries with valid ones
wildfire_all <- st_make_valid(wildfire_all)
western_states <- st_make_valid(western_states)
ca <- st_make_valid(ca)

# clip burns to CA
CA_wildfire <- st_intersection(wildfire_all, western_states)

wildfire <- CA_wildfire


# Convert FIA plot data to an sf object 
  # first remove NAs
  ca_plot_clean <- west_plot %>%
  filter(!is.na(LON) & !is.na(LAT))
  
  #then use coords to convert to sf
  fia_sf <- st_as_sf(ca_plot_clean, coords = c("LON", "LAT"), crs = 4326)
  
# # read in CA resource regions
# resource_reg <- read_sf("CAResourceAreas/CAResourceAreas.shp")
#   head(resource_reg)
#     # info: 6 resource areas (ResourceAr)
#   # match crs to fia
#   resource_reg <- st_transform(resource_reg, st_crs(fia_sf))
#   
#   # Dissolve multipolygons into single polygons by region name
#   resource_reg_clean <- resource_reg %>%
#     group_by(ResourceAr) %>%
#     summarise(geometry = st_union(geometry), .groups = "drop") %>%
#     st_as_sf()
#   
#   # join resource reg to fia data
#   fia_with_region <- st_join(fia_sf, resource_reg_clean, left = TRUE)

fia_with_region <- fia_sf
  
# Reproject wildfire data to match FIA CRS
wildfire <- st_transform(wildfire, st_crs(fia_sf))

# This returns a list: one element per plot, containing indices of intersecting fire polygons
wildfire_intersections <- st_intersects(fia_with_region, wildfire, sparse = TRUE)

# Check if each plot has any intersections (i.e., non-empty list element)
fia_with_region$wildfire_intersect <- lengths(wildfire_intersections) > 0

# Convert to 1/0 instead of TRUE/FALSE 
fia_with_region$wildfire_intersect <- as.integer(fia_with_region$wildfire_intersect)

# now we will join to ca_grm_travel
  # First create a plot ID column to match ca_grm_travel$pltID. This is UNITCD_STATECD_COUNTYCD_PLOT.
    fia_with_region$pltID <- paste(fia_with_region$UNITCD, fia_with_region$STATECD, fia_with_region$COUNTYCD, fia_with_region$PLOT, sep = "_")

  # Then extract just the plot ID, resource area, and fire_intersect column from the sf object
  wildfire_flag_df <- fia_with_region %>%
  sf::st_drop_geometry() %>%
  dplyr::select(pltID, wildfire_intersect) 

  # make pltID a character
  wildfire_flag_df$pltID <- as.character(wildfire_flag_df$pltID)
  
  # make sure each plot appears only once, in case a plot experiences multiple fires
  wildfire_flag_unique <- wildfire_flag_df %>%
  group_by(pltID) %>%
  summarise(
    wildfire_intersect = max(wildfire_intersect, na.rm = TRUE),
    # ResourceAr = first(ResourceAr),   # keep the resource region
    .groups = "drop"
  )
  
  
  # now join to ca_grm
  ca_grm_travel_wildfire <- ca_grm_travel %>%
  dplyr::left_join(wildfire_flag_unique, by = "pltID")
  
  # make fire_intersect a factor
  ca_grm_travel_wildfire$wildfire_intersect <- as.factor(ca_grm_travel_wildfire$wildfire_intersect)
  
  # make resource area a factor
  # ca_grm_travel_wildfire$ResourceAr <- as.factor(ca_grm_travel_wildfire$ResourceAr)  
 

```

```{r}
# Look at the data
summary(wildfire)
head(wildfire)
```

```{r}
# MTBS Burn Severity

# get files
sev_dir <- "/Users/alyssajain/FIA/CA_mtbs_unzipped"

sev_files <- list.files(sev_dir, full.names = TRUE, pattern = "\\.tif$")
length(sev_files)



```

**Findings:**

*Q1)* We ran a logistic regression to determine how distance from the nearest mill affects the probability of fire, using CA resource region as a random effect.

```{r}
  # Merchantable
   
  # no regions
   q1M <- glm(wildfire_intersect ~ OWH_Merchantable, data = ca_grm_travel_wildfire, family = "binomial")
    summary(q1M)
    
    # regions with fixed effect (effect controlling for resource area)
    q1M1_fixed <- glm(wildfire_intersect ~ OWH_Merchantable + ResourceAr,
            data = ca_grm_travel_wildfire, family = binomial)
    summary(q1M1_fixed)
    
    # regions as fixed effects with interaction term (effect by resource area)
    q1M1 <- glm(wildfire_intersect ~ OWH_Merchantable * ResourceAr,
            data = ca_grm_travel_wildfire, family = binomial)
    summary(q1M1)
    
    # Extract coefficients (log-odds)
    fixef_vals_q1M1 <- fixef(q1M1)
    
    # Tidy summary with SE, z, p-values
    fixed_summary_q1M1 <- broom.mixed::tidy(q1M1, effects = "fixed")
    
    # Add odds ratios for interpretation
    fixed_summary_q1M1 <- fixed_summary_q1M1 %>%
      mutate(
        odds_ratio = exp(estimate),
        prob_change = plogis(estimate)
      )
    
    print(fixed_summary_q1M1)
    view(fixed_summary_q1M1)
    
    
    # Marginal effects (predicted probabilities)
    predsq1M1 <- ggpredict(q1M1, terms = c("OWH_Merchantable", "ResourceAr"))

    # Plot
    ggplot(predsq1M1, aes(x = x, y = predicted, color = group)) +
      geom_line(size = 1.2) +
      geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.15, color = NA) +
      labs(
        x = "OWH Merchantable",
        y = "Predicted probability of wildfire",
        color = "Resource Area",
        fill  = "Resource Area",
        title = "Interaction between OWH_Merchantable and ResourceAr"
      ) +
      theme_minimal(base_size = 14)
    

        
  # Chips
    
    # no regions
    q1C <- glm(fire_intersect ~ OWH_Chips, data = ca_grm_travel_wildfire, family = "binomial")
    summary(q1C)
    
     # regions as fixed effects with interaction term
    q1C1 <- glm(wildfire_intersect ~ OWH_Chips * ResourceAr,
            data = ca_grm_travel_wildfire, family = binomial)
    summary(q1C1)
    
    # Extract coefficients (log-odds)
    fixef_vals_q1C1 <- fixef(q1C1)
    
    # Tidy summary with SE, z, p-values
    fixed_summary_q1C1 <- broom.mixed::tidy(q1C1, effects = "fixed")
    
    # Add odds ratios for interpretation
    fixed_summary_q1C1 <- fixed_summary_q1C1 %>%
      mutate(
        odds_ratio = exp(estimate),
        prob_change = plogis(estimate)
      )
    
    print(fixed_summary_q1C1)
    view(fixed_summary_q1C1)
    
    
    # Marginal effects (predicted probabilities)
    predsq1C1 <- ggpredict(q1C1, terms = c("OWH_Chips", "ResourceAr"))

    # Plot
    ggplot(predsq1C1, aes(x = x, y = predicted, color = group)) +
      geom_line(size = 1.2) +
      geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.15, color = NA) +
      labs(
        x = "OWH Chips",
        y = "Predicted probability of wildfire",
        color = "Resource Area",
        fill  = "Resource Area",
        title = "Interaction between OWH_Chips and ResourceAr"
      ) +
      theme_minimal(base_size = 14)
    
    
```

Using the fire perimeter data to determine fire occurrence, we find that for every additional hour in distance to a merchantable mill, the log odds of fire increase by 0.09028 (p = 0.000762). Similarly, for every additional hour in distance to a chips mill, the log odds of fire increase by 0.15161 (p = 6.79e-10). These results indicate that plots farther away from mills are significantly more likely to experience fire.

*Q2)* To see if decreased management is the mechanism increasing probability of fire farther from mills, we ran a logistic regression to see how distance from mills affects probability of treatment (which mostly refers to cutting, as well as a little preparation for planting).

```{r}
#| echo: false
# clean data for TRTCD = 10 (cutting) or 20 (Site preparation - Clearing, slash burning, chopping, disking, bedding, or other practices clearly intended to prepare a site for either natural or artificial regeneration.)
  ca_grm_travel_trtcd_grp <- ca_grm_travel_wildfire %>%
  mutate(
    TRTCD_CUT_PREP = if_else(
      as.numeric(as.character(TRTCD1)) %in% 10:20 |
      as.numeric(as.character(TRTCD2)) %in% 10:20 |
      as.numeric(as.character(TRTCD3)) %in% 10:20,
      1, 0
    )
  )
  

```

```{r}
# Merchantable
  q2aM <- glm(TRTCD_CUT_PREP ~ OWH_Merchantable, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2aM)
  
  # Chips
  q2aC <- glm(TRTCD_CUT_PREP ~ OWH_Chips, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2aC)
```

The log odds of treatment decrease by 0.56457 and 0.57778 for merchantable and chip mills, respectively (p \<2e-16). 

We then ran a logistic regression to see if there is more fire when there is less management, controlling for distance.

```{r}
# Merchantable
  q2bM <- glm(wildfire_intersect ~ OWH_Merchantable + TRTCD_CUT_PREP, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2bM)
  
  # Chips
  q2bC <- glm(wildfire_intersect ~ OWH_Chips + TRTCD_CUT_PREP, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2bC) 
```

For merchantable mills, the log odds of fire decrease by 0.25618 when there is treatment, compared to plots without treatment (p = 0.00519). Further, for every additional hour in distance to a merchantable mill, the log odds of fire increase by 0.08316 (p = 0.00202). This is down from 0.09028 when treatment is not considered, which means that treatment mediates the effect of distance on fire.

A causal mediation analysis reveals that 13% of the effect of distance from merchantable mills on the probability of fire is explained by reduced management (p = 0.01). We found a similar trend for Chip mills: the log odds of fire decrease by 0.25067 when there is treatment compared to no treatment (p = 0.00534). For every additional hour in distance to a chip mill, the log odds of fire increase by 0.14422 (p = 5.3e-09). This is down from 0.1516 when treatment is not considered, meaning that treatment mediates the effect of distance on fire. A causal mediation analysis reveals that reduced management explains 8.9% of the effect of distance from merchantable mills on the probability of fire (p= 0.002).

```{r}
# Mediation analysis: does management effect on fire interact with mill distance?
  
  # Merchantable
  med_outM <- mediate(q2aM, q2bM,
                     treat = "OWH_Merchantable",
                     mediator = "TRTCD_CUT_PREP",
                     boot = TRUE, sims = 1000)
  
  summary(med_outM)
  
  #Chips
  med_outC <- mediate(q2aC, q2bC,
                     treat = "OWH_Chips",
                     mediator = "TRTCD_CUT_PREP",
                     boot = TRUE, sims = 1000)
  
  summary(med_outC)
```

We are also going to run a Cox Proportional Hazards model. First we figured out which variables might be predictors of wildfire by running generalized linear models

```{r}
# test stocking density
  qsdi <- glm(wildfire_intersect ~ GSSTKCD, data = ca_grm_travel_wildfire, family = "binomial")
  summary(qsdi)
   # not significant predictor
  
# test forest type
  qftyp <- glm(wildfire_intersect ~ FORTYPCD, data = ca_grm_travel_wildfire, family = "binomial")
  summary(qftyp)
   # some types are  significant predictor
  
# test live basal area
  qbal <- glm(wildfire_intersect ~ BALIVE, data = ca_grm_travel_wildfire, family = "binomial")
  summary(qbal)
   # small effect but significant

# test physiographic class code
  qphys <- glm(wildfire_intersect ~ PHYSCLCD, data = ca_grm_travel_wildfire, family = "binomial")
  summary(qphys)
   # some xeric and mesic sites decrease wildfire risk, but that makes no sense, they should increase it...
  
# test disturbance code
  qdstrb <- glm(wildfire_intersect ~ DSTRBCD1, data = ca_grm_travel_wildfire, family = "binomial")
  summary(qdstrb)
    # negative for disease, weather, livestock, and positive for fire (obviously)
  
  
# make sure variables are factors if need be
ca_grm_travel_trtcd_grp$OWNGRPCD <- as.factor(ca_grm_travel_trtcd_grp$OWNGRPCD)
ca_grm_travel_trtcd_grp$FORTYPCD <- as.factor(ca_grm_travel_trtcd_grp$FORTYPCD)
ca_grm_travel_trtcd_grp$TRTCD1 <- as.factor(ca_grm_travel_trtcd_grp$TRTCD1)
ca_grm_travel_trtcd_grp$TRTCD2 <- as.factor(ca_grm_travel_trtcd_grp$TRTCD2)
ca_grm_travel_trtcd_grp$TRTCD3 <- as.factor(ca_grm_travel_trtcd_grp$TRTCD3)
ca_grm_travel_trtcd_grp$TRTYR1 <- as.factor(ca_grm_travel_trtcd_grp$TRTYR1)
ca_grm_travel_trtcd_grp$TRTYR2 <- as.factor(ca_grm_travel_trtcd_grp$TRTYR2)
ca_grm_travel_trtcd_grp$TRTYR3 <- as.factor(ca_grm_travel_trtcd_grp$TRTYR3)
ca_grm_travel_trtcd_grp$SITECLCD <- as.factor(ca_grm_travel_trtcd_grp$SITECLCD)
ca_grm_travel_trtcd_grp$PHYSCLCD <- as.factor(ca_grm_travel_trtcd_grp$PHYSCLCD)
ca_grm_travel_trtcd_grp$RESERVCD <- as.factor(ca_grm_travel_trtcd_grp$RESERVCD)
ca_grm_travel_trtcd_grp$YEAR <- as.factor(ca_grm_travel_trtcd_grp$YEAR)
ca_grm_travel_trtcd_grp$DSTRBCD1 <- as.factor(ca_grm_travel_trtcd_grp$DSTRBCD1)
ca_grm_travel_trtcd_grp$DSTRBCD2 <- as.factor(ca_grm_travel_trtcd_grp$DSTRBCD2)
ca_grm_travel_trtcd_grp$DSTRBCD3 <- as.factor(ca_grm_travel_trtcd_grp$DSTRBCD3)
ca_grm_travel_trtcd_grp$wildfire_intersect <- as.factor(ca_grm_travel_trtcd_grp$wildfire_intersect)
  

```

## Condition Level Cox

prep data

```{r}
# derive cutting year or the year cutting was observed
#------------------------------------------------------
# Define cutting-related treatment codes 
cut_codes <- c(10:21) 

# For convenience, pivot the 3 treatment slots into long format
cut_year_grm_all <- ca_grm_travel_trtcd_grp %>%
  mutate(across(starts_with("TRTYR"), ~ as.numeric(as.character(.x)))) %>%
  # calculate earliest cutting year (rowwise lets us work across columns for each plot)
  rowwise() %>%
  mutate(cut_year = min(
    c(TRTYR1[TRTCD1 %in% cut_codes],
      TRTYR2[TRTCD2 %in% cut_codes],
      TRTYR3[TRTCD3 %in% cut_codes]),
    na.rm = TRUE
  )) %>%
  # if no cutting treatment, set cut_year to NA instead of Inf
  mutate(cut_year = ifelse(is.infinite(cut_year), NA, cut_year)) %>%
  ungroup()

# get condition- plot ID that is pltID_CONDID
  cut_year_grm_all$plt_condID <- paste(cut_year_grm_all$pltID, cut_year_grm_all$CONDID, sep = "_")
```

merge with fire to get year

```{r}
# Convert df to sf for spatial intersection with fire polygons
# ---------------------------
# first collapse fia_with_region so only one row for each pltID
fia_region_unique <- fia_with_region %>%
  group_by(pltID) %>%
  slice(1) %>%        # keep just the first geometry for each pltID
  ungroup()

# get geometry from fia_with_region and join with the pltID
cut_year_grm_all_sf <- cut_year_grm_all %>%
  left_join(
    fia_region_unique %>% dplyr::select((pltID))
  ) %>%
  st_as_sf()


# mow intersect with wildfire polygons to get year
cut_year_grm_all_sf_fire <- st_join(
  cut_year_grm_all_sf,
  wildfire_clean %>% dplyr::select(YEAR_),
  join = st_intersects,
  left = TRUE
)
```

now run cox

```{r}
# Drop geometry from ca_grm
cut_year_grm_all_fire_nogeo <- st_drop_geometry(cut_year_grm_all_sf_fire)

# Collapse to earliest fire year per plot
fire_first <- cut_year_grm_all_fire_nogeo %>%
  group_by(pltID) %>%
  summarise(first_fire_year = min(YEAR_, na.rm = TRUE)) %>%
  mutate(first_fire_year = ifelse(is.infinite(first_fire_year), NA, first_fire_year))

# Merge back with GRM/treatment data
cox_df <- cut_year_grm_all_fire_nogeo %>%
  left_join(fire_first, by = "pltID")

# Define survival object
last_fire_year <- 2025
# Time = years since treatment (or since plot established) until first fire or censoring
cox_df <- cox_df %>%
  mutate(
    start_year = ifelse(!is.na(cut_year), cut_year, INVYR),   # start at cutting if cut, otherwise at inventory
    end_year   = ifelse(!is.na(first_fire_year), first_fire_year, last_fire_year), # event or censor
    time       = end_year - start_year,
    event      = ifelse(!is.na(first_fire_year), 1, 0)
  )

# 4. Cox model
cox_model <- coxph(
  Surv(time, event) ~ OWH_Chips + OWH_Merchantable + TRTCD_CUT_PREP + FORTYPCD + BALIVE + OWNGRPCD,
  data = cox_df
)

summary(cox_model)
```

## Plot Level Cox

Now we are going to collapse each entry to the plot level. Now, there might be multiple entries per plot, if a plot has multiple conditions. However, we don't want to look at fire at the condition level because the conditions are not independent, they may be part of the same plot so they would have the same fire outcome (since fire is mapped over plots)

```{r}
# collapse conditions to plot level, using the proportion of the plot that the condition covers to get the dominant or mean of each variables. 
plot_level_grm <- ca_grm_travel_trtcd_grp %>%
  group_by(pltID, INVYR) %>%
  summarise(
    # Weighted mean for all numeric vars
    across(where(is.numeric), 
           ~ weighted.mean(.x, w = CONDPROP_UNADJ, na.rm = TRUE), # keep in mind that now we can only use per acre variables, because the gross amounts should have been summed, not averaged. can fix code later to address that
           .names = "{.col}_wmean"),

    # For categoricals: take the value from the dominant condition
    across(where(is.character), 
           ~ .x[which.max(CONDPROP_UNADJ)], 
           .names = "{.col}_dom"),
    across(where(is.factor), 
           ~ .x[which.max(CONDPROP_UNADJ)], 
           .names = "{.col}_dom"),
  ) %>%
  ungroup()
```

Now we will prep the data.

```{r}
# derive cutting year or the year cutting was observed
#------------------------------------------------------
# Define cutting-related treatment codes 
cut_codes <- c(10:21) 

# For convenience, pivot the 3 treatment slots into long format
cut_year_df <- plot_level_grm %>%
  dplyr::select(pltID,
         TRTCD1_dom, TRTYR1_dom,
         TRTCD2_dom, TRTYR2_dom,
         TRTCD3_dom, TRTYR3_dom) %>%
  # make sure treatment years are numeric
  mutate(across(starts_with("TRTYR"), ~ as.numeric(as.character(.x)))) %>%
  # calculate earliest cutting year (rowwise lets us work across columns for each plot)
  rowwise() %>%
  mutate(cut_year = min(
    c(TRTYR1_dom[TRTCD1_dom %in% cut_codes],
      TRTYR2_dom[TRTCD2_dom %in% cut_codes],
      TRTYR3_dom[TRTCD3_dom %in% cut_codes]),
    na.rm = TRUE
  )) %>%
  # if no cutting treatment, set cut_year to NA instead of Inf
  mutate(cut_year = ifelse(is.infinite(cut_year), NA, cut_year)) %>%
  ungroup()
# Join back to plot-level dataset
plot_level_grm_cut <- plot_level_grm %>%
  left_join(cut_year_df, by = "pltID")
```

```{r}
# Convert plot_level_grm_cut to sf for spatial intersection with fire polygons
# ---------------------------
# first collapse fia_with_region so only one row for each pltID
fia_region_unique <- fia_with_region %>%
  group_by(pltID) %>%
  slice(1) %>%        # keep just the first geometry for each pltID
  ungroup()

# get geometry from fia_with_region and join with the pltID
plot_level_grm_cut_sf <- plot_level_grm_cut %>%
  left_join(
    fia_region_unique %>% dplyr::select((pltID))
  ) %>%
  st_as_sf()


# mow intersect with wildfire polygons to get year
plot_level_grm_cut_sf_fire <- st_join(
  plot_level_grm_cut_sf,
  wildfire_clean %>% dplyr::select(YEAR_),
  join = st_intersects,
  left = TRUE
)
```

Then we ran a Cox Proportional Hazards model to see how the covariates affected time until wildfire

```{r}
# Drop geometry from ca_grm
plot_level_grm_cut_fire_nogeo <- st_drop_geometry(plot_level_grm_cut_sf_fire)

# Collapse to earliest fire year per plot
fire_first <- plot_level_grm_cut_fire_nogeo %>%
  group_by(pltID) %>%
  summarise(first_fire_year = min(YEAR_, na.rm = TRUE)) %>%
  mutate(first_fire_year = ifelse(is.infinite(first_fire_year), NA, first_fire_year))

# Merge back with GRM/treatment data
cox_df <- plot_level_grm_cut_fire_nogeo %>%
  left_join(fire_first, by = "pltID")

# Define survival object
last_fire_year <- 2025
# Time = years since treatment (or since plot established) until first fire or censoring
cox_df <- cox_df %>%
  mutate(
    start_year = ifelse(!is.na(cut_year), cut_year, INVYR),   # start at cutting if cut, otherwise at inventory
    end_year   = ifelse(!is.na(first_fire_year), first_fire_year, last_fire_year), # event or censor
    time       = end_year - start_year,
    event      = ifelse(!is.na(first_fire_year), 1, 0)
  )

# 4. Cox model
cox_model <- coxph(
  Surv(time, event) ~ OWH_Chips_wmean+ OWH_Merchantable_wmean + TRTCD_CUT_PREP_wmean + FORTYPCD_dom + BALIVE_wmean + OWNGRPCD_dom,
  data = cox_df
)

summary(cox_model)
```

Let's check against log reg

```{r}
# all vars
glm_fin <- glm(wildfire_intersect ~ OWH_Chips + OWH_Merchantable + TRTCD_CUT_PREP + BALIVE + FORTYPCD + OWNGRPCD, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(glm_fin)
    #                    Estimate Std. Error z value Pr(>|z|)    
# (Intercept)      -7.008e-01  4.334e-01  -1.617 0.105876    
# OWH_Chips         8.532e-02  7.513e-02   1.136 0.256142    
# OWH_Merchantable  1.020e-01  8.449e-02   1.207 0.227347    
# TRTCD_CUT_PREP   -3.028e-01  1.059e-01  -2.859 0.004243 ** 
  
# is cutting because of OWH?
glm_cut_owh <- glm(TRTCD_CUT_PREP ~ OWH_Chips + OWH_Merchantable, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(glm_cut_owh)

```

look at Kaplan - Meier curves to see how travel time relates to time to fire

```{r}
# Load libraries
library(survival)
library(survminer)
library(dplyr)

# --- 1. Bin OWH into categories (0–1 hr, 1–2 hrs, …, 4–5 hrs) ---
cox_df <- cox_df %>%
  mutate(
    OWH_merch_bin = cut(
      OWH_Merchantable_wmean, breaks = seq(0, 5, by = 1), 
      labels = c("0–1 hr", "1–2 hrs", "2–3 hrs", "3–4 hrs", "4–5 hrs"), 
      right = TRUE, include.lowest = TRUE
    )
  )

# --- 2. Define survival object ---
# time = follow-up length (years until fire or censoring)
# event = 1 if fire occurred, 0 if censored
surv_obj <- Surv(time = cox_df$time, event = cox_df$event)

# --- 3. Kaplan–Meier fit stratified by OWH_bin ---
fit <- survfit(surv_obj ~ OWH_merch_bin, data = cox_df)

# --- 4. Plot with survminer ---
ggsurvplot(
  fit,
  data = cox_df,
  pval = TRUE,              # log-rank test p-value
  risk.table = TRUE,        # show number at risk below plot
  conf.int = TRUE,          # show confidence bands
  xlab = "Years since treatment/inventory",
  ylab = "Survival probability (no fire)",
  legend.title = "Time to mill (OWH)",
  legend.labs = levels(cox_df$OWH_merch_bin),
  ggtheme = theme_minimal()
)

```

## Fixed Effects Linear Model

Now we will run a fixed effects linear model

```{r}
felm_test2 <- felm(
  wildfire_intersect ~  BALIVE + OWNGRPCD + TRTCD_CUT_PREP | ResourceAr + YEAR + FORTYPCD| 0 | pltID,
  data = ca_grm_travel_trtcd_grp
)

summary(felm_test2)


felm_test <- felm(
  wildfire_intersect ~  BALIVE + OWNGRPCD + OWH_Chips | ResourceAr + YEAR + FORTYPCD| 0 | pltID,
  data = ca_grm_travel_trtcd_grp
)

summary(felm_test)


felm_trt <- felm(
  wildfire_intersect ~  BALIVE + OWNGRPCD | ResourceAr + YEAR + FORTYPCD| (TRTCD_CUT_PREP ~ OWH_Chips) | pltID,
  data = ca_grm_travel_trtcd_grp
)

summary(felm_trt)

felm_remv <- felm(
  wildfire_intersect ~  BALIVE + FORTYPCD + OWNGRPCD | ResourceAr + YEAR | (REMV_CARB_AG_ACRE ~ OWH_Chips) | pltID,
  data = ca_grm_travel_trtcd_grp
)

summary(felm_remv) # less of an impact

```

*Q3)* To determine if tree mortality increases with distance from mills, we ran a linear regression using mortality carbon per acre to represent mortality. For every additional hour in distance to a chip mill, the mortality carbon per acre increased by 80.28 tons (p = 0.00124). However, we found no significant effect on distance from merchantable mills (p = 0.0804). These results demonstrate that tree mortality significantly increases with distance to chip mills.

```{r}
# Merchantable
q3aM <- lm(MORT_CARB_AG_ACRE ~ OWH_Merchantable, data = ca_grm_travel_trtcd_grp)
summary(q3aM)

#Chips
q3aC <- lm(MORT_CARB_AG_ACRE ~ OWH_Chips, data = ca_grm_travel_trtcd_grp)
summary(q3aC)
```

Figures

**map of distance to mill by UNITCD**

BROKEN CODE – returns map with messed up UNITCD

```{r}
# Extract coordinates from fia sfinto separate columns 
coords_only <- fia_sf %>%
  mutate(
    LON = st_coordinates(.)[,1],
    LAT = st_coordinates(.)[,2]
  ) %>%
  dplyr::select(PLOT, LAT, LON, geometry)

# rename PLOT to match with ca_grm PLOT_CD
names(coords_only)[names(coords_only) == "PLOT"] <- "PLOT_CD"

#make PLOT_CD a character
coords_only$PLOT_CD <- as.character(coords_only$PLOT_CD)

# join coordinates from fia sf to plot/travel time table
ca_grm_travel_sf <- ca_grm_travel %>%
  left_join(coords_only, by = "PLOT_CD")

###### note: this is just for mercantable mills for now
# aggregate distance to mill by UNITCD
unit_avg <- ca_grm_travel_sf %>%
  st_drop_geometry() %>%
  group_by(UNITCD) %>%
  summarise(AVG_OWH_UNITCD = mean(OWH_Merchantable, na.rm = TRUE))

# join the county average back onto the original sf object
ca_grm_travel_unitcd_avg_sf <- ca_grm_travel_sf %>%
  left_join(unit_avg, by = "UNITCD")

# use coords to convert to sf
  ca_grm_travel_unitcd_avg_sf <- st_as_sf(ca_grm_travel_unitcd_avg_sf, coords = c("LON", "LAT"), crs = 4326)

# for plotting, get an sf that is just geometry and UNITCD_AVG
ca_grm_travel_unitcd_avg_ONLY <- ca_grm_travel_unitcd_avg_sf %>%
  dplyr::select(PLOT, AVG_OWH_UNITCD, geometry)
  
# plot
plot(ca_grm_travel_unitcd_avg_ONLY)

ggplot(ca_grm_travel_unitcd_avg_sf) +
  geom_sf(aes(fill = AVG_OWH_UNITCD)) +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  labs(fill = "Avg. Merchantable OWH by UNITCD")
```

## Binning

### Merchantable

get bins

```{r}

# get the range of travel times
summary(ca_grm_travel_trtcd_grp$OWH_Merchantable) # max: 4.9980

tmp <- ca_grm_travel_trtcd_grp %>%
  mutate(test = ifelse(
    OWH_Merchantable >= 0 & OWH_Merchantable < 0.5, 0, 
    ifelse(OWH_Merchantable >= 0.5 & OWH_Merchantable < 1, 1,
         ifelse(OWH_Merchantable >= 1 & OWH_Merchantable < 1.5, 2,
              ifelse(OWH_Merchantable >= 1.5 & OWH_Merchantable < 2, 3,
                     ifelse(OWH_Merchantable >= 2 & OWH_Merchantable < 2.5, 4,
                            ifelse(OWH_Merchantable >= 2.5 & OWH_Merchantable < 3, 5,
                                   ifelse(OWH_Merchantable >= 3 & OWH_Merchantable < 3.5, 6,
                                          ifelse(OWH_Merchantable >= 3.5 & OWH_Merchantable < 4, 7,
                                                 ifelse(OWH_Merchantable >= 4 & OWH_Merchantable < 4.5, 8,
                                                        ifelse(OWH_Merchantable >= 4.5 & OWH_Merchantable < 5, 9, 10)))))))))))
  

model <- felm(formula = wildfire_intersect ~ as.factor(test)  |    YEAR + FORTYPCD,     data = tmp)

summary(model)
                       
```

plot nonlinear relationship using felm

```{r}
{r}
# --- 1️⃣ Extract bin-level coefficients from felm model ---
coefs <- broom::tidy(model, conf.int = TRUE) %>%
  filter(grepl("as.factor\\(test\\)", term))

# --- 2️⃣ Clean up term names and assign bin midpoints ---
coefs <- coefs %>%
  mutate(
    bin = as.numeric(gsub("as\\.factor\\(test\\)", "", term)),
    hours_mid = bin * 0.5 + 0.25  # adjust if your bins differ
  )

# --- 3️⃣ Add baseline (reference bin = intercept) ---
intercept <- broom::tidy(model) %>%
  filter(term == "(Intercept)") %>%
  pull(estimate)

coefs <- coefs %>%
  add_row(
    term = "(Intercept)",
    estimate = intercept,
    conf.low = intercept,
    conf.high = intercept,
    bin = -1,
    hours_mid = 0
  )

# --- 4️⃣ Create Brookings-style plot ---
ggplot(coefs, aes(x = hours_mid, y = estimate)) +
  
  # Confidence ribbon (subtle, transparent)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              fill = "#74add1", alpha = 0.25) +
  
  # Smooth nonlinear curve (primary visual focus)
  geom_smooth(
    method = "loess", se = FALSE, span = 0.6,
    color = "#0570b0", size = 1.3, linetype = "solid"
  ) +
  
  # Axis and label formatting
  scale_x_continuous(
    name = "Hours to Merchantable Mill",
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  scale_y_continuous(
    name = "Wildfire Probability Coefficient",
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  
  # Title and subtitle (optional)
  labs(
    title = "Nonlinear Relationship between Hours to Merchantable Mill and Wildfire Probability",
    subtitle = "Estimated marginal effects with fixed effects = resource area and year"
  ) +
  
  # Clean, professional theme
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#08306b"),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray85", linetype = "dotted"),
    axis.text = element_text(color = "gray25"),
    plot.margin = margin(10, 15, 10, 15)
  )

```

Let's look for reasons for decline after 2.5 hours

```{r}
# Get # of plots in each bin
counts <- tmp %>% count(test)
print(counts)

```

Potential covariates of mill distance that could drive wildfire prob?

```{r}
# stocking density
   gsstk_lm <- lm(GSSTKCD ~ OWH_Merchantable , data = ca_grm_travel_wildfire)
    summary(gsstk_lm)
      # increases with distance to mill (goes from overstocked to nonstocked)
    
  # check felm, excluding fixed effects
    gsstk_felm <- felm(formula = GSSTKCD ~ as.factor(test)  |      0,     data = tmp)
    summary(gsstk_felm)
    
  # felm with fixed effects
    gsstk_felm <- felm(formula = GSSTKCD ~ as.factor(test)  |      ResourceAr + YEAR + FORTYPCD,     data = tmp)
    summary(gsstk_felm)
    # coefficients change with fixed effects, so stocking density doesn't independently covary with mill distance

# moisture
    
    # make it numeric (even though it is not ordinal)
    #-------------------------------------------------
        phys_lm <- lm(as.numeric(PHYSCLCD) ~ OWH_Merchantable , data = ca_grm_travel_trtcd_grp)
        summary(phys_lm)
          # increases with distance to mill 
        
      # check felm, excluding fixed effects
        phys_felm <- felm(formula = as.numeric(PHYSCLCD) ~ as.factor(test)  |      0,     data = tmp)
        summary(phys_felm)
        
      # felm with fixed effects
        phys_felm <- felm(formula = as.numeric(PHYSCLCD) ~ as.factor(test)  |      ResourceAr + YEAR + FORTYPCD,     data = tmp)
        summary(phys_felm)
        
    # check chi sq
    #-------------------------------------------------
      # make table
        phys_table = table(tmp$PHYSCLCD, tmp$test)
        print(phys_table)
        
        print(chisq.test(phys_table))
        
    # multinom log reg
      multinom_logreg <- multinom(PHYSCLCD ~ OWH_Merchantable + ResourceAr + YEAR + FORTYPCD,                                    data = ca_grm_travel_trtcd_grp)
      
      # Summary
      summary(multinom_logreg)
      
      # Get p-values (since nnet::multinom doesn't print them by default)
      z <- summary(multinom_logreg)$coefficients / summary(multinom_logreg)$standard.errors
      p <- 2 * (1 - pnorm(abs(z)))
      p
        

# Basal area living trees
    balive_lm <- lm(BALIVE ~ OWH_Merchantable , data = ca_grm_travel_trtcd_grp)
    summary(balive_lm)
     
    
  # check felm, excluding fixed effects
    balive_felm <- felm(formula = BALIVE ~ as.factor(test)  |      0,     data = tmp)
    summary(balive_felm)
    
  # felm with fixed effects
    balive_felm <- felm(formula = BALIVE ~ as.factor(test)  |      ResourceAr + YEAR + FORTYPCD,     data = tmp)
    summary(balive_felm)
```

### Chips

get bins

```{r}
# get the range of travel times
summary(ca_grm_travel_trtcd_grp$OWH_Chips) # max:  4.959  
summary(ca_grm_travel_trtcd_grp$OWH_Merchantable) # max: 4.9980

tmp <- ca_grm_travel_trtcd_grp %>%
  mutate(test = ifelse(
    OWH_Chips >= 0 & OWH_Chips < 0.5, 0, 
    ifelse(OWH_Chips >= 0.5 & OWH_Chips < 1, 1,
         ifelse(OWH_Chips >= 1 & OWH_Chips < 1.5, 2,
              ifelse(OWH_Chips >= 1.5 & OWH_Chips < 2, 3,
                     ifelse(OWH_Chips >= 2 & OWH_Chips < 2.5, 4,
                            ifelse(OWH_Chips >= 2.5 & OWH_Chips < 3, 5,
                                   ifelse(OWH_Chips >= 3 & OWH_Chips < 3.5, 6,
                                          ifelse(OWH_Chips >= 3.5 & OWH_Chips < 4, 7,
                                                 ifelse(OWH_Chips >= 4 & OWH_Chips < 4.5, 8,
                                                        ifelse(OWH_Chips >= 4.5 & OWH_Chips < 5, 9, 10)))))))))))
  

model <- felm(formula = wildfire_intersect ~ as.factor(test)  |      ResourceAr + YEAR + FORTYPCD,     data = tmp)

summary(model)
                       
```

plot nonlinear curve using felm

```{r}
# --- 1️⃣ Extract bin-level coefficients from felm model ---
coefs <- broom::tidy(model, conf.int = TRUE) %>%
  filter(grepl("as.factor\\(test\\)", term))

# --- 2️⃣ Clean up term names and assign bin midpoints ---
coefs <- coefs %>%
  mutate(
    bin = as.numeric(gsub("as\\.factor\\(test\\)", "", term)),
    hours_mid = bin * 0.5 + 0.25  # adjust if your bins differ
  )

# --- 3️⃣ Add baseline (reference bin = intercept) ---
intercept <- broom::tidy(model) %>%
  filter(term == "(Intercept)") %>%
  pull(estimate)

coefs <- coefs %>%
  add_row(
    term = "(Intercept)",
    estimate = intercept,
    conf.low = intercept,
    conf.high = intercept,
    bin = -1,
    hours_mid = 0
  )

# --- 4️⃣ Create Brookings-style plot ---
ggplot(coefs, aes(x = hours_mid, y = estimate)) +
  
  # Confidence ribbon (subtle, transparent)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              fill = "#74add1", alpha = 0.25) +
  
  # Smooth nonlinear curve (primary visual focus)
  geom_smooth(
    method = "loess", se = FALSE, span = 0.6,
    color = "#0570b0", size = 1.3, linetype = "solid"
  ) +
  
  # Axis and label formatting
  scale_x_continuous(
    name = "Hours to Chip Mill",
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  scale_y_continuous(
    name = "Wildfire Probability Coefficient",
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  
  # Title and subtitle (optional)
  labs(
    title = "Nonlinear Relationship between Hours to Chip Mill and Wildfire Probability",
    subtitle = "Estimated marginal effects with fixed effects = resource area and year"
  ) +
  
  # Clean, professional theme
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#08306b"),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray85", linetype = "dotted"),
    axis.text = element_text(color = "gray25"),
    plot.margin = margin(10, 15, 10, 15)
  )

```

**Histogram of Average Mill Distances**

```{r}
hist(ca_grm_travel$OWH_Merchantable)
hist(ca_grm_travel$OWH_Chips)

```

**Map of fire and distance to mills**

```{r}
# use coords to make sure data is sf
  ca_grm_travel_sf <- st_as_sf(ca_grm_travel_sf, coords = c("LON", "LAT"), crs = 4326)


# plot
ggplot() +
  geom_sf(data = fire_per, fill = "orange", color = "darkred", alpha = 0.3) +
  geom_sf(data = ca_grm_travel_sf, aes(color = OWH_Merchantable), size = 2) +
  scale_color_viridis_c(option = "C", name = "Distance to Chips Mill (hrs)") +
  theme_minimal() +
  labs(title = "Fire Perimeters and Plot Distances to Mill",
       subtitle = "Plots color-coded by distance to nearest chips mill")

```
