---
title: "rFIA_mill_distance"
author: "Alyssa N Jain"
date: "6/1/2025"
format: html
editor: visual
---

```{r setup, include=FALSE}
#| echo: false
library(rFIA)
library(ggplot2)
library(dplyr)
library(stats)
library(knitr)
library(tidyverse)
library(tidyr)
library(stringr)
library(readr)
library(sf)
library(multcomp)
library(mediation)
library(lme4)
library(broom.mixed)
library(ggeffects)
```

## How does distance to mill affect FIA plot characteristics?

**Questions:** We aim to answer the following questions:

*Q1) Does distance from the nearest mill affect the probability of fire?*

*Q2) Does distance increase the probability of fire because there’s less management happening farther from mills?*

*Q3) Does tree mortality increase as distance from mills increases?*

We separate analyses by CA resource regions

**Code Setup**

FIA data and GRM

```{r}
#| echo: false
# read in FIA data
ca_data <- readFIA("/Volumes/Alyssa_Backup/CA_CSV") 

ca_cond <- ca_data$COND #cond table
ca_plot <- ca_data$PLOT #plot table
ca_tree <- ca_data$TREE #tree table

# get grm by plot
grm_plot <- growMort(ca_data, byPlot = TRUE, stateVar = "CARB_AG")

# join
ca_grm_table <- grm_plot %>%
  # Join tables
  left_join(ca_cond, by = 'PLT_CN') # The foreign key for the COND table is PLT_CN. There is always a match of the PLT_CN value to the CN value in the PLOT table.

# make variables factors
ca_grm_table$OWNGRPCD <- as.factor(ca_grm_table$OWNGRPCD)
ca_grm_table$FORTYPCD <- as.factor(ca_grm_table$FORTYPCD)
ca_grm_table$TRTCD1 <- as.factor(ca_grm_table$TRTCD1)
ca_grm_table$SITECLCD <- as.factor(ca_grm_table$SITECLCD)
ca_grm_table$PHYSCLCD <- as.factor(ca_grm_table$PHYSCLCD)
ca_grm_table$RESERVCD <- as.factor(ca_grm_table$RESERVCD)
ca_grm_table$YEAR <- as.factor(ca_grm_table$YEAR)
ca_grm_table$DSTRBCD1 <- as.factor(ca_grm_table$DSTRBCD1)
ca_grm_table$DSTRBCD2 <- as.factor(ca_grm_table$DSTRBCD2)
ca_grm_table$DSTRBCD3 <- as.factor(ca_grm_table$DSTRBCD3)
```

Mill Travel Time

```{r}
#| echo: false
# read in mill travel time data
min_travel_time <- read_csv("min_travel_time.csv")

# filter by CA
ca_min_travel_time <- min_travel_time[min_travel_time$STATE == 'CA',] #this leaves some plots with FIPS code 41 (OR), even though STATE = CA
ca_min_travel_time <- min_travel_time[min_travel_time$STATECD == '6',]

# the last 5 digits of pltID in ca_grm_table are PLOT IDs from ca_min_travel time

# make new column un ca_grm_with last 5 digits of pltID
ca_grm_table1 <- ca_grm_table

ca_grm_table1$PLOT_CD <- substring(ca_grm_table$pltID, nchar(ca_grm_table$pltID) - 4)

# now pivot so that time to chips and merchantable facilities are separate columns
ca_min_travel_time_clean <- ca_min_travel_time %>%
  dplyr::select(PLOT, BIOCD_DEF, ONE_WAY_HOURS) %>%
  group_by(PLOT, BIOCD_DEF) %>%
  # summarise(ONE_WAY_HOURS = min(ONE_WAY_HOURS), .groups = "drop") %>%
  pivot_wider(
    names_from = BIOCD_DEF,
    values_from = ONE_WAY_HOURS,
    names_prefix = "OWH_"
  )

# rename PLOT to match with ca_grm PLOT_CD
names(ca_min_travel_time_clean)[names(ca_min_travel_time_clean) == "PLOT"] <- "PLOT_CD"

# make PLOT_CD a character
ca_min_travel_time_clean$PLOT_CD <- as.character(ca_min_travel_time_clean$PLOT_CD)

# join with ca_grm
ca_grm_travel <- ca_grm_table1 %>%
  full_join(ca_min_travel_time_clean, by = 'PLOT_CD')
```

Fire Spatial Data

```{r}
#| echo: false
# read in spatial fire perimeters
fire_path <- "/Users/alyssajain/FIA/fire24_1.gdb"

#find layers in gdb
st_layers(fire_path)

# get wildfire only
wildfire <- st_read(fire_path, layer = "firep24_1") 

# Convert FIA plot data to an sf object 
  # first remove NAs
  ca_plot_clean <- ca_plot %>%
  filter(!is.na(LON) & !is.na(LAT))
  
  #then use coords to convert to sf
  fia_sf <- st_as_sf(ca_plot_clean, coords = c("LON", "LAT"), crs = 4326)
  
# read in CA resource regions
resource_reg <- read_sf("CAResourceAreas/CAResourceAreas.shp")
  head(resource_reg)
    # info: 6 resource areas (ResourceAr)
  # match crs to fia
  resource_reg <- st_transform(resource_reg, st_crs(fia_sf))
  
  # Dissolve multipolygons into single polygons by region name
  resource_reg_clean <- resource_reg %>%
    group_by(ResourceAr) %>%
    summarise(geometry = st_union(geometry), .groups = "drop") %>%
    st_as_sf()
  
  # join resource reg to fia data
  fia_with_region <- st_join(fia_sf, resource_reg_clean, left = TRUE)

# Reproject wildfire data to match FIA CRS
wildfire <- st_transform(wildfire, st_crs(fia_sf))

# get rid of multisurface
wildfire_clean <- wildfire %>%
  filter(st_geometry_type(st_geometry(.)) %in% c("POLYGON", "MULTIPOLYGON"))

# fix invalid geometries
wildfire_geom <- st_geometry(wildfire_clean)

invalid_idx <- which(!st_is_valid(wildfire_geom))

if(length(invalid_idx) > 0){
  fixed_geoms <- st_make_valid(wildfire_geom[invalid_idx])
  wildfire_geom[invalid_idx] <- fixed_geoms
  st_geometry(wildfire_clean) <- wildfire_geom
}

# Now test validity again
sum(!st_is_valid(wildfire_clean))  # Should be 0 or very low
  
# This returns a list: one element per plot, containing indices of intersecting fire polygons
wildfire_intersections <- st_intersects(fia_with_region, wildfire_clean, sparse = TRUE)

# Check if each plot has any intersections (i.e., non-empty list element)
fia_with_region$wildfire_intersect <- lengths(wildfire_intersections) > 0

# Convert to 1/0 instead of TRUE/FALSE 
fia_with_region$wildfire_intersect <- as.integer(fia_with_region$wildfire_intersect)

# now we will join to ca_grm_travel
  # First create a plot ID column to match ca_grm_travel$pltID. This is UNITCD_STATECD_COUNTYCD_PLOT.
    fia_with_region$pltID <- paste(fia_with_region$UNITCD, fia_with_region$STATECD, fia_with_region$COUNTYCD, fia_with_region$PLOT, sep = "_")

  # Then extract just the plot ID, resource area, and fire_intersect column from the sf object
  wildfire_flag_df <- fia_with_region %>%
  sf::st_drop_geometry() %>%
  dplyr::select(pltID, wildfire_intersect, ResourceAr) 

  # make pltID a character
  wildfire_flag_df$pltID <- as.character(wildfire_flag_df$pltID)
  
  # make sure each plot appears only once, in case a plot experiences multiple fires
  wildfire_flag_unique <- wildfire_flag_df %>%
  group_by(pltID) %>%
  summarise(
    wildfire_intersect = max(wildfire_intersect, na.rm = TRUE),
    ResourceAr = first(ResourceAr),   # keep the resource region
    .groups = "drop"
  )
  
  
  # now join to ca_grm
  ca_grm_travel_wildfire <- ca_grm_travel %>%
  dplyr::left_join(wildfire_flag_unique, by = "pltID")
  
  # make fire_intersect a factor
  ca_grm_travel_wildfire$wildfire_intersect <- as.factor(ca_grm_travel_wildfire$wildfire_intersect)
  
  # make resource area a factor
  ca_grm_travel_wildfire$ResourceAr <- as.factor(ca_grm_travel_wildfire$ResourceAr)  
 


```

**Findings:**

*Q1)* We ran a logistic regression to determine how distance from the nearest mill affects the probability of fire, using CA resource region as a random effect.

```{r}
  # Merchantable
   
  # no regions
   q1M <- glm(wildfire_intersect ~ OWH_Merchantable, data = ca_grm_travel_wildfire, family = "binomial")
    summary(q1M)
    
    # regions as fixed effects with interaction term
    q1M1 <- glm(wildfire_intersect ~ OWH_Merchantable * ResourceAr,
            data = ca_grm_travel_wildfire, family = binomial)
    summary(q1M1)
    
    # Extract coefficients (log-odds)
    fixef_vals_q1M1 <- fixef(q1M1)
    
    # Tidy summary with SE, z, p-values
    fixed_summary_q1M1 <- broom.mixed::tidy(q1M1, effects = "fixed")
    
    # Add odds ratios for interpretation
    fixed_summary_q1M1 <- fixed_summary_q1M1 %>%
      mutate(
        odds_ratio = exp(estimate),
        prob_change = plogis(estimate)
      )
    
    print(fixed_summary_q1M1)
    view(fixed_summary_q1M1)
    
    
    # Marginal effects (predicted probabilities)
    predsq1M1 <- ggpredict(q1M1, terms = c("OWH_Merchantable", "ResourceAr"))

    # Plot
    ggplot(predsq1M1, aes(x = x, y = predicted, color = group)) +
      geom_line(size = 1.2) +
      geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.15, color = NA) +
      labs(
        x = "OWH Merchantable",
        y = "Predicted probability of wildfire",
        color = "Resource Area",
        fill  = "Resource Area",
        title = "Interaction between OWH_Merchantable and ResourceAr"
      ) +
      theme_minimal(base_size = 14)
    

        
  # Chips
    
    # no regions
    q1C <- glm(fire_intersect ~ OWH_Chips, data = ca_grm_travel_fire, family = "binomial")
    summary(q1C)
    
     # regions as fixed effects with interaction term
    q1C1 <- glm(wildfire_intersect ~ OWH_Chips * ResourceAr,
            data = ca_grm_travel_wildfire, family = binomial)
    summary(q1C1)
    
    # Extract coefficients (log-odds)
    fixef_vals_q1C1 <- fixef(q1C1)
    
    # Tidy summary with SE, z, p-values
    fixed_summary_q1C1 <- broom.mixed::tidy(q1C1, effects = "fixed")
    
    # Add odds ratios for interpretation
    fixed_summary_q1C1 <- fixed_summary_q1C1 %>%
      mutate(
        odds_ratio = exp(estimate),
        prob_change = plogis(estimate)
      )
    
    print(fixed_summary_q1C1)
    view(fixed_summary_q1C1)
    
    
    # Marginal effects (predicted probabilities)
    predsq1C1 <- ggpredict(q1C1, terms = c("OWH_Chips", "ResourceAr"))

    # Plot
    ggplot(predsq1C1, aes(x = x, y = predicted, color = group)) +
      geom_line(size = 1.2) +
      geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.15, color = NA) +
      labs(
        x = "OWH Chips",
        y = "Predicted probability of wildfire",
        color = "Resource Area",
        fill  = "Resource Area",
        title = "Interaction between OWH_Chips and ResourceAr"
      ) +
      theme_minimal(base_size = 14)
    
    
```

Using the fire perimeter data to determine fire occurrence, we find that for every additional hour in distance to a merchantable mill, the log odds of fire increase by 0.09028 (p = 0.000762). Similarly, for every additional hour in distance to a chips mill, the log odds of fire increase by 0.15161 (p = 6.79e-10). These results indicate that plots farther away from mills are significantly more likely to experience fire.

*Q2)* To see if decreased management is the mechanism increasing probability of fire farther from mills, we ran a logistic regression to see how distance from mills affects probability of treatment (which mostly refers to cutting, as well as a little preparation for planting).

```{r}
#| echo: false
# clean data for TRTCD = 10 (cutting) or 20 (Site preparation - Clearing, slash burning, chopping, disking, bedding, or other practices clearly intended to prepare a site for either natural or artificial regeneration.)
  ca_grm_travel_trtcd_grp <- ca_grm_travel_fire %>%
  mutate(
    TRTCD_CUT_PREP = if_else(
      as.numeric(as.character(TRTCD1)) %in% 10:20 |
      as.numeric(as.character(TRTCD2)) %in% 10:20 |
      as.numeric(as.character(TRTCD3)) %in% 10:20,
      1, 0
    )
  )
```

```{r}
# Merchantable
  q2aM <- glm(TRTCD_CUT_PREP ~ OWH_Merchantable, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2aM)
  
  # Chips
  q2aC <- glm(TRTCD_CUT_PREP ~ OWH_Chips, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2aC)
```

The log odds of treatment decrease by 0.56457 and 0.57778 for merchantable and chip mills, respectively (p \<2e-16). 

We then ran a logistic regression to see if there is more fire when there is less management, controlling for distance.

```{r}
# Merchantable
  q2bM <- glm(fire_intersect ~ OWH_Merchantable + TRTCD_CUT_PREP, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2bM)
  
  # Chips
  q2bC <- glm(fire_intersect ~ OWH_Chips + TRTCD_CUT_PREP, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2bC) 
```

For merchantable mills, the log odds of fire decrease by 0.25618 when there is treatment, compared to plots without treatment (p = 0.00519). Further, for every additional hour in distance to a merchantable mill, the log odds of fire increase by 0.08316 (p = 0.00202). This is down from 0.09028 when treatment is not considered, which means that treatment mediates the effect of distance on fire.

A causal mediation analysis reveals that 13% of the effect of distance from merchantable mills on the probability of fire is explained by reduced management (p = 0.01). We found a similar trend for Chip mills: the log odds of fire decrease by 0.25067 when there is treatment compared to no treatment (p = 0.00534). For every additional hour in distance to a chip mill, the log odds of fire increase by 0.14422 (p = 5.3e-09). This is down from 0.1516 when treatment is not considered, meaning that treatment mediates the effect of distance on fire. A causal mediation analysis reveals that reduced management explains 8.9% of the effect of distance from merchantable mills on the probability of fire (p= 0.002).

```{r}
# Mediation analysis: does management effect on fire interact with mill distance?
  
  # Merchantable
  med_outM <- mediate(q2aM, q2bM,
                     treat = "OWH_Merchantable",
                     mediator = "TRTCD_CUT_PREP",
                     boot = TRUE, sims = 1000)
  
  summary(med_outM)
  
  #Chips
  med_outC <- mediate(q2aC, q2bC,
                     treat = "OWH_Chips",
                     mediator = "TRTCD_CUT_PREP",
                     boot = TRUE, sims = 1000)
  
  summary(med_outC)
```

We are also going to run a Cox Proportional Hazards model. First we figured out which variables might be predictors of wildfire by running generalized linear models

```{r}
# test stocking density
  qsdi <- glm(wildfire_intersect ~ GSSTKCD, data = ca_grm_travel_wildfire, family = "binomial")
  summary(qsdi)
   # not significant predictor
  
# test forest type
  qftyp <- glm(wildfire_intersect ~ FORTYPCD, data = ca_grm_travel_wildfire, family = "binomial")
  summary(qftyp)
   # some types are significant significant predictor
  
# test live basal area
  qbal <- glm(wildfire_intersect ~ BALIVE, data = ca_grm_travel_wildfire, family = "binomial")
  summary(qbal)
   # small effect but significant

# test physiographic class code
  qphys <- glm(wildfire_intersect ~ PHYSCLCD, data = ca_grm_travel_wildfire, family = "binomial")
  summary(qphys)
   # some xeric and mesic sites decrease wildfire risk, but that makes no sense, they should increase it...
  
# test disturbance code
  qdstrb <- glm(wildfire_intersect ~ DSTRBCD1, data = ca_grm_travel_wildfire, family = "binomial")
  summary(qdstrb)
  

```

*Q3)* To determine if tree mortality increases with distance from mills, we ran a linear regression using mortality carbon per acre to represent mortality. For every additional hour in distance to a chip mill, the mortality carbon per acre increased by 80.28 tons (p = 0.00124). However, we found no significant effect on distance from merchantable mills (p = 0.0804). These results demonstrate that tree mortality significantly increases with distance to chip mills.

```{r}
# Merchantable
q3aM <- lm(MORT_CARB_AG_ACRE ~ OWH_Merchantable, data = ca_grm_travel_trtcd_grp)
summary(q3aM)

#Chips
q3aC <- lm(MORT_CARB_AG_ACRE ~ OWH_Chips, data = ca_grm_travel_trtcd_grp)
summary(q3aC)
```

Figures

**map of distance to mill by UNITCD**

BROKEN CODE – returns map with messed up UNITCD

```{r}
# Extract coordinates from fia sfinto separate columns 
coords_only <- fia_sf %>%
  mutate(
    LON = st_coordinates(.)[,1],
    LAT = st_coordinates(.)[,2]
  ) %>%
  dplyr::select(PLOT, LAT, LON, geometry)

# rename PLOT to match with ca_grm PLOT_CD
names(coords_only)[names(coords_only) == "PLOT"] <- "PLOT_CD"

#make PLOT_CD a character
coords_only$PLOT_CD <- as.character(coords_only$PLOT_CD)

# join coordinates from fia sf to plot/travel time table
ca_grm_travel_sf <- ca_grm_travel %>%
  left_join(coords_only, by = "PLOT_CD")

###### note: this is just for mercantable mills for now
# aggregate distance to mill by UNITCD
unit_avg <- ca_grm_travel_sf %>%
  st_drop_geometry() %>%
  group_by(UNITCD) %>%
  summarise(AVG_OWH_UNITCD = mean(OWH_Merchantable, na.rm = TRUE))

# join the county average back onto the original sf object
ca_grm_travel_unitcd_avg_sf <- ca_grm_travel_sf %>%
  left_join(unit_avg, by = "UNITCD")

# use coords to convert to sf
  ca_grm_travel_unitcd_avg_sf <- st_as_sf(ca_grm_travel_unitcd_avg_sf, coords = c("LON", "LAT"), crs = 4326)

# for plotting, get an sf that is just geometry and UNITCD_AVG
ca_grm_travel_unitcd_avg_ONLY <- ca_grm_travel_unitcd_avg_sf %>%
  dplyr::select(PLOT, AVG_OWH_UNITCD, geometry)
  
# plot
plot(ca_grm_travel_unitcd_avg_ONLY)

ggplot(ca_grm_travel_unitcd_avg_sf) +
  geom_sf(aes(fill = AVG_OWH_UNITCD)) +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  labs(fill = "Avg. Merchantable OWH by UNITCD")
```

**Histogram of Average Mill Distances**

```{r}
hist(ca_grm_travel$OWH_Merchantable)
hist(ca_grm_travel$OWH_Chips)

```

**Map of fire and distance to mills**

```{r}
# use coords to make sure data is sf
  ca_grm_travel_sf <- st_as_sf(ca_grm_travel_sf, coords = c("LON", "LAT"), crs = 4326)


# plot
ggplot() +
  geom_sf(data = fire_per, fill = "orange", color = "darkred", alpha = 0.3) +
  geom_sf(data = ca_grm_travel_sf, aes(color = OWH_Merchantable), size = 2) +
  scale_color_viridis_c(option = "C", name = "Distance to Chips Mill (hrs)") +
  theme_minimal() +
  labs(title = "Fire Perimeters and Plot Distances to Mill",
       subtitle = "Plots color-coded by distance to nearest chips mill")

```
