---
title: "rFIA carbon"
author: "Alyssa"
date: "4/9/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rFIA)
library(ggplot2)
library(dplyr)
library(stats)
library(knitr)
library(tidyverse)
library(tidyr)
library(stringr)
library(readr)
library(sf)
library(multcomp)
library(mediation)
```

load data
```{r}
ca_data <- readFIA("/Volumes/Alyssa_Backup/CA_CSV") 
min_travel_time <- read_csv("min_travel_time.csv")
              
```

## Carbon Calculations

Carbon by pool

```{r}
carbon(db = ca_data, byPool = TRUE)
```
Carbon by ownership

```{r}
#create table of carbon by pool and ownership
ownership <- carbon(db = ca_data, byPool = TRUE, grpBy = OWNGRPCD)

#aggregate by ownership and pool
ownership_summary <- ownership %>%
  group_by(OWNGRPCD, POOL) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# factorize
ownership_summary$POOL <- factor(ownership_summary$POOL)
ownership_summary$OWNGRPCD <- as.factor(ownership_summary$OWNGRPCD)

# plot
ggplot(ownership_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = POOL)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )


```
# Carbon by management

regeneration
```{r}
# natural (0) or artificial (1) regeneration
stand_regeneration <- carbon(db = ca_data, byPool = FALSE, grpBy = c(STDORGCD, OWNGRPCD))
stand_regeneration

#aggregate by ownership and pool
stand_regeneration_summary <- stand_regeneration %>%
  group_by(OWNGRPCD, STDORGCD) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# factorize
stand_regeneration_summary$STDORGCD <- factor(stand_regeneration_summary$STDORGCD)
stand_regeneration_summary$OWNGRPCD <- as.factor(stand_regeneration_summary$OWNGRPCD)

# plot
ggplot(stand_regeneration_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = STDORGCD)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )

```
treatment
```{r}
# treatment code 1
treatment1 <- carbon(db = ca_data, grpBy = TRTCD1)
treatment1

#aggregate by trtcd
carbon_by_treatment <- aggregate(CARB_ACRE ~ TRTCD1, data = treatment1, sum)

# barplot
barplot(carbon_by_treatment$CARB_ACRE,
        names.arg = carbon_by_treatment$TRTCD1,
        main = "Carbon by Treatment",
        xlab = "Treatment Code",
        ylab = "Carbon per Acre",
        las = 2)  # makes x-axis labels vertical for readability

# treatment and ownership
treatment_own <- carbon(db = ca_data, grpBy = c(TRTCD1, OWNGRPCD))

#factorize
treatment_own$TRTCD1 <- as.factor(treatment_own$TRTCD1)
levels(treatment_own$TRTCD1) # get different factor levels
levels(treatment_own$TRTCD1) <- c("None", "Cutting", "Regeneration Prep", "Artificial Regen", "Natural Regen", "Other") # rename for readability

treatment_own$OWNGRPCD <- as.factor(treatment_own$OWNGRPCD)
levels(treatment_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")


#aggregate by treatment and ownership
treatment_own_summary <- treatment_own %>%
  group_by(OWNGRPCD, TRTCD1) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(treatment_own_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = TRTCD1)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )
 

#plot without SE
ggplot(treatment_own, aes(fill=TRTCD1, y=CARB_ACRE, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")

```
cutting
```{r}
# stumps (cutting) and ownership
cutting_own <- carbon(db = ca_data, grpBy = c(STUMP_CD_PNWRS, OWNGRPCD))

#factorize
cutting_own$STUMP_CD_PNWRS <- as.factor(cutting_own$STUMP_CD_PNWRS)
levels(cutting_own$STUMP_CD_PNWRS) # get different factor levels

cutting_own$OWNGRPCD <- as.factor(cutting_own$OWNGRPCD)
levels(cutting_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#aggregate by treatment and ownership
cutting_own_summary <- cutting_own %>%
  group_by(OWNGRPCD, STUMP_CD_PNWRS) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(cutting_own_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = STUMP_CD_PNWRS)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )


#plot without SE
ggplot(cutting_own, aes(fill=STUMP_CD_PNWRS, y=CARB_ACRE, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")
```

site productivity code
```{r}
#site productivity and ownership
prod_own <- carbon(db = ca_data, grpBy = c(SITECLCD, OWNGRPCD))


#factorize
prod_own$SITECLCD <- as.factor(prod_own$SITECLCD)
levels(prod_own$SITECLCD) # get different factor levels

prod_own$OWNGRPCD <- as.factor(prod_own$OWNGRPCD)
levels(prod_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#aggregate by productivity and ownership
prod_own_summary <- prod_own %>%
  group_by(OWNGRPCD, SITECLCD) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(prod_own_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = SITECLCD)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )


#plot without SE
ggplot(prod_own, aes(fill=SITECLCD, y=CARB_ACRE, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")
```
forest type
```{r}
# read in fortypcd and group csv to aggregate later
fortypgrp <- read.csv("FORTYPCD_GRPCD.csv")

#find most common forest types by acre
fortyp_area <- area(db = ca_data, grpBy = FORTYPCD)
agg = aggregate(fortyp_area,
by = list(fortyp_area$FORTYPCD),
FUN = mean) #aggregate years

fortyp_order <- agg %>%
  arrange(desc(AREA_TOTAL))

top_fortyp <- agg %>%
  arrange(desc(AREA_TOTAL)) %>%
  slice_head(n = 6) %>%
  pull(FORTYPCD) # get top 6 forest type cd
top_fortyp

#now filter carbon by top fortypcd when doing carbon ownership analysis
top_fortyp_own <- carbon(db = ca_data, grpBy = c(FORTYPCD, OWNGRPCD))
top_fortyp_own$FORTYPCD <- as.factor(top_fortyp_own$FORTYPCD)

#filter
top_fortyp_own <- top_fortyp_own %>%
  filter(FORTYPCD %in% top_fortyp)
# check
summary(top_fortyp_own$FORTYPCD)

top_fortyp_own$OWNGRPCD <- as.factor(top_fortyp_own$OWNGRPCD)
levels(top_fortyp_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#plot
ggplot(top_fortyp_own, aes(fill=FORTYPCD, y=CARB_ACRE, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")

# error bars
#aggregate by fortyp and ownership
top_fortyp_own_summary <- top_fortyp_own %>%
  group_by(OWNGRPCD, FORTYPCD) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(top_fortyp_own_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = FORTYPCD)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )
####################
# fortype by owner by area
fortyp_area_own <- area(db = ca_data, grpBy = c(FORTYPCD, OWNGRPCD))
fortyp_area_own$FORTYPCD <- as.factor(fortyp_area_own$FORTYPCD)

#filter
top_fortyp_area_own <- fortyp_area_own %>%
  filter(FORTYPCD %in% top_fortyp)
# check
summary(top_fortyp_area_own$FORTYPCD)

top_fortyp_area_own$OWNGRPCD <- as.factor(top_fortyp_area_own$OWNGRPCD)
levels(top_fortyp_area_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#plot
ggplot(top_fortyp_area_own, aes(fill=FORTYPCD, y=AREA_TOTAL, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")

ggplot(top_fortyp_area_own, aes(fill=FORTYPCD, y=PERC_AREA, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")

# error bars
#aggregate by fortyp and ownership
top_fortyp_area_own_summary <- top_fortyp_area_own %>%
  group_by(OWNGRPCD, FORTYPCD) %>%
  summarise(
    PERC_AREA = mean(PERC_AREA),
    PERC_AREA_SE = mean(PERC_AREA_SE),
    .groups = 'drop'
  )

# plot
ggplot(top_fortyp_area_own_summary, aes(x = OWNGRPCD, y = PERC_AREA, fill = FORTYPCD)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = PERC_AREA - PERC_AREA_SE, ymax = PERC_AREA + PERC_AREA_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )

###################

#carbon
fortyp_own <- carbon(db = ca_data, grpBy = c(FORTYPCD, OWNGRPCD))
fortyp_own
fortyp_own$FORTYPCD <- as.factor(fortyp_own$FORTYPCD)

levels(fortyp_own$FORTYPCD) # get different factor levels


fortyp_own$OWNGRPCD <- as.factor(fortyp_own$OWNGRPCD)
levels(fortyp_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability


#plot
ggplot(fortyp_own, aes(fill=FORTYPCD, y=CARB_ACRE, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")
```
```{r}
# disturbance
dstrb_own <- carbon(db = ca_data, grpBy = c(DSTRBCD1, OWNGRPCD))

# factorize
dstrb_own$DSTRBCD1 <- as.factor(dstrb_own$DSTRBCD1)
levels(dstrb_own$DSTRBCD1) # get different factor levels

# aggregate disturbance codes
dstrb_own_grp <- dstrb_own %>%
  mutate(
    DSTRBCD1 = case_when(
      DSTRBCD1 %in% 10:12 ~ "Insect",
      DSTRBCD1 %in% 20:22 ~ "Disease",
      DSTRBCD1 %in% 30:32 ~ "Fire",
      DSTRBCD1 %in% 40:46 ~ "Animal",
      DSTRBCD1 %in% 50:54 ~ "Weather",
      DSTRBCD1 == 60:62 ~ "Vegetation",
      DSTRBCD1 == 70:79 ~ "Unknown",
      DSTRBCD1 == 80    ~ "Human",
      DSTRBCD1 %in% 90:95 ~ "Geologic",
      TRUE                        ~ "Other"
    )
  )

# factor owners
dstrb_own_grp$OWNGRPCD <- as.factor(dstrb_own_grp$OWNGRPCD)
levels(dstrb_own_grp$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#aggregate by disturbance and ownership
dstrb_own_grp_summary <- dstrb_own_grp %>%
  group_by(OWNGRPCD, DSTRBCD1) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(dstrb_own_grp_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = DSTRBCD1)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )

```


available moisture
```{r}
#site productivity and ownership
moist_own <- carbon(db = ca_data, grpBy = c(PHYSCLCD, OWNGRPCD))

#factorize
moist_own$PHYSCLCD <- as.factor(moist_own$PHYSCLCD)
levels(moist_own$PHYSCLCD) # get different factor levels

moist_own$OWNGRPCD <- as.factor(moist_own$OWNGRPCD)
levels(moist_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#aggregate by moist and ownership
moist_own_summary <- moist_own %>%
  group_by(OWNGRPCD, PHYSCLCD) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(moist_own_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = PHYSCLCD)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )


#plot
ggplot(moist_own, aes(fill=PHYSCLCD, y=CARB_ACRE, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")
```
stocking
```{r}
#site productivity and ownership
stock_own <- carbon(db = ca_data, grpBy = c(GSSTKCD, OWNGRPCD))

#factorize
stock_own$GSSTKCD <- as.factor(stock_own$GSSTKCD)
levels(stock_own$GSSTKCD) # get different factor levels

stock_own$OWNGRPCD <- as.factor(stock_own$OWNGRPCD)
levels(stock_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability


#aggregate by stocking and ownership
stock_own_summary <- stock_own %>%
  group_by(OWNGRPCD, GSSTKCD) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(stock_own_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = GSSTKCD)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )

```
## growMort
```{r}
# get carbon flux by ownership
carb_grm_agg <- growMort(ca_data, stateVar = "CARB", grpBy = "OWNGRPCD")

#factorize
carb_grm_agg$OWNGRPCD <- as.factor(carb_grm_agg$OWNGRPCD)
levels(carb_grm_agg$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#aggregate by ownership
carb_grm_summary <- carb_grm_agg %>%
  group_by(OWNGRPCD) %>%
  summarise(
    RECR_CARB_ACRE = mean(RECR_CARB_ACRE),
    RECR_CARB_ACRE_SE = mean(RECR_CARB_ACRE_SE),
    MORT_CARB_ACRE = mean(MORT_CARB_ACRE),
    MORT_CARB_ACRE_SE = mean(MORT_CARB_ACRE_SE),
    REMV_CARB_ACRE = mean(REMV_CARB_ACRE),
    REMV_CARB_ACRE_SE = mean(REMV_CARB_ACRE_SE),
    GROW_CARB_ACRE = mean(GROW_CARB_ACRE),
    GROW_CARB_ACRE_SE = mean(GROW_CARB_ACRE_SE),
    CHNG_CARB_ACRE = mean(CHNG_CARB_ACRE),
    .groups = 'drop'
  )

# make mort and remv negative
carb_grm_summary$MORT_CARB_ACRE <- carb_grm_summary$MORT_CARB_ACRE * -1
carb_grm_summary$REMV_CARB_ACRE <- carb_grm_summary$REMV_CARB_ACRE * -1
## comparison

# 1. Pivot GRM values to long format
grm_vals <- carb_grm_summary %>%
  select(OWNGRPCD, RECR_CARB_ACRE, MORT_CARB_ACRE, REMV_CARB_ACRE, GROW_CARB_ACRE) %>%
  pivot_longer(
    cols = ends_with("CARB_ACRE"),
    names_to = "GRM",
    values_to = "CARB_ACRE"
  ) %>%
  mutate(GRM = str_remove(GRM, "_CARB_ACRE"))

# 2. Pivot SE values to long format
grm_ses <- carb_grm_summary %>%
  select(OWNGRPCD, RECR_CARB_ACRE_SE, MORT_CARB_ACRE_SE, REMV_CARB_ACRE_SE, GROW_CARB_ACRE_SE) %>%
  pivot_longer(
    cols = ends_with("_SE"),
    names_to = "GRM",
    values_to = "CARB_ACRE_SE"
  ) %>%
  mutate(GRM = str_remove(GRM, "_CARB_ACRE_SE"))

# 3. Join the values + SEs
grm_combined <- left_join(grm_vals, grm_ses, by = c("OWNGRPCD", "GRM"))


# 4. Plot
# stacked
  # Pivot CHNG values to long format
  chng_vals <- carb_grm_summary %>%
    select(OWNGRPCD, CHNG_CARB_ACRE) %>%
    pivot_longer(
      cols = ends_with("CARB_ACRE"),
      names_to = "GRM",
      values_to = "CARB_ACRE"
    ) %>%
    mutate(GRM = str_remove(GRM, "_CARB_ACRE"))
  
  #plot
  ggplot(grm_combined, aes(x = OWNGRPCD, y = CARB_ACRE, fill = GRM)) +
    geom_bar(stat = "identity", position = "stack") + 
    geom_point(
    data = chng_vals,
    aes(x = OWNGRPCD, y = CARB_ACRE),
    color = "red",
    size = 3
)


# staggered
ggplot(grm_combined, aes(x = OWNGRPCD, y = CARB_ACRE, fill = GRM)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.3,
    color = "black"
  ) +
  theme_minimal() +
  labs(
    x = "Ownership Group",
    y = "Carbon per Acre",
    fill = "GRM Component"
  )

```

## spatial ownership dynamics
```{r}
# load in taskforce regions
regions <- read_sf("/Volumes/Alyssa_Backup/rFIA_work/RRK_boundaries4/RRK_boundaries4.shp")

# get carbon for ownership by region
carbon_own_region <- carbon(db = ca_data, byPool = FALSE, polys = regions, grpBy = OWNGRPCD, totals = TRUE, returnSpatial = TRUE)

# plot
plot(carbon_own_region["CARB_ACRE"])
```

## Regression
# GRM
```{r}
ca_cond <- ca_data$COND #cond table
ca_plot <- ca_data$PLOT #plot table
ca_tree <- ca_data$TREE #tree table

# get grm by plot
grm_plot <- growMort(ca_data, byPlot = TRUE, stateVar = "CARB_AG")

# join
ca_grm_table <- grm_plot %>%
  # Join tables
  left_join(ca_cond, by = 'PLT_CN') # The foreign key for the COND table is PLT_CN. There is always a match of the PLT_CN value to the CN value in the PLOT table.

# # filter by top fortypcd
# top_fortyp_ca_grm_table <- ca_grm_table %>%
#   filter(FORTYPCD %in% top_fortyp)
# # check
# summary(top_fortyp_ca_grm_table$FORTYPCD)

# # aggregate disturbances
# # aggregate disturbance codes
# ca_grm_table_grp <- ca_grm_table %>%
#   mutate(
#     DSTRBCD1 = case_when(
#       DSTRBCD1 %in% 10:12 ~ "Insect",
#       DSTRBCD1 %in% 20:22 ~ "Disease",
#       DSTRBCD1 %in% 30:32 ~ "Fire",
#       DSTRBCD1 %in% 40:46 ~ "Animal",
#       DSTRBCD1 %in% 50:54 ~ "Weather",
#       DSTRBCD1 == 60:62 ~ "Vegetation",
#       DSTRBCD1 == 70:79 ~ "Unknown",
#       DSTRBCD1 == 80    ~ "Human",
#       DSTRBCD1 %in% 90:95 ~ "Geologic",
#       TRUE                        ~ "Other"
#     )
#   )


# make variables factors
ca_grm_table$OWNGRPCD <- as.factor(ca_grm_table$OWNGRPCD)
ca_grm_table$FORTYPCD <- as.factor(ca_grm_table$FORTYPCD)
ca_grm_table$TRTCD1 <- as.factor(ca_grm_table$TRTCD1)
ca_grm_table$SITECLCD <- as.factor(ca_grm_table$SITECLCD)
ca_grm_table$PHYSCLCD <- as.factor(ca_grm_table$PHYSCLCD)
ca_grm_table$RESERVCD <- as.factor(ca_grm_table$RESERVCD)
ca_grm_table$YEAR <- as.factor(ca_grm_table$YEAR)
ca_grm_table$DSTRBCD1 <- as.factor(ca_grm_table$DSTRBCD1)
ca_grm_table$DSTRBCD2 <- as.factor(ca_grm_table$DSTRBCD2)
ca_grm_table$DSTRBCD3 <- as.factor(ca_grm_table$DSTRBCD3)
```

remv estimates (lm4_remv is best so far)
```{r}
lm1_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD, data = ca_grm_table)
summary(lm1_remv) # Adjusted R-squared:  0.02081  

lm2_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD, data = ca_grm_table_grp)
summary(lm2_remv) # Adjusted R-squared:  0.07301 

lm3_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1, data = ca_grm_table_grp)
summary(lm3_remv) # Adjusted R-squared:  0.234

lm4_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD, data = ca_grm_table_grp)
summary(lm4_remv) # Adjusted R-squared:  0.2423

lm5_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD, data = ca_grm_table_grp)
summary(lm5_remv) # Adjusted R-squared:  0.2422

lm6_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD + DSTRBCD1, data = ca_grm_table_grp)
summary(lm6_remv) # Adjusted R-squared:  0.242

lmDSTRB_remv <- lm(REMV_CARB_AG_ACRE ~ DSTRBCD1 + DSTRBCD2 + DSTRBCD3, data = ca_grm_table_grp)
summary(lmDSTRB_remv) # Adjusted R-squared: 0.01728

lm1DSTRB_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + DSTRBCD1 + DSTRBCD2 + DSTRBCD3, data = ca_grm_table_grp)
summary(lm1DSTRB_remv) # Adjusted R-squared: 0.08971

```

mortality estimates
```{r}
lm1_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD, data = ca_grm_table_grp)
summary(lm1_mort) # Adjusted R-squared:  0.03153  

lm2_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD, data = ca_grm_table_grp)
summary(lm2_mort) # Adjusted R-squared:  0.09861 

lm3_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1, data = ca_grm_table_grp)
summary(lm3_mort) # Adjusted R-squared:  0.09875

lm4_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD, data = ca_grm_table_grp)
summary(lm4_mort) # Adjusted R-squared:  0.1289

lm5_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD, data = ca_grm_table_grp)
summary(lm5_mort) # Adjusted R-squared:  0.1346

lm6_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD + DSTRBCD1, data = ca_grm_table_grp)
summary(lm6_mort) # Adjusted R-squared:  0.2948

lm7_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD + DSTRBCD1 + YEAR, data = ca_grm_table)
summary(lm7_mort) # Adjusted R-squared:  0.3274

lmDSTRB_mort <- lm(MORT_CARB_AG_ACRE ~ DSTRBCD1 + DSTRBCD2 + DSTRBCD3, data = ca_grm_table_grp)
summary(lmDSTRB_mort) # Adjusted R-squared: 0.2692 

lm1DSTRB_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + DSTRBCD1 + DSTRBCD2 + DSTRBCD3, data = ca_grm_table_grp)
summary(lm1DSTRB_mort) # Adjusted R-squared: 0.2747 

lmyr_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + YEAR + FORTYPCD + DSTRBCD1 + DSTRBCD2 + DSTRBCD3, data = ca_grm_table)
summary(lmyr_mort) # Adjusted R-squared: 0.3101

lmyr_mort2 <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + YEAR + FORTYPCD + DSTRBCD1 + DSTRBCD2 + DSTRBCD3 + OWNGRPCD * DSTRBCD1, data = ca_grm_table)
summary(lmyr_mort2) # Adjusted R-squared: 0.3276

#interaction terms
lmi_mort1 <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD * DSTRBCD1, data = ca_grm_table)
summary(lmi_mort1) # Adjusted R-squared: 0.2655 

lmi_mort2 <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD * FORTYPCD, data = ca_grm_table)
summary(lmi_mort2) # Adjusted R-squared:  0.1107 

lmi_mort1_2 <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD * DSTRBCD1 + OWNGRPCD * FORTYPCD + DSTRBCD2 + DSTRBCD3, data = ca_grm_table) 
summary(lmi_mort1_2) # Adjusted R-squared: 0.3326

lmi_mort3 <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD * TRTCD1, data = ca_grm_table)
summary(lmi_mort3)
```



growth estimates
```{r}
lm1_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD, data = ca_grm_table_grp)
summary(lm1_grow) # Adjusted R-squared:  0.01448  

lm2_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD, data = ca_grm_table_grp)
summary(lm2_grow) # Adjusted R-squared:  0.3749 

lm3_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1, data = ca_grm_table_grp)
summary(lm3_grow) # Adjusted R-squared:  0.3752

lm4_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD, data = ca_grm_table_grp)
summary(lm4_grow) # Adjusted R-squared:  0.4797

lm5_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD, data = ca_grm_table_grp)
summary(lm5_grow) # Adjusted R-squared:  0.4937

lm6_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD + DSTRBCD1, data = ca_grm_table_grp)
summary(lm6_grow) # Adjusted R-squared:  0.5063

lm7_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD + DSTRBCD1 + YEAR, data = ca_grm_table_grp)
summary(lm7_grow) # Adjusted R-squared:  0.5113

lmDSTRB_grow <- lm(GROW_CARB_AG_ACRE ~ DSTRBCD1 + DSTRBCD2 + DSTRBCD3, data = ca_grm_table_grp)
summary(lmDSTRB_grow) # Adjusted R-squared: 0.05833 

# interacting terms
lmi_grow1 <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD * FORTYPCD, data = ca_grm_table)
summary(lmi_grow1) # Adjusted R-squared:  0.3821

lmi_grow2 <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD * FORTYPCD + OWNGRPCD * SITECLCD, data = ca_grm_table)
summary(lmi_grow2) # Adjusted R-squared:  0.4858 

lmi_growsite <-  lm(GROW_CARB_AG_ACRE ~ OWNGRPCD * SITECLCD, data = ca_grm_table)
summary(lmi_growsite) # Adjusted R-squared:  0.4007

lmi_grow3 <-  lm(GROW_CARB_AG_ACRE ~ OWNGRPCD * TRTCD1, data = ca_grm_table)
summary(lmi_grow3) # Adjusted R-squared:  0.02122

lmi_grow4 <-  lm(GROW_CARB_AG_ACRE ~ OWNGRPCD * FORTYPCD + OWNGRPCD * SITECLCD + PHYSCLCD, data = ca_grm_table)
summary(lmi_grow4) # Adjusted R-squared:  0.4986

lmi_grow5 <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD * DSTRBCD1, data = ca_grm_table)
summary(lmi_grow5) # Adjusted R-squared:  0.09258

# analyze linear regressions

result_grow <- AIC(lm3_grow, lm4_grow, lm5_grow, lm6_grow, lm7_grow)

models_grow <- list(lm3_grow, lm4_grow, lm5_grow, lm6_grow, lm7_grow)
result_grow$BIC <- sapply(models_grow, BIC) 
model_summary_grow <- lapply(models_grow, summary)

for(i in 1:length(models_grow)){ #this creates a variable i that starts with the value i=1
  result_grow$rsq[i] <- model_summary_grow[[i]]$r.squared #we assign the rsq value from model i to the i'th row of the column 'rsq' in the table 'results'
  result_grow$adj_rsq[i] <- model_summary_grow[[i]]$adj.r.squared #same for adjusted rsq
} #now we go back to the beginning of the for-loop, add 1 to the value of i, and do everything again

kable(result_grow, digits = 2, align = "c")


```

do private lands have less disturbance? (answer: yes)
```{r}
# prepare data
  # copy data
  test_ca_grm_table <- ca_grm_table
 
 # flag if plot has insect damage
  test_ca_grm_table$insect_damage <- ifelse(test_ca_grm_table$DSTRBCD1 %in% c(10, 11, 12), 1, 0)
  
  # flag if plot has disease 
  test_ca_grm_table$disease <- ifelse(test_ca_grm_table$DSTRBCD1 %in% c(20, 21, 22), 1, 0)
  
  # flag if plot has fire 
  test_ca_grm_table$fire <- ifelse(test_ca_grm_table$DSTRBCD1 %in% c(30, 31, 32), 1, 0)
  
  # flag if plot has drought
  test_ca_grm_table$drought <- ifelse(test_ca_grm_table$DSTRBCD1 == 54, 1, 0)
  
  # flag if plot has no disturbance
  test_ca_grm_table$none <- ifelse(test_ca_grm_table$DSTRBCD1 == 0, 1, 0)
  
  
#calculate proportion of plots with disturbance by ownership
  disturbance_rates <- test_ca_grm_table %>%
    group_by(OWNGRPCD) %>%
    summarize(
      n = n(), # dount plots in ownership group
      insect_rate = mean(insect_damage, na.rm = TRUE), # mean = proportion of plots with this kind of disturbance,
      disease_rate = mean(disease, na.rm = TRUE),
      fire_rate = mean(fire, na.rm = TRUE),
      drought_rate = mean(drought, na.rm = TRUE),
      none_rate = mean(none, na.rm = TRUE)
    )

# analyze
disturbance_rates #view

# export
write.csv(disturbance_rates, file = "disturbance_rates_ownership_CA.csv")

  # chi sq test
    # insect
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$insect_damage)) # p-value = 0.0001722
    fisher.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$insect_damage)) # p-value = 9.493e-05
    
    # disease
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$disease)) # p-value < 2.2e-16
    
    # fire
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$fire)) # p-value < 2.2e-16
    
    # drought
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$drought)) # p-value = 0.1483
    
    # none
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$none)) # p-value = 1.613e-08
    
```

do private lands have different treatment codes?
```{r}
# prepare data
  # # copy data (only if you did not run previous chunk abt disturbance)
  # test_ca_grm_table <- ca_grm_table
 
 # flag if plot has no treatment
  test_ca_grm_table$no_treatment <- ifelse(test_ca_grm_table$TRTCD1 == 0, 1, 0)
  
  # flag if plot has cutting 
  test_ca_grm_table$cut <- ifelse(test_ca_grm_table$TRTCD1 == 10, 1, 0)
  
   # flag if plot has other prep for regen (Clearing, slash burning, chopping, disking, bedding)
  test_ca_grm_table$prep <- ifelse(test_ca_grm_table$TRTCD1 == 20, 1, 0)
  
  # flag if plot has artificial regen 
  test_ca_grm_table$art_regen <- ifelse(test_ca_grm_table$TRTCD1 == 30, 1, 0)
  
  # flag if plot has natural regen
  test_ca_grm_table$nat_regen <- ifelse(test_ca_grm_table$TRTCD1 == 40, 1, 0)
  
  # flag if plot has silviculture treatment (fertilizers, herbicides, girdling, pruning,)
  test_ca_grm_table$silv <- ifelse(test_ca_grm_table$DSTRBCD1 == 50, 1, 0)
  
  
#calculate proportion of plots with disturbance by ownership
  treatment_rates <- test_ca_grm_table %>%
    group_by(OWNGRPCD) %>%
    summarize(
      n = n(), # dount plots in ownership group
      no_treatment_rate = mean(no_treatment, na.rm = TRUE), # mean = proportion of plots with this kind of disturbance,
      cut_rate = mean(cut, na.rm = TRUE),
      prep_rate = mean(prep, na.rm = TRUE),
      art_regen_rate = mean(art_regen, na.rm = TRUE),
      nat_regen_rate = mean(nat_regen, na.rm = TRUE),
      silv_rate = mean(silv, na.rm = TRUE)
    )

# analyze
treatment_rates #view

  # chi sq test
    # none
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$no_treatment)) # p-value < 2.2e-16
    
    # cutting
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$cut)) # p-value < 2.2e-16 
    
    # prep
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$prep)) # p-value = 0.764
    
    # art regen
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$art_regen)) # p-value = 0.2798
    
    # nat regen
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$nat_regen)) # p-value = 0.5924
    
    # silv
    fisher.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$silv)) # p-value = 2.608e-10
    
```

## min travel time

prep data
```{r}
# filter by CA
ca_min_travel_time <- min_travel_time[min_travel_time$STATE == 'CA',] #this leaves some plots with FIPS code 41 (OR), even though STATE = CA
ca_min_travel_time <- min_travel_time[min_travel_time$STATECD == '6',]

# the last 5 digits of pltID in ca_grm_table are PLOT IDs from ca_min_travel time

# make new column un ca_grm_with last 5 digits of pltID
ca_grm_table1 <- ca_grm_table

ca_grm_table1$PLOT_CD <- substring(ca_grm_table$pltID, nchar(ca_grm_table$pltID) - 4)

# now pivot so that time to chips and merchantable facilities are separate columns
ca_min_travel_time_clean <- ca_min_travel_time %>%
  dplyr::select(PLOT, BIOCD_DEF, ONE_WAY_HOURS) %>%
  group_by(PLOT, BIOCD_DEF) %>%
  # summarise(ONE_WAY_HOURS = min(ONE_WAY_HOURS), .groups = "drop") %>%
  pivot_wider(
    names_from = BIOCD_DEF,
    values_from = ONE_WAY_HOURS,
    names_prefix = "OWH_"
  )

# rename PLOT to match with ca_grm PLOT_CD
names(ca_min_travel_time_clean)[names(ca_min_travel_time_clean) == "PLOT"] <- "PLOT_CD"

# make PLOT_CD a character
ca_min_travel_time_clean$PLOT_CD <- as.character(ca_min_travel_time_clean$PLOT_CD)

# join with ca_grm
ca_grm_travel <- ca_grm_table1 %>%
  full_join(ca_min_travel_time_clean, by = 'PLOT_CD')

```

## fire spatial data
```{r}
# read in spatial fire perimeters
fire_per <- read_sf("/Volumes/Alyssa_Backup/InterAgencyFirePerimeterHistory_All_Years_View/InterAgencyFirePerimeterHistory_All_Years_View.shp")

# read in ca boundaries
ca <- read_sf("/Volumes/Alyssa_Backup/tl_2022_06_tract/tl_2022_06_tract.shp")
#plot(ca)

# Convert FIA plot data to an sf object 
  # first remove NAs
  ca_plot_clean <- ca_plot %>%
  filter(!is.na(LON) & !is.na(LAT))
  
  #then use coords to convert to sf
  fia_sf <- st_as_sf(ca_plot_clean, coords = c("LON", "LAT"), crs = 4326)
  
# match crs
  fire_per <- st_transform(fire_per, st_crs(fia_sf))
  
# This returns a list: one element per plot, containing indices of intersecting fire polygons
intersections <- st_intersects(fia_sf, fire_per, sparse = TRUE)

# Check if each plot has any intersections (i.e., non-empty list element)
fia_sf$fire_intersect <- lengths(intersections) > 0

# Convert to 1/0 instead of TRUE/FALSE 
fia_sf$fire_intersect <- as.integer(fia_sf$fire_intersect)

# now we will join to ca_grm_travel

  # First extract just the plot ID and fire_intersect column from the sf object
  fire_flag_df <- fia_sf %>%
  sf::st_drop_geometry() %>%
  dplyr::select(PLOT, fire_intersect) 
  
  # rename PLOT to match with ca_grm PLOT_CD
  names(fire_flag_df)[names(fire_flag_df) == "PLOT"] <- "PLOT_CD"

  # make PLOT_CD a character
  fire_flag_df$PLOT_CD <- as.character(fire_flag_df$PLOT_CD)
  
  # make sure each plot appears only once
  fire_flag_unique <- fire_flag_df %>%
  group_by(PLOT_CD) %>%
  summarise(fire_intersect = max(fire_intersect, na.rm = TRUE), .groups = "drop")
  
  # now join
  ca_grm_travel_fire <- ca_grm_travel %>%
  dplyr::left_join(fire_flag_unique, by = "PLOT_CD")
  
  # make fire_intersect a factor
  ca_grm_travel_fire$fire_intersect <- as.factor(ca_grm_travel_fire$fire_intersect)

############ this code fails due to memory limit
# # if fia plot coords intersect with fire polygons, 1; else 0
# fia_with_fire <- fia_sf %>%
#   mutate(fire_intersect = as.integer(rowSums(st_intersects(., fire_per, sparse = FALSE)) > 0))
############


```

analyze
```{r}
# plot GRM
plot(OWH_Chips ~ GROW_CARB_AG_ACRE, data = ca_grm_travel)
plot(OWH_Chips ~ REMV_CARB_AG_ACRE, data = ca_grm_travel)
plot(OWH_Chips ~ MORT_CARB_AG_ACRE, data = ca_grm_travel)

plot(OWH_Merchantable ~ GROW_CARB_AG_ACRE, data = ca_grm_travel)
plot(OWH_Merchantable ~ REMV_CARB_AG_ACRE, data = ca_grm_travel)
plot(OWH_Merchantable ~ MORT_CARB_AG_ACRE, data = ca_grm_travel)

# plot basal area of live trees
plot(OWH_Chips ~ BALIVE, data = ca_grm_travel)
plot(OWH_Merchantable ~ BALIVE, data = ca_grm_travel)

# plot growing stock stocking rate
plot(OWH_Chips ~ GSSTK, data = ca_grm_travel)
plot(OWH_Merchantable ~ GSSTK, data = ca_grm_travel)

# plot CARBON_DOWN_DEAD
plot(CARBON_DOWN_DEAD ~ OWH_Chips, data = ca_grm_travel)
plot(CARBON_DOWN_DEAD ~ OWH_Merchantable, data = ca_grm_travel)

# plot Disturbance
  # # aggregate disturbance codes
ca_grm_travel_grp <- ca_grm_travel %>%
  mutate(
    DSTRBCD1 = case_when(
      DSTRBCD1 %in% 10:12 ~ "Insect",
      DSTRBCD1 %in% 20:22 ~ "Disease",
      DSTRBCD1 %in% 30:32 ~ "Fire",
      DSTRBCD1 %in% 40:46 ~ "Animal",
      DSTRBCD1 %in% 50:54 ~ "Weather",
      DSTRBCD1 == 60:62 ~ "Vegetation",
      DSTRBCD1 == 70:79 ~ "Unknown",
      DSTRBCD1 == 80    ~ "Human",
      DSTRBCD1 %in% 90:95 ~ "Geologic",
      TRUE                        ~ "Other"
    )
  )
  # chips
    plot(OWH_Chips ~ DSTRBCD1, data = ca_grm_travel)
    # anova
        res_aov <- aov(OWH_Chips ~ DSTRBCD1,
        data = ca_grm_travel
      )
      
        summary(res_aov) # p = 6.57e-10
        
        # dunnett's
        post_test <- glht(res_aov,
      linfct = mcp(DSTRBCD1 = "Dunnett")
    )
    
    summary(post_test) # for fire p = 0.00865
    
  # merch
    plot(OWH_Merchantable ~ DSTRBCD1, data = ca_grm_travel)
    # anova
        res_aov <- aov(OWH_Merchantable ~ DSTRBCD1,
        data = ca_grm_travel
      )
      
        summary(res_aov) # p = 1.85e-13
        
        # dunnett's
        post_test <- glht(res_aov,
      linfct = mcp(DSTRBCD1 = "Dunnett")
    )
    
    summary(post_test) # for fire p = 0.02753


# plot spatial fire
  # chips
   boxplot(OWH_Chips ~ fire_intersect, data = ca_grm_travel_fire)
    # anova
        res_aov <- aov(OWH_Chips ~ fire_intersect,
        data = ca_grm_travel_fire
      )
      
        summary(res_aov) # p = 5.88e-10
    # means
    ca_grm_travel_fire %>%
    group_by(fire_intersect) %>%
    summarise(mean_OWH_Chips = mean(OWH_Chips, na.rm = TRUE),
              n = n()) # no fire =	1.613747 hours; fire = 	1.734607 hours	
    
  # merch
    boxplot(OWH_Merchantable ~ fire_intersect, data = ca_grm_travel_fire)
    # anova
        res_aov <- aov(OWH_Merchantable ~ fire_intersect,
        data = ca_grm_travel_fire
      )
      
        summary(res_aov) # p = 0.000751
  
      

# plot treatment
  # chips
    plot(OWH_Chips ~ TRTCD1, data = ca_grm_travel)
      # anova
        res_aov <- aov(OWH_Chips ~ TRTCD1,
        data = ca_grm_travel
      )
      
        summary(res_aov) # p<2e-16
        
        # dunnett's
        post_test <- glht(res_aov,
      linfct = mcp(TRTCD1 = "Dunnett")
    )
    
        summary(post_test) # for cutting p <1e-06 
    
  # merch
    plot(OWH_Merchantable ~ TRTCD1, data = ca_grm_travel)
  # anova
    res_aov <- aov(OWH_Merchantable ~ TRTCD1,
    data = ca_grm_travel
  )
  
    summary(res_aov) # p<2e-16
    
    # dunnett's
    post_test <- glht(res_aov,
  linfct = mcp(TRTCD1 = "Dunnett")
)

    summary(post_test) # for cutting p <1e-06 



# removal
lm1_remv_chips <- lm(REMV_CARB_AG_ACRE ~ OWH_Chips, data = ca_grm_travel)
summary(lm1_remv_chips) # Adjusted R-squared:  0.009079

lmt5_remv_chips <- lm(REMV_CARB_AG_ACRE ~ OWH_Chips + OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD, data = ca_grm_travel)
summary(lmt5_remv_chips) # Adjusted R-squared:  0.2447

lmt1_remv_merch <- lm(REMV_CARB_AG_ACRE ~ OWH_Merchantable, data = ca_grm_travel)
summary(lmt1_remv_merch) # Adjusted R-squared:  0.007083


lmt5_remv_merch <- lm(REMV_CARB_AG_ACRE ~ OWH_Merchantable + OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD, data = ca_grm_travel)
summary(lmt5_remv_merch) # Adjusted R-squared:  0.2397

lmtest <- lm(CARBON_DOWN_DEAD ~ OWH_Chips + OWNGRPCD, data = ca_grm_travel)
summary(lmtest)

lmtest <- lm(fire_intersect ~ OWH_Chips, data = ca_grm_travel_fire)


```
# statistical tests for mill distance write-up
Q1) Does distance from the nearest mill affect the probability of fire?
```{r}
# Fire Perimeters

  # Merchantable
    q1M <- glm(fire_intersect ~ OWH_Merchantable, data = ca_grm_travel_fire, family = "binomial")
    summary(q1M)
      # for every additional hour in distance to a Merchantable mill, the log odds of fire increase by 0.09028 (P = 0.000762)
    
    
  # Chips
    
    q1C <- glm(fire_intersect ~ OWH_Chips, data = ca_grm_travel_fire, family = "binomial")
    summary(q1C)
      # for every additional hour in distance to a Chips mill, the log odds of fire increase by 0.15161 (P = 6.79e-10)
    
#DSTRBCD
  # first duplicate the data so we don't mess it up
  ca_grm_travel_fiaFire <- ca_grm_travel_fire
    
  # clean data so fire is a binary variable
  ca_grm_travel_fiaFire_grp <- ca_grm_travel_fiaFire %>%
  mutate(
    DSTRBCD_FIRE = if_else(
      as.numeric(as.character(DSTRBCD1)) %in% 30:32 |
      as.numeric(as.character(DSTRBCD2)) %in% 30:32 |
      as.numeric(as.character(DSTRBCD3)) %in% 30:32,
      1, 0
    )
  )
  
  # now we can analyze
  
    # Merchantable
      q1Mfia <- glm(DSTRBCD_FIRE ~ OWH_Merchantable, data = ca_grm_travel_fiaFire_grp, family = "binomial")
      summary(q1Mfia)
        # for every additional hour in distance to a Merchantable mill, the log odds of fire increase by 0.18472 (P = 4.46e-05)
          # why is this twice the fire perimeter one? 
      
      
    # Chips
      
      q1Cfia <- glm(DSTRBCD_FIRE ~ OWH_Chips, data = ca_grm_travel_fiaFire_grp, family = "binomial")
      summary(q1Cfia)
        # for every additional hour in distance to a Chips mill, the log odds of fire increase by 0.20745 (P = 5.46e-07)
  
  
  
  
```
Q2) Does distance increase the probability of fire because there’s less management happening farther from mills (TRTCD1, Removals)
```{r}
# clean data for TRTCD = 10 (cutting) or 20 (Site preparation - Clearing, slash burning, chopping, disking, bedding, or other practices clearly intended to prepare a site for either natural or artificial regeneration.)
  ca_grm_travel_trtcd_grp <- ca_grm_travel_fiaFire %>%
  mutate(
    TRTCD_CUT_PREP = if_else(
      as.numeric(as.character(TRTCD1)) %in% 10:20 |
      as.numeric(as.character(TRTCD2)) %in% 10:20 |
      as.numeric(as.character(TRTCD3)) %in% 10:20,
      1, 0
    )
  )

# do the same for the fire dataset so we can analyze trtmnt and fire together: clean data for TRTCD = 10 (cutting) or 20 (Site preparation - Clearing, slash burning, chopping, disking, bedding, or other practices clearly intended to prepare a site for either natural or artificial regeneration.)
  ca_grm_travel_trtcd_grp <- ca_grm_travel_fiaFire_grp %>%
  mutate(
    TRTCD_CUT_PREP = if_else(
      as.numeric(as.character(TRTCD1)) %in% 10:20 |
      as.numeric(as.character(TRTCD2)) %in% 10:20 |
      as.numeric(as.character(TRTCD3)) %in% 10:20,
      1, 0
    )
  )

# is there less management farther from mills?
  
  # Merchantable
  q2aM <- glm(TRTCD_CUT_PREP ~ OWH_Merchantable, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2aM)
    # for every additional hour in distance to a Merchantable mill, the log odds of treatment decrease by 0.56457 (P <2e-16)
  
  # Chips
  q2aC <- glm(TRTCD_CUT_PREP ~ OWH_Chips, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2aC)
    # for every additional hour in distance to a Chip mill, the log odds of treatment decrease by 0.57778 (P <2e-16)

# is there more fire when there is less management, controlling for distance?

  # Merchantable
  q2bM <- glm(fire_intersect ~ OWH_Merchantable + TRTCD_CUT_PREP, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2bM)
    # for every additional hour in distance to a Merchantable mill, the log odds of fire increase by 0.08316 (P = 0.00202). This is down from 0.09028 when treatment is not considered, which means that treatment mediates the effect of distance on fire. The log odds of fire decrease by 0.25618 if there is treatment vs no treatment (p = 0.00519) 
  
  # Chips
  q2bC <- glm(fire_intersect ~ OWH_Chips + TRTCD_CUT_PREP, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2bC)
    # for every additional hour in distance to a Chip mill, the log odds of fire increase by 0.14422 (P = 5.3e-09). This is down from 0.1516 when treatment is not considered, meaning that treatment mediates the effect of distance on fire. The log odds of fire decrease by 0.25067 if there is treatment (p = 0.00534). 
  
  
# Mediation analysis: does management effect on fire interact with mill distance?
  
  # Merchantable
  med_outM <- mediate(q2aM, q2bM,
                     treat = "OWH_Merchantable",
                     mediator = "TRTCD_CUT_PREP",
                     boot = TRUE, sims = 1000)
  
  summary(med_outM)
    # Prop. Mediated (average) 0.132741 (P = 0.01), so 13% of the effect of distance from merchantable mill on fire is explained by reduced management
  
  #Chips
  med_outC <- mediate(q2aC, q2bC,
                     treat = "OWH_Chips",
                     mediator = "TRTCD_CUT_PREP",
                     boot = TRUE, sims = 1000)
  
  summary(med_outC)
    # Prop. Mediated (average) 0.08860 (P = 0.002), so 8.9% of the effect of distance from chip mill on fire is explained by reduced management
  
```

Q3) Does tree mortality increase as distance from mills increases?
```{r}
# Merchantable
q3aM <- lm(MORT_CARB_AG_ACRE ~ OWH_Merchantable, data = ca_grm_travel_fire)
summary(q3aM)

#Chips
q3aC <- lm(MORT_CARB_AG_ACRE ~ OWH_Chips, data = ca_grm_travel_fire)
summary(q3aC)
```

Q4) Does removal decrease with distance from mill?
```{r}
#Merchantable
q3bM <- lm(REMV_CARB_AG_ACRE ~ OWH_Merchantable, data = ca_grm_travel_fire)
summary(q3bM)
  # coefficient = -153.92, p = 8e-13 

#Chips
q3bC <- lm(REMV_CARB_AG_ACRE ~ OWH_Chips, data = ca_grm_travel_fire)
summary(q3bC)
  # coefficient = -164.52, p = 2.59e-16 

```
Q4: Does more carbon removed increase probability of fire?
```{r}
q4a <- glm(fire_intersect ~ REMV_CARB_AG_ACRE, data = ca_grm_travel_trtcd_grp, family = "binomial")
summary(q4a)
  # coefficient = -3.206e-05, p-value = 0.0473

```
#######################
#######################



## carbon
Carbon
```{r}
ca_cond <- ca_data$COND #cond table
ca_plot <- ca_data$PLOT #plot table
ca_tree <- ca_data$TREE #tree table

# get carbon by plot
carbon_plot<- carbon(ca_data, byPlot = TRUE)

# join
ca_table <- carbon_plot %>%
  # Join tables
  left_join(ca_cond, by = 'PLT_CN') # The foreign key for the COND table is PLT_CN. There is always a match of the PLT_CN value to the CN value in the PLOT table.

# make variables factors
ca_table$OWNGRPCD <- as.factor(ca_table$OWNGRPCD)
ca_table$FORTYPCD <- as.factor(ca_table$FORTYPCD)
ca_table$TRTCD1 <- as.factor(ca_table$TRTCD1)
ca_table$SITECLCD <- as.factor(ca_table$SITECLCD)
ca_table$PHYSCLCD <- as.factor(ca_table$PHYSCLCD)
ca_table$RESERVCD <- as.factor(ca_table$RESERVCD)


# carbon per acre estimates

lm1 <- lm(CARB_ACRE ~ OWNGRPCD, data = ca_table)
summary(lm1) # Adjusted R-squared:  0.0181 

lm2 <- lm(CARB_ACRE ~ OWNGRPCD + FORTYPCD, data = ca_table)
summary(lm2) # Adjusted R-squared:  0.1025 

lm3 <- lm(CARB_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1, data = ca_table)
summary(lm3) # Adjusted R-squared:  0.1028

lm4 <- lm(CARB_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD, data = ca_table)
summary(lm4) # Adjusted R-squared:  0.1152

lm5 <- lm(CARB_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD, data = ca_table)
summary(lm5) # Adjusted R-squared:  0.1221

```


analyze linear regressions
```{r}
result <- AIC(lm1, lm2, lm3, lm4, lm5)

models <- list(lm1, lm2, lm3, lm4, lm5)
result$BIC <- sapply(models, BIC) 
model_summary <- lapply(models, summary)

for(i in 1:length(models)){ #this creates a variable i that starts with the value i=1
  result$rsq[i] <- model_summary[[i]]$r.squared #we assign the rsq value from model i to the i'th row of the column 'rsq' in the table 'results'
  result$adj_rsq[i] <- model_summary[[i]]$adj.r.squared #same for adjusted rsq
} #now we go back to the beginning of the for-loop, add 1 to the value of i, and do everything again

kable(result, digits = 2, align = "c")


```