---
title: "rFIA carbon"
author: "Alyssa"
date: "4/9/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rFIA)
library(ggplot2)
library(dplyr)
library(stats)
library(knitr)
library(tidyverse)
library(tidyr)
library(stringr)
library(readr)
library(sf)
library(multcomp)
library(mediation)
library(terra)
library(mgcv)
library(purrr)
library(DHARMa)
library(nnet)
library(MASS)
```

load data
```{r}
ca_data <- readFIA("/Volumes/Alyssa_Backup/CA_CSV") 
min_travel_time <- read_csv("min_travel_time.csv")
              
```

## Carbon Calculations

Carbon by pool

```{r}
carbon(db = ca_data, byPool = TRUE)
```
Carbon by ownership

```{r}
#create table of carbon by pool and ownership
ownership <- carbon(db = ca_data, byPool = TRUE, grpBy = OWNGRPCD)

#aggregate by ownership and pool
ownership_summary <- ownership %>%
  group_by(OWNGRPCD, POOL) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# factorize
ownership_summary$POOL <- factor(ownership_summary$POOL)
ownership_summary$OWNGRPCD <- as.factor(ownership_summary$OWNGRPCD)

# plot
ggplot(ownership_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = POOL)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )


```
# Carbon by management

regeneration
```{r}
# natural (0) or artificial (1) regeneration
stand_regeneration <- carbon(db = ca_data, byPool = FALSE, grpBy = c(STDORGCD, OWNGRPCD))
stand_regeneration

#aggregate by ownership and pool
stand_regeneration_summary <- stand_regeneration %>%
  group_by(OWNGRPCD, STDORGCD) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# factorize
stand_regeneration_summary$STDORGCD <- factor(stand_regeneration_summary$STDORGCD)
stand_regeneration_summary$OWNGRPCD <- as.factor(stand_regeneration_summary$OWNGRPCD)

# plot
ggplot(stand_regeneration_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = STDORGCD)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )

```
treatment
```{r}
# treatment code 1
treatment1 <- carbon(db = ca_data, grpBy = TRTCD1)
treatment1

#aggregate by trtcd
carbon_by_treatment <- aggregate(CARB_ACRE ~ TRTCD1, data = treatment1, sum)

# barplot
barplot(carbon_by_treatment$CARB_ACRE,
        names.arg = carbon_by_treatment$TRTCD1,
        main = "Carbon by Treatment",
        xlab = "Treatment Code",
        ylab = "Carbon per Acre",
        las = 2)  # makes x-axis labels vertical for readability

# treatment and ownership
treatment_own <- carbon(db = ca_data, grpBy = c(TRTCD1, OWNGRPCD))

#factorize
treatment_own$TRTCD1 <- as.factor(treatment_own$TRTCD1)
levels(treatment_own$TRTCD1) # get different factor levels
levels(treatment_own$TRTCD1) <- c("None", "Cutting", "Regeneration Prep", "Artificial Regen", "Natural Regen", "Other") # rename for readability

treatment_own$OWNGRPCD <- as.factor(treatment_own$OWNGRPCD)
levels(treatment_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")


#aggregate by treatment and ownership
treatment_own_summary <- treatment_own %>%
  group_by(OWNGRPCD, TRTCD1) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(treatment_own_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = TRTCD1)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )
 

#plot without SE
ggplot(treatment_own, aes(fill=TRTCD1, y=CARB_ACRE, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")

```
cutting
```{r}
# stumps (cutting) and ownership
cutting_own <- carbon(db = ca_data, grpBy = c(STUMP_CD_PNWRS, OWNGRPCD))

#factorize
cutting_own$STUMP_CD_PNWRS <- as.factor(cutting_own$STUMP_CD_PNWRS)
levels(cutting_own$STUMP_CD_PNWRS) # get different factor levels

cutting_own$OWNGRPCD <- as.factor(cutting_own$OWNGRPCD)
levels(cutting_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#aggregate by treatment and ownership
cutting_own_summary <- cutting_own %>%
  group_by(OWNGRPCD, STUMP_CD_PNWRS) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(cutting_own_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = STUMP_CD_PNWRS)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )


#plot without SE
ggplot(cutting_own, aes(fill=STUMP_CD_PNWRS, y=CARB_ACRE, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")
```

site productivity code
```{r}
#site productivity and ownership
prod_own <- carbon(db = ca_data, grpBy = c(SITECLCD, OWNGRPCD))


#factorize
prod_own$SITECLCD <- as.factor(prod_own$SITECLCD)
levels(prod_own$SITECLCD) # get different factor levels

prod_own$OWNGRPCD <- as.factor(prod_own$OWNGRPCD)
levels(prod_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#aggregate by productivity and ownership
prod_own_summary <- prod_own %>%
  group_by(OWNGRPCD, SITECLCD) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(prod_own_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = SITECLCD)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )


#plot without SE
ggplot(prod_own, aes(fill=SITECLCD, y=CARB_ACRE, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")
```
forest type
```{r}
# read in fortypcd and group csv to aggregate later
fortypgrp <- read.csv("FORTYPCD_GRPCD.csv")

#find most common forest types by acre
fortyp_area <- area(db = ca_data, grpBy = FORTYPCD)
agg = aggregate(fortyp_area,
by = list(fortyp_area$FORTYPCD),
FUN = mean) #aggregate years

fortyp_order <- agg %>%
  arrange(desc(AREA_TOTAL))

top_fortyp <- agg %>%
  arrange(desc(AREA_TOTAL)) %>%
  slice_head(n = 6) %>%
  pull(FORTYPCD) # get top 6 forest type cd
top_fortyp

#now filter carbon by top fortypcd when doing carbon ownership analysis
top_fortyp_own <- carbon(db = ca_data, grpBy = c(FORTYPCD, OWNGRPCD))
top_fortyp_own$FORTYPCD <- as.factor(top_fortyp_own$FORTYPCD)

#filter
top_fortyp_own <- top_fortyp_own %>%
  filter(FORTYPCD %in% top_fortyp)
# check
summary(top_fortyp_own$FORTYPCD)

top_fortyp_own$OWNGRPCD <- as.factor(top_fortyp_own$OWNGRPCD)
levels(top_fortyp_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#plot
ggplot(top_fortyp_own, aes(fill=FORTYPCD, y=CARB_ACRE, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")

# error bars
#aggregate by fortyp and ownership
top_fortyp_own_summary <- top_fortyp_own %>%
  group_by(OWNGRPCD, FORTYPCD) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(top_fortyp_own_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = FORTYPCD)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )
####################
# fortype by owner by area
fortyp_area_own <- area(db = ca_data, grpBy = c(FORTYPCD, OWNGRPCD))
fortyp_area_own$FORTYPCD <- as.factor(fortyp_area_own$FORTYPCD)

#filter
top_fortyp_area_own <- fortyp_area_own %>%
  filter(FORTYPCD %in% top_fortyp)
# check
summary(top_fortyp_area_own$FORTYPCD)

top_fortyp_area_own$OWNGRPCD <- as.factor(top_fortyp_area_own$OWNGRPCD)
levels(top_fortyp_area_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#plot
ggplot(top_fortyp_area_own, aes(fill=FORTYPCD, y=AREA_TOTAL, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")

ggplot(top_fortyp_area_own, aes(fill=FORTYPCD, y=PERC_AREA, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")

# error bars
#aggregate by fortyp and ownership
top_fortyp_area_own_summary <- top_fortyp_area_own %>%
  group_by(OWNGRPCD, FORTYPCD) %>%
  summarise(
    PERC_AREA = mean(PERC_AREA),
    PERC_AREA_SE = mean(PERC_AREA_SE),
    .groups = 'drop'
  )

# plot
ggplot(top_fortyp_area_own_summary, aes(x = OWNGRPCD, y = PERC_AREA, fill = FORTYPCD)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = PERC_AREA - PERC_AREA_SE, ymax = PERC_AREA + PERC_AREA_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )

###################

#carbon
fortyp_own <- carbon(db = ca_data, grpBy = c(FORTYPCD, OWNGRPCD))
fortyp_own
fortyp_own$FORTYPCD <- as.factor(fortyp_own$FORTYPCD)

levels(fortyp_own$FORTYPCD) # get different factor levels


fortyp_own$OWNGRPCD <- as.factor(fortyp_own$OWNGRPCD)
levels(fortyp_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability


#plot
ggplot(fortyp_own, aes(fill=FORTYPCD, y=CARB_ACRE, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")
```
```{r}
# disturbance
dstrb_own <- carbon(db = ca_data, grpBy = c(DSTRBCD1, OWNGRPCD))

# factorize
dstrb_own$DSTRBCD1 <- as.factor(dstrb_own$DSTRBCD1)
levels(dstrb_own$DSTRBCD1) # get different factor levels

# aggregate disturbance codes
dstrb_own_grp <- dstrb_own %>%
  mutate(
    DSTRBCD1 = case_when(
      DSTRBCD1 %in% 10:12 ~ "Insect",
      DSTRBCD1 %in% 20:22 ~ "Disease",
      DSTRBCD1 %in% 30:32 ~ "Fire",
      DSTRBCD1 %in% 40:46 ~ "Animal",
      DSTRBCD1 %in% 50:54 ~ "Weather",
      DSTRBCD1 == 60:62 ~ "Vegetation",
      DSTRBCD1 == 70:79 ~ "Unknown",
      DSTRBCD1 == 80    ~ "Human",
      DSTRBCD1 %in% 90:95 ~ "Geologic",
      TRUE                        ~ "Other"
    )
  )

# factor owners
dstrb_own_grp$OWNGRPCD <- as.factor(dstrb_own_grp$OWNGRPCD)
levels(dstrb_own_grp$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#aggregate by disturbance and ownership
dstrb_own_grp_summary <- dstrb_own_grp %>%
  group_by(OWNGRPCD, DSTRBCD1) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(dstrb_own_grp_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = DSTRBCD1)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )

```


available moisture
```{r}
#site productivity and ownership
moist_own <- carbon(db = ca_data, grpBy = c(PHYSCLCD, OWNGRPCD))

#factorize
moist_own$PHYSCLCD <- as.factor(moist_own$PHYSCLCD)
levels(moist_own$PHYSCLCD) # get different factor levels

moist_own$OWNGRPCD <- as.factor(moist_own$OWNGRPCD)
levels(moist_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#aggregate by moist and ownership
moist_own_summary <- moist_own %>%
  group_by(OWNGRPCD, PHYSCLCD) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(moist_own_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = PHYSCLCD)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )


#plot
ggplot(moist_own, aes(fill=PHYSCLCD, y=CARB_ACRE, x=OWNGRPCD)) + 
    geom_bar(position="dodge", stat="identity")
```
stocking
```{r}
#site productivity and ownership
stock_own <- carbon(db = ca_data, grpBy = c(GSSTKCD, OWNGRPCD))

#factorize
stock_own$GSSTKCD <- as.factor(stock_own$GSSTKCD)
levels(stock_own$GSSTKCD) # get different factor levels

stock_own$OWNGRPCD <- as.factor(stock_own$OWNGRPCD)
levels(stock_own$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability


#aggregate by stocking and ownership
stock_own_summary <- stock_own %>%
  group_by(OWNGRPCD, GSSTKCD) %>%
  summarise(
    CARB_ACRE = mean(CARB_ACRE),
    CARB_ACRE_SE = mean(CARB_ACRE_SE),
    .groups = 'drop'
  )

# plot
ggplot(stock_own_summary, aes(x = OWNGRPCD, y = CARB_ACRE, fill = GSSTKCD)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.4,
    colour = "black",
    alpha = 0.9,
    size = 0.5
  )

```
## growMort
```{r}
# get carbon flux by ownership
carb_grm_agg <- growMort(ca_data, stateVar = "CARB", grpBy = "OWNGRPCD")

#factorize
carb_grm_agg$OWNGRPCD <- as.factor(carb_grm_agg$OWNGRPCD)
levels(carb_grm_agg$OWNGRPCD) <- c("Forest Service", "Other Federal", "State/Local Gov", "Private")  # rename for readability

#aggregate by ownership
carb_grm_summary <- carb_grm_agg %>%
  group_by(OWNGRPCD) %>%
  summarise(
    RECR_CARB_ACRE = mean(RECR_CARB_ACRE),
    RECR_CARB_ACRE_SE = mean(RECR_CARB_ACRE_SE),
    MORT_CARB_ACRE = mean(MORT_CARB_ACRE),
    MORT_CARB_ACRE_SE = mean(MORT_CARB_ACRE_SE),
    REMV_CARB_ACRE = mean(REMV_CARB_ACRE),
    REMV_CARB_ACRE_SE = mean(REMV_CARB_ACRE_SE),
    GROW_CARB_ACRE = mean(GROW_CARB_ACRE),
    GROW_CARB_ACRE_SE = mean(GROW_CARB_ACRE_SE),
    CHNG_CARB_ACRE = mean(CHNG_CARB_ACRE),
    .groups = 'drop'
  )

# make mort and remv negative
carb_grm_summary$MORT_CARB_ACRE <- carb_grm_summary$MORT_CARB_ACRE * -1
carb_grm_summary$REMV_CARB_ACRE <- carb_grm_summary$REMV_CARB_ACRE * -1
## comparison

# 1. Pivot GRM values to long format
grm_vals <- carb_grm_summary %>%
  select(OWNGRPCD, RECR_CARB_ACRE, MORT_CARB_ACRE, REMV_CARB_ACRE, GROW_CARB_ACRE) %>%
  pivot_longer(
    cols = ends_with("CARB_ACRE"),
    names_to = "GRM",
    values_to = "CARB_ACRE"
  ) %>%
  mutate(GRM = str_remove(GRM, "_CARB_ACRE"))

# 2. Pivot SE values to long format
grm_ses <- carb_grm_summary %>%
  select(OWNGRPCD, RECR_CARB_ACRE_SE, MORT_CARB_ACRE_SE, REMV_CARB_ACRE_SE, GROW_CARB_ACRE_SE) %>%
  pivot_longer(
    cols = ends_with("_SE"),
    names_to = "GRM",
    values_to = "CARB_ACRE_SE"
  ) %>%
  mutate(GRM = str_remove(GRM, "_CARB_ACRE_SE"))

# 3. Join the values + SEs
grm_combined <- left_join(grm_vals, grm_ses, by = c("OWNGRPCD", "GRM"))


# 4. Plot
# stacked
  # Pivot CHNG values to long format
  chng_vals <- carb_grm_summary %>%
    select(OWNGRPCD, CHNG_CARB_ACRE) %>%
    pivot_longer(
      cols = ends_with("CARB_ACRE"),
      names_to = "GRM",
      values_to = "CARB_ACRE"
    ) %>%
    mutate(GRM = str_remove(GRM, "_CARB_ACRE"))
  
  #plot
  ggplot(grm_combined, aes(x = OWNGRPCD, y = CARB_ACRE, fill = GRM)) +
    geom_bar(stat = "identity", position = "stack") + 
    geom_point(
    data = chng_vals,
    aes(x = OWNGRPCD, y = CARB_ACRE),
    color = "red",
    size = 3
)


# staggered
ggplot(grm_combined, aes(x = OWNGRPCD, y = CARB_ACRE, fill = GRM)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_errorbar(
    aes(ymin = CARB_ACRE - CARB_ACRE_SE, ymax = CARB_ACRE + CARB_ACRE_SE),
    position = position_dodge(width = 0.9),
    width = 0.3,
    color = "black"
  ) +
  theme_minimal() +
  labs(
    x = "Ownership Group",
    y = "Carbon per Acre",
    fill = "GRM Component"
  )

```

## spatial ownership dynamics
```{r}
# load in taskforce regions
regions <- read_sf("/Volumes/Alyssa_Backup/rFIA_work/RRK_boundaries4/RRK_boundaries4.shp")

# get carbon for ownership by region
carbon_own_region <- carbon(db = ca_data, byPool = FALSE, polys = regions, grpBy = OWNGRPCD, totals = TRUE, returnSpatial = TRUE)

# plot
plot(carbon_own_region["CARB_ACRE"])
```

## Regression
# GRM
```{r}
ca_cond <- ca_data$COND #cond table
ca_plot <- ca_data$PLOT #plot table
ca_tree <- ca_data$TREE #tree table

# get grm by plot
grm_plot <- growMort(ca_data, byPlot = TRUE, stateVar = "CARB_AG")

# join
ca_grm_table <- grm_plot %>%
  # Join tables
  left_join(ca_cond, by = 'PLT_CN') # The foreign key for the COND table is PLT_CN. There is always a match of the PLT_CN value to the CN value in the PLOT table.

# # filter by top fortypcd
# top_fortyp_ca_grm_table <- ca_grm_table %>%
#   filter(FORTYPCD %in% top_fortyp)
# # check
# summary(top_fortyp_ca_grm_table$FORTYPCD)

# # aggregate disturbances
# # aggregate disturbance codes
# ca_grm_table_grp <- ca_grm_table %>%
#   mutate(
#     DSTRBCD1 = case_when(
#       DSTRBCD1 %in% 10:12 ~ "Insect",
#       DSTRBCD1 %in% 20:22 ~ "Disease",
#       DSTRBCD1 %in% 30:32 ~ "Fire",
#       DSTRBCD1 %in% 40:46 ~ "Animal",
#       DSTRBCD1 %in% 50:54 ~ "Weather",
#       DSTRBCD1 == 60:62 ~ "Vegetation",
#       DSTRBCD1 == 70:79 ~ "Unknown",
#       DSTRBCD1 == 80    ~ "Human",
#       DSTRBCD1 %in% 90:95 ~ "Geologic",
#       TRUE                        ~ "Other"
#     )
#   )


# make variables factors
ca_grm_table$OWNGRPCD <- as.factor(ca_grm_table$OWNGRPCD)
ca_grm_table$FORTYPCD <- as.factor(ca_grm_table$FORTYPCD)
ca_grm_table$TRTCD1 <- as.factor(ca_grm_table$TRTCD1)
ca_grm_table$SITECLCD <- as.factor(ca_grm_table$SITECLCD)
ca_grm_table$PHYSCLCD <- as.factor(ca_grm_table$PHYSCLCD)
ca_grm_table$RESERVCD <- as.factor(ca_grm_table$RESERVCD)
ca_grm_table$YEAR <- as.factor(ca_grm_table$YEAR)
ca_grm_table$DSTRBCD1 <- as.factor(ca_grm_table$DSTRBCD1)
ca_grm_table$DSTRBCD2 <- as.factor(ca_grm_table$DSTRBCD2)
ca_grm_table$DSTRBCD3 <- as.factor(ca_grm_table$DSTRBCD3)
```

remv estimates (lm4_remv is best so far)
```{r}
lm1_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD, data = ca_grm_table)
summary(lm1_remv) # Adjusted R-squared:  0.02081  

lm2_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD, data = ca_grm_table_grp)
summary(lm2_remv) # Adjusted R-squared:  0.07301 

lm3_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1, data = ca_grm_table_grp)
summary(lm3_remv) # Adjusted R-squared:  0.234

lm4_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD, data = ca_grm_table_grp)
summary(lm4_remv) # Adjusted R-squared:  0.2423

lm5_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD, data = ca_grm_table_grp)
summary(lm5_remv) # Adjusted R-squared:  0.2422

lm6_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD + DSTRBCD1, data = ca_grm_table_grp)
summary(lm6_remv) # Adjusted R-squared:  0.242

lmDSTRB_remv <- lm(REMV_CARB_AG_ACRE ~ DSTRBCD1 + DSTRBCD2 + DSTRBCD3, data = ca_grm_table_grp)
summary(lmDSTRB_remv) # Adjusted R-squared: 0.01728

lm1DSTRB_remv <- lm(REMV_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + DSTRBCD1 + DSTRBCD2 + DSTRBCD3, data = ca_grm_table_grp)
summary(lm1DSTRB_remv) # Adjusted R-squared: 0.08971

```

mortality estimates
```{r}
lm1_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD, data = ca_grm_table_grp)
summary(lm1_mort) # Adjusted R-squared:  0.03153  

lm2_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD, data = ca_grm_table_grp)
summary(lm2_mort) # Adjusted R-squared:  0.09861 

lm3_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1, data = ca_grm_table_grp)
summary(lm3_mort) # Adjusted R-squared:  0.09875

lm4_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD, data = ca_grm_table_grp)
summary(lm4_mort) # Adjusted R-squared:  0.1289

lm5_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD, data = ca_grm_table_grp)
summary(lm5_mort) # Adjusted R-squared:  0.1346

lm6_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD + DSTRBCD1, data = ca_grm_table_grp)
summary(lm6_mort) # Adjusted R-squared:  0.2948

lm7_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD + DSTRBCD1 + YEAR, data = ca_grm_table)
summary(lm7_mort) # Adjusted R-squared:  0.3274

lmDSTRB_mort <- lm(MORT_CARB_AG_ACRE ~ DSTRBCD1 + DSTRBCD2 + DSTRBCD3, data = ca_grm_table_grp)
summary(lmDSTRB_mort) # Adjusted R-squared: 0.2692 

lm1DSTRB_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + DSTRBCD1 + DSTRBCD2 + DSTRBCD3, data = ca_grm_table_grp)
summary(lm1DSTRB_mort) # Adjusted R-squared: 0.2747 

lmyr_mort <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + YEAR + FORTYPCD + DSTRBCD1 + DSTRBCD2 + DSTRBCD3, data = ca_grm_table)
summary(lmyr_mort) # Adjusted R-squared: 0.3101

lmyr_mort2 <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD + YEAR + FORTYPCD + DSTRBCD1 + DSTRBCD2 + DSTRBCD3 + OWNGRPCD * DSTRBCD1, data = ca_grm_table)
summary(lmyr_mort2) # Adjusted R-squared: 0.3276

#interaction terms
lmi_mort1 <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD * DSTRBCD1, data = ca_grm_table)
summary(lmi_mort1) # Adjusted R-squared: 0.2655 

lmi_mort2 <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD * FORTYPCD, data = ca_grm_table)
summary(lmi_mort2) # Adjusted R-squared:  0.1107 

lmi_mort1_2 <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD * DSTRBCD1 + OWNGRPCD * FORTYPCD + DSTRBCD2 + DSTRBCD3, data = ca_grm_table) 
summary(lmi_mort1_2) # Adjusted R-squared: 0.3326

lmi_mort3 <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD * TRTCD1, data = ca_grm_table)
summary(lmi_mort3)
```



growth estimates
```{r}
lm1_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD, data = ca_grm_table_grp)
summary(lm1_grow) # Adjusted R-squared:  0.01448  

lm2_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD, data = ca_grm_table_grp)
summary(lm2_grow) # Adjusted R-squared:  0.3749 

lm3_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1, data = ca_grm_table_grp)
summary(lm3_grow) # Adjusted R-squared:  0.3752

lm4_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD, data = ca_grm_table_grp)
summary(lm4_grow) # Adjusted R-squared:  0.4797

lm5_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD, data = ca_grm_table_grp)
summary(lm5_grow) # Adjusted R-squared:  0.4937

lm6_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD + DSTRBCD1, data = ca_grm_table_grp)
summary(lm6_grow) # Adjusted R-squared:  0.5063

lm7_grow <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD + DSTRBCD1 + YEAR, data = ca_grm_table_grp)
summary(lm7_grow) # Adjusted R-squared:  0.5113

lmDSTRB_grow <- lm(GROW_CARB_AG_ACRE ~ DSTRBCD1 + DSTRBCD2 + DSTRBCD3, data = ca_grm_table_grp)
summary(lmDSTRB_grow) # Adjusted R-squared: 0.05833 

# interacting terms
lmi_grow1 <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD * FORTYPCD, data = ca_grm_table)
summary(lmi_grow1) # Adjusted R-squared:  0.3821

lmi_grow2 <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD * FORTYPCD + OWNGRPCD * SITECLCD, data = ca_grm_table)
summary(lmi_grow2) # Adjusted R-squared:  0.4858 

lmi_growsite <-  lm(GROW_CARB_AG_ACRE ~ OWNGRPCD * SITECLCD, data = ca_grm_table)
summary(lmi_growsite) # Adjusted R-squared:  0.4007

lmi_grow3 <-  lm(GROW_CARB_AG_ACRE ~ OWNGRPCD * TRTCD1, data = ca_grm_table)
summary(lmi_grow3) # Adjusted R-squared:  0.02122

lmi_grow4 <-  lm(GROW_CARB_AG_ACRE ~ OWNGRPCD * FORTYPCD + OWNGRPCD * SITECLCD + PHYSCLCD, data = ca_grm_table)
summary(lmi_grow4) # Adjusted R-squared:  0.4986

lmi_grow5 <- lm(GROW_CARB_AG_ACRE ~ OWNGRPCD * DSTRBCD1, data = ca_grm_table)
summary(lmi_grow5) # Adjusted R-squared:  0.09258

# analyze linear regressions

result_grow <- AIC(lm3_grow, lm4_grow, lm5_grow, lm6_grow, lm7_grow)

models_grow <- list(lm3_grow, lm4_grow, lm5_grow, lm6_grow, lm7_grow)
result_grow$BIC <- sapply(models_grow, BIC) 
model_summary_grow <- lapply(models_grow, summary)

for(i in 1:length(models_grow)){ #this creates a variable i that starts with the value i=1
  result_grow$rsq[i] <- model_summary_grow[[i]]$r.squared #we assign the rsq value from model i to the i'th row of the column 'rsq' in the table 'results'
  result_grow$adj_rsq[i] <- model_summary_grow[[i]]$adj.r.squared #same for adjusted rsq
} #now we go back to the beginning of the for-loop, add 1 to the value of i, and do everything again

kable(result_grow, digits = 2, align = "c")


```

do private lands have less disturbance? (answer: yes)
```{r}
# prepare data
  # copy data
  test_ca_grm_table <- ca_grm_table
 
 # flag if plot has insect damage
  test_ca_grm_table$insect_damage <- ifelse(test_ca_grm_table$DSTRBCD1 %in% c(10, 11, 12), 1, 0)
  
  # flag if plot has disease 
  test_ca_grm_table$disease <- ifelse(test_ca_grm_table$DSTRBCD1 %in% c(20, 21, 22), 1, 0)
  
  # flag if plot has fire 
  test_ca_grm_table$fire <- ifelse(test_ca_grm_table$DSTRBCD1 %in% c(30, 31, 32), 1, 0)
  
  # flag if plot has drought
  test_ca_grm_table$drought <- ifelse(test_ca_grm_table$DSTRBCD1 == 54, 1, 0)
  
  # flag if plot has no disturbance
  test_ca_grm_table$none <- ifelse(test_ca_grm_table$DSTRBCD1 == 0, 1, 0)
  
  
#calculate proportion of plots with disturbance by ownership
  disturbance_rates <- test_ca_grm_table %>%
    group_by(OWNGRPCD) %>%
    summarize(
      n = n(), # dount plots in ownership group
      insect_rate = mean(insect_damage, na.rm = TRUE), # mean = proportion of plots with this kind of disturbance,
      disease_rate = mean(disease, na.rm = TRUE),
      fire_rate = mean(fire, na.rm = TRUE),
      drought_rate = mean(drought, na.rm = TRUE),
      none_rate = mean(none, na.rm = TRUE)
    )

# analyze
disturbance_rates #view

# export
write.csv(disturbance_rates, file = "disturbance_rates_ownership_CA.csv")

  # chi sq test
    # insect
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$insect_damage)) # p-value = 0.0001722
    fisher.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$insect_damage)) # p-value = 9.493e-05
    
    # disease
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$disease)) # p-value < 2.2e-16
    
    # fire
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$fire)) # p-value < 2.2e-16
    
    # drought
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$drought)) # p-value = 0.1483
    
    # none
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$none)) # p-value = 1.613e-08
    
```

do private lands have different treatment codes?
```{r}
# prepare data
  # # copy data (only if you did not run previous chunk abt disturbance)
  # test_ca_grm_table <- ca_grm_table
 
 # flag if plot has no treatment
  test_ca_grm_table$no_treatment <- ifelse(test_ca_grm_table$TRTCD1 == 0, 1, 0)
  
  # flag if plot has cutting 
  test_ca_grm_table$cut <- ifelse(test_ca_grm_table$TRTCD1 == 10, 1, 0)
  
   # flag if plot has other prep for regen (Clearing, slash burning, chopping, disking, bedding)
  test_ca_grm_table$prep <- ifelse(test_ca_grm_table$TRTCD1 == 20, 1, 0)
  
  # flag if plot has artificial regen 
  test_ca_grm_table$art_regen <- ifelse(test_ca_grm_table$TRTCD1 == 30, 1, 0)
  
  # flag if plot has natural regen
  test_ca_grm_table$nat_regen <- ifelse(test_ca_grm_table$TRTCD1 == 40, 1, 0)
  
  # flag if plot has silviculture treatment (fertilizers, herbicides, girdling, pruning,)
  test_ca_grm_table$silv <- ifelse(test_ca_grm_table$DSTRBCD1 == 50, 1, 0)
  
  
#calculate proportion of plots with disturbance by ownership
  treatment_rates <- test_ca_grm_table %>%
    group_by(OWNGRPCD) %>%
    summarize(
      n = n(), # dount plots in ownership group
      no_treatment_rate = mean(no_treatment, na.rm = TRUE), # mean = proportion of plots with this kind of disturbance,
      cut_rate = mean(cut, na.rm = TRUE),
      prep_rate = mean(prep, na.rm = TRUE),
      art_regen_rate = mean(art_regen, na.rm = TRUE),
      nat_regen_rate = mean(nat_regen, na.rm = TRUE),
      silv_rate = mean(silv, na.rm = TRUE)
    )

# analyze
treatment_rates #view

  # chi sq test
    # none
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$no_treatment)) # p-value < 2.2e-16
    
    # cutting
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$cut)) # p-value < 2.2e-16 
    
    # prep
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$prep)) # p-value = 0.764
    
    # art regen
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$art_regen)) # p-value = 0.2798
    
    # nat regen
    chisq.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$nat_regen)) # p-value = 0.5924
    
    # silv
    fisher.test(table(test_ca_grm_table$OWNGRPCD, test_ca_grm_table$silv)) # p-value = 2.608e-10
    
```

## min travel time

prep data
```{r}
# filter by CA
ca_min_travel_time <- min_travel_time[min_travel_time$STATE == 'CA',] #this leaves some plots with FIPS code 41 (OR), even though STATE = CA
ca_min_travel_time <- min_travel_time[min_travel_time$STATECD == '6',]

# the last 5 digits of pltID in ca_grm_table are PLOT IDs from ca_min_travel time

# make new column un ca_grm_with last 5 digits of pltID
ca_grm_table1 <- ca_grm_table

ca_grm_table1$PLOT_CD <- substring(ca_grm_table$pltID, nchar(ca_grm_table$pltID) - 4)

# now pivot so that time to chips and merchantable facilities are separate columns
ca_min_travel_time_clean <- ca_min_travel_time %>%
  dplyr::select(PLOT, BIOCD_DEF, ONE_WAY_HOURS) %>%
  group_by(PLOT, BIOCD_DEF) %>%
  # summarise(ONE_WAY_HOURS = min(ONE_WAY_HOURS), .groups = "drop") %>%
  pivot_wider(
    names_from = BIOCD_DEF,
    values_from = ONE_WAY_HOURS,
    names_prefix = "OWH_"
  )

# rename PLOT to match with ca_grm PLOT_CD
names(ca_min_travel_time_clean)[names(ca_min_travel_time_clean) == "PLOT"] <- "PLOT_CD"

# make PLOT_CD a character
ca_min_travel_time_clean$PLOT_CD <- as.character(ca_min_travel_time_clean$PLOT_CD)

# join with ca_grm
ca_grm_travel <- ca_grm_table1 %>%
  full_join(ca_min_travel_time_clean, by = 'PLOT_CD')

```

## fire spatial data
```{r}
# read in spatial fire perimeters
fire_per <- read_sf("/Volumes/Alyssa_Backup/InterAgencyFirePerimeterHistory_All_Years_View/InterAgencyFirePerimeterHistory_All_Years_View.shp")

# read in ca boundaries
ca <- read_sf("/Volumes/Alyssa_Backup/tl_2022_06_tract/tl_2022_06_tract.shp")
#plot(ca)

# Convert FIA plot data to an sf object 
  # first remove NAs
  ca_plot_clean <- ca_plot %>%
  filter(!is.na(LON) & !is.na(LAT))
  
  #then use coords to convert to sf
  fia_sf <- st_as_sf(ca_plot_clean, coords = c("LON", "LAT"), crs = 4326)
  
# match crs
  fire_per <- st_transform(fire_per, st_crs(fia_sf))
  
# This returns a list: one element per plot, containing indices of intersecting fire polygons
intersections <- st_intersects(fia_sf, fire_per, sparse = TRUE)

# Check if each plot has any intersections (i.e., non-empty list element)
fia_sf$fire_intersect <- lengths(intersections) > 0

# Convert to 1/0 instead of TRUE/FALSE 
fia_sf$fire_intersect <- as.integer(fia_sf$fire_intersect)

# now we will join to ca_grm_travel

  # First extract just the plot ID and fire_intersect column from the sf object
  fire_flag_df <- fia_sf %>%
  sf::st_drop_geometry() %>%
  dplyr::select(PLOT, fire_intersect) 
  
  # rename PLOT to match with ca_grm PLOT_CD
  names(fire_flag_df)[names(fire_flag_df) == "PLOT"] <- "PLOT_CD"

  # make PLOT_CD a character
  fire_flag_df$PLOT_CD <- as.character(fire_flag_df$PLOT_CD)
  
  # make sure each plot appears only once
  fire_flag_unique <- fire_flag_df %>%
  group_by(PLOT_CD) %>%
  summarise(fire_intersect = max(fire_intersect, na.rm = TRUE), .groups = "drop")
  
  # now join
  ca_grm_travel_fire <- ca_grm_travel %>%
  dplyr::left_join(fire_flag_unique, by = "PLOT_CD")
  
  # make fire_intersect a factor
  ca_grm_travel_fire$fire_intersect <- as.factor(ca_grm_travel_fire$fire_intersect)

############ this code fails due to memory limit
# # if fia plot coords intersect with fire polygons, 1; else 0
# fia_with_fire <- fia_sf %>%
#   mutate(fire_intersect = as.integer(rowSums(st_intersects(., fire_per, sparse = FALSE)) > 0))
############


```

analyze
```{r}
# plot GRM
plot(OWH_Chips ~ GROW_CARB_AG_ACRE, data = ca_grm_travel)
plot(OWH_Chips ~ REMV_CARB_AG_ACRE, data = ca_grm_travel)
plot(OWH_Chips ~ MORT_CARB_AG_ACRE, data = ca_grm_travel)

plot(OWH_Merchantable ~ GROW_CARB_AG_ACRE, data = ca_grm_travel)
plot(OWH_Merchantable ~ REMV_CARB_AG_ACRE, data = ca_grm_travel)
plot(OWH_Merchantable ~ MORT_CARB_AG_ACRE, data = ca_grm_travel)

# plot basal area of live trees
plot(OWH_Chips ~ BALIVE, data = ca_grm_travel)
plot(OWH_Merchantable ~ BALIVE, data = ca_grm_travel)

# plot growing stock stocking rate
plot(OWH_Chips ~ GSSTK, data = ca_grm_travel)
plot(OWH_Merchantable ~ GSSTK, data = ca_grm_travel)

# plot CARBON_DOWN_DEAD
plot(CARBON_DOWN_DEAD ~ OWH_Chips, data = ca_grm_travel)
plot(CARBON_DOWN_DEAD ~ OWH_Merchantable, data = ca_grm_travel)

# plot Disturbance
  # # aggregate disturbance codes
ca_grm_travel_grp <- ca_grm_travel %>%
  mutate(
    DSTRBCD1 = case_when(
      DSTRBCD1 %in% 10:12 ~ "Insect",
      DSTRBCD1 %in% 20:22 ~ "Disease",
      DSTRBCD1 %in% 30:32 ~ "Fire",
      DSTRBCD1 %in% 40:46 ~ "Animal",
      DSTRBCD1 %in% 50:54 ~ "Weather",
      DSTRBCD1 == 60:62 ~ "Vegetation",
      DSTRBCD1 == 70:79 ~ "Unknown",
      DSTRBCD1 == 80    ~ "Human",
      DSTRBCD1 %in% 90:95 ~ "Geologic",
      TRUE                        ~ "Other"
    )
  )
  # chips
    plot(OWH_Chips ~ DSTRBCD1, data = ca_grm_travel)
    # anova
        res_aov <- aov(OWH_Chips ~ DSTRBCD1,
        data = ca_grm_travel
      )
      
        summary(res_aov) # p = 6.57e-10
        
        # dunnett's
        post_test <- glht(res_aov,
      linfct = mcp(DSTRBCD1 = "Dunnett")
    )
    
    summary(post_test) # for fire p = 0.00865
    
  # merch
    plot(OWH_Merchantable ~ DSTRBCD1, data = ca_grm_travel)
    # anova
        res_aov <- aov(OWH_Merchantable ~ DSTRBCD1,
        data = ca_grm_travel
      )
      
        summary(res_aov) # p = 1.85e-13
        
        # dunnett's
        post_test <- glht(res_aov,
      linfct = mcp(DSTRBCD1 = "Dunnett")
    )
    
    summary(post_test) # for fire p = 0.02753


# plot spatial fire
  # chips
   boxplot(OWH_Chips ~ fire_intersect, data = ca_grm_travel_fire)
    # anova
        res_aov <- aov(OWH_Chips ~ fire_intersect,
        data = ca_grm_travel_fire
      )
      
        summary(res_aov) # p = 5.88e-10
    # means
    ca_grm_travel_fire %>%
    group_by(fire_intersect) %>%
    summarise(mean_OWH_Chips = mean(OWH_Chips, na.rm = TRUE),
              n = n()) # no fire =	1.613747 hours; fire = 	1.734607 hours	
    
  # merch
    boxplot(OWH_Merchantable ~ fire_intersect, data = ca_grm_travel_fire)
    # anova
        res_aov <- aov(OWH_Merchantable ~ fire_intersect,
        data = ca_grm_travel_fire
      )
      
        summary(res_aov) # p = 0.000751
  
      

# plot treatment
  # chips
    plot(OWH_Chips ~ TRTCD1, data = ca_grm_travel)
      # anova
        res_aov <- aov(OWH_Chips ~ TRTCD1,
        data = ca_grm_travel
      )
      
        summary(res_aov) # p<2e-16
        
        # dunnett's
        post_test <- glht(res_aov,
      linfct = mcp(TRTCD1 = "Dunnett")
    )
    
        summary(post_test) # for cutting p <1e-06 
    
  # merch
    plot(OWH_Merchantable ~ TRTCD1, data = ca_grm_travel)
  # anova
    res_aov <- aov(OWH_Merchantable ~ TRTCD1,
    data = ca_grm_travel
  )
  
    summary(res_aov) # p<2e-16
    
    # dunnett's
    post_test <- glht(res_aov,
  linfct = mcp(TRTCD1 = "Dunnett")
)

    summary(post_test) # for cutting p <1e-06 



# removal
lm1_remv_chips <- lm(REMV_CARB_AG_ACRE ~ OWH_Chips, data = ca_grm_travel)
summary(lm1_remv_chips) # Adjusted R-squared:  0.009079

lmt5_remv_chips <- lm(REMV_CARB_AG_ACRE ~ OWH_Chips + OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD, data = ca_grm_travel)
summary(lmt5_remv_chips) # Adjusted R-squared:  0.2447

lmt1_remv_merch <- lm(REMV_CARB_AG_ACRE ~ OWH_Merchantable, data = ca_grm_travel)
summary(lmt1_remv_merch) # Adjusted R-squared:  0.007083


lmt5_remv_merch <- lm(REMV_CARB_AG_ACRE ~ OWH_Merchantable + OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD, data = ca_grm_travel)
summary(lmt5_remv_merch) # Adjusted R-squared:  0.2397

lmtest <- lm(CARBON_DOWN_DEAD ~ OWH_Chips + OWNGRPCD, data = ca_grm_travel)
summary(lmtest)

lmtest <- lm(fire_intersect ~ OWH_Chips, data = ca_grm_travel_fire)


```
# statistical tests for mill distance write-up
Q1) Does distance from the nearest mill affect the probability of fire?
```{r}
# Fire Perimeters

  # Merchantable
    q1M <- glm(fire_intersect ~ OWH_Merchantable, data = ca_grm_travel_fire, family = "binomial")
    summary(q1M)
      # for every additional hour in distance to a Merchantable mill, the log odds of fire increase by 0.09028 (P = 0.000762)
    
    
  # Chips
    
    q1C <- glm(fire_intersect ~ OWH_Chips, data = ca_grm_travel_fire, family = "binomial")
    summary(q1C)
      # for every additional hour in distance to a Chips mill, the log odds of fire increase by 0.15161 (P = 6.79e-10)
    
#DSTRBCD
  # first duplicate the data so we don't mess it up
  ca_grm_travel_fiaFire <- ca_grm_travel_fire
    
  # clean data so fire is a binary variable
  ca_grm_travel_fiaFire_grp <- ca_grm_travel_fiaFire %>%
  mutate(
    DSTRBCD_FIRE = if_else(
      as.numeric(as.character(DSTRBCD1)) %in% 30:32 |
      as.numeric(as.character(DSTRBCD2)) %in% 30:32 |
      as.numeric(as.character(DSTRBCD3)) %in% 30:32,
      1, 0
    )
  )
  
  # now we can analyze
  
    # Merchantable
      q1Mfia <- glm(DSTRBCD_FIRE ~ OWH_Merchantable, data = ca_grm_travel_fiaFire_grp, family = "binomial")
      summary(q1Mfia)
        # for every additional hour in distance to a Merchantable mill, the log odds of fire increase by 0.18472 (P = 4.46e-05)
          # why is this twice the fire perimeter one? 
      
      
    # Chips
      
      q1Cfia <- glm(DSTRBCD_FIRE ~ OWH_Chips, data = ca_grm_travel_fiaFire_grp, family = "binomial")
      summary(q1Cfia)
        # for every additional hour in distance to a Chips mill, the log odds of fire increase by 0.20745 (P = 5.46e-07)
  
  
  
  
```
Q2) Does distance increase the probability of fire because there’s less management happening farther from mills (TRTCD1, Removals)
```{r}
# clean data for TRTCD = 10 (cutting) or 20 (Site preparation - Clearing, slash burning, chopping, disking, bedding, or other practices clearly intended to prepare a site for either natural or artificial regeneration.)
  ca_grm_travel_trtcd_grp <- ca_grm_travel_fiaFire %>%
  mutate(
    TRTCD_CUT_PREP = if_else(
      as.numeric(as.character(TRTCD1)) %in% 10:20 |
      as.numeric(as.character(TRTCD2)) %in% 10:20 |
      as.numeric(as.character(TRTCD3)) %in% 10:20,
      1, 0
    )
  )

# do the same for the fire dataset so we can analyze trtmnt and fire together: clean data for TRTCD = 10 (cutting) or 20 (Site preparation - Clearing, slash burning, chopping, disking, bedding, or other practices clearly intended to prepare a site for either natural or artificial regeneration.)
  ca_grm_travel_trtcd_grp <- ca_grm_travel_fiaFire_grp %>%
  mutate(
    TRTCD_CUT_PREP = if_else(
      as.numeric(as.character(TRTCD1)) %in% 10:20 |
      as.numeric(as.character(TRTCD2)) %in% 10:20 |
      as.numeric(as.character(TRTCD3)) %in% 10:20,
      1, 0
    )
  )

# is there less management farther from mills?
  
  # Merchantable
  q2aM <- glm(TRTCD_CUT_PREP ~ OWH_Merchantable, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2aM)
    # for every additional hour in distance to a Merchantable mill, the log odds of treatment decrease by 0.56457 (P <2e-16)
  
  # Chips
  q2aC <- glm(TRTCD_CUT_PREP ~ OWH_Chips, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2aC)
    # for every additional hour in distance to a Chip mill, the log odds of treatment decrease by 0.57778 (P <2e-16)

# is there more fire when there is less management, controlling for distance?

  # Merchantable
  q2bM <- glm(fire_intersect ~ OWH_Merchantable + TRTCD_CUT_PREP, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2bM)
    # for every additional hour in distance to a Merchantable mill, the log odds of fire increase by 0.08316 (P = 0.00202). This is down from 0.09028 when treatment is not considered, which means that treatment mediates the effect of distance on fire. The log odds of fire decrease by 0.25618 if there is treatment vs no treatment (p = 0.00519) 
  
  # Chips
  q2bC <- glm(fire_intersect ~ OWH_Chips + TRTCD_CUT_PREP, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2bC)
    # for every additional hour in distance to a Chip mill, the log odds of fire increase by 0.14422 (P = 5.3e-09). This is down from 0.1516 when treatment is not considered, meaning that treatment mediates the effect of distance on fire. The log odds of fire decrease by 0.25067 if there is treatment (p = 0.00534). 
  
  
# Mediation analysis: does management effect on fire interact with mill distance?
  
  # Merchantable
  med_outM <- mediate(q2aM, q2bM,
                     treat = "OWH_Merchantable",
                     mediator = "TRTCD_CUT_PREP",
                     boot = TRUE, sims = 1000)
  
  summary(med_outM)
    # Prop. Mediated (average) 0.132741 (P = 0.01), so 13% of the effect of distance from merchantable mill on fire is explained by reduced management
  
  #Chips
  med_outC <- mediate(q2aC, q2bC,
                     treat = "OWH_Chips",
                     mediator = "TRTCD_CUT_PREP",
                     boot = TRUE, sims = 1000)
  
  summary(med_outC)
    # Prop. Mediated (average) 0.08860 (P = 0.002), so 8.9% of the effect of distance from chip mill on fire is explained by reduced management
  
```

Q3) Does tree mortality increase as distance from mills increases?
```{r}
# Merchantable
q3aM <- lm(MORT_CARB_AG_ACRE ~ OWH_Merchantable, data = ca_grm_travel_fire)
summary(q3aM)

#Chips
q3aC <- lm(MORT_CARB_AG_ACRE ~ OWH_Chips, data = ca_grm_travel_fire)
summary(q3aC)
```

Q4) Does removal decrease with distance from mill?
```{r}
#Merchantable
q3bM <- lm(REMV_CARB_AG_ACRE ~ OWH_Merchantable, data = ca_grm_travel_fire)
summary(q3bM)
  # coefficient = -153.92, p = 8e-13 

#Chips
q3bC <- lm(REMV_CARB_AG_ACRE ~ OWH_Chips, data = ca_grm_travel_fire)
summary(q3bC)
  # coefficient = -164.52, p = 2.59e-16 

```
Q4: Does more carbon removed increase probability of fire?
```{r}
q4a <- glm(fire_intersect ~ REMV_CARB_AG_ACRE, data = ca_grm_travel_trtcd_grp, family = "binomial")
summary(q4a)
  # coefficient = -3.206e-05, p-value = 0.0473

```

# same Qs but with fire perim database focused on wildfire
```{r}
#find layers in gdb
fire_path <- "/Users/alyssajain/FIA/fire24_1.gdb"
st_layers(fire_path)

# get wildfire only
wildfire <- st_read(fire_path, layer = "firep24_1") 

# Reproject wildfire data to match FIA CRS
wildfire <- st_transform(wildfire, st_crs(fia_sf))

# get rid of multisurface
wildfire_clean <- wildfire %>%
  filter(st_geometry_type(st_geometry(.)) %in% c("POLYGON", "MULTIPOLYGON"))

# fix invalid geometries
wildfire_geom <- st_geometry(wildfire_clean)

invalid_idx <- which(!st_is_valid(wildfire_geom))

if(length(invalid_idx) > 0){
  fixed_geoms <- st_make_valid(wildfire_geom[invalid_idx])
  wildfire_geom[invalid_idx] <- fixed_geoms
  st_geometry(wildfire_clean) <- wildfire_geom
}

# Now test validity again
sum(!st_is_valid(wildfire_clean))  # Should be 0 or very low
  
# This returns a list: one element per plot, containing indices of intersecting fire polygons
wildfire_intersections <- st_intersects(fia_sf, wildfire_clean, sparse = TRUE)

# Check if each plot has any intersections (i.e., non-empty list element)
fia_sf$wildfire_intersect <- lengths(wildfire_intersections) > 0

# Convert to 1/0 instead of TRUE/FALSE 
fia_sf$wildfire_intersect <- as.integer(fia_sf$wildfire_intersect)

# now we will join to ca_grm_travel

  # First extract just the plot ID and fire_intersect column from the sf object
  wildfire_flag_df <- fia_sf %>%
  sf::st_drop_geometry() %>%
  dplyr::select(PLOT, wildfire_intersect) 
  
  # rename PLOT to match with ca_grm PLOT_CD
  names(wildfire_flag_df)[names(wildfire_flag_df) == "PLOT"] <- "PLOT_CD"

  # make PLOT_CD a character
  wildfire_flag_df$PLOT_CD <- as.character(wildfire_flag_df$PLOT_CD)
  
  # make sure each plot appears only once
  wildfire_flag_unique <- wildfire_flag_df %>%
  group_by(PLOT_CD) %>%
  summarise(wildfire_intersect = max(wildfire_intersect, na.rm = TRUE), .groups = "drop")
  
  # now join
  ca_grm_travel_wildfire <- ca_grm_travel %>%
  dplyr::left_join(wildfire_flag_unique, by = "PLOT_CD")
  
  # make fire_intersect a factor
  ca_grm_travel_wildfire$wildfire_intersect <- as.factor(ca_grm_travel_wildfire$wildfire_intersect)
```  
# now analyze if distance to mill causes more fire
```{r}
  # Merchantable
    q1M <- glm(wildfire_intersect ~ OWH_Merchantable, data = ca_grm_travel_wildfire, family = "binomial")
    summary(q1M)
      # for every additional hour in distance to a Merchantable mill, the log odds of fire increase by 0.08274 (P = 0.00201)
exp(0.08274)

    # plot residuals
    res = simulateResiduals(q1M)
    plot(res)
      
  # Chips
    
    q1C <- glm(wildfire_intersect ~ OWH_Chips, data = ca_grm_travel_wildfire, family = "binomial")
    summary(q1C)
      # for every additional hour in distance to a Chips mill, the log odds of fire increase by 0.15013 (P = 9.52e-10)
    
    # plot residuals
    res = simulateResiduals(q1C)
    plot(res)
      


```
# now analyze stocking density
```{r}
# Merchantable

# Make sure stocking density is treated as an ordered factor
ca_grm_travel_wildfire$GSSTKCD <- factor(
  ca_grm_travel_wildfire$GSSTKCD,
  levels = c(1, 2, 3, 4, 5),
  ordered = TRUE
)

# Fit ordinal logistic regression
stockingM <- polr(GSSTKCD ~ OWH_Merchantable,
                  data = ca_grm_travel_wildfire,
                  Hess = TRUE)

# Summary
summary(stockingM)

# Get p-values for coefficients
(ctable <- coef(summary(stockingM)))
p_values <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
p_values

# McFadden's pseudo-R² for model fit
null_logLik <- logLik(polr(GSSTKCD ~ 1, 
                           data = ca_grm_travel_wildfire, Hess = TRUE))
model_logLik <- logLik(stockingM)

pseudoR2 <- 1 - (as.numeric(model_logLik) / as.numeric(null_logLik))
pseudoR2

# plot predicted probabilities

# Generate new data across the observed range of OWH_Merchantable
new_data <- data.frame(
  OWH_Merchantable = seq(
    min(ca_grm_travel_wildfire$OWH_Merchantable, na.rm = TRUE),
    max(ca_grm_travel_wildfire$OWH_Merchantable, na.rm = TRUE),
    length.out = 100
  )
)

# Predict probabilities for each stocking density category
pred_probs <- predict(stockingM, newdata = new_data, type = "probs")

# Convert to long format for ggplot
pred_df <- cbind(new_data, pred_probs)
pred_df_long <- tidyr::pivot_longer(pred_df,
                                    cols = starts_with("1"):starts_with("5"),
                                    names_to = "GSSTKCD",
                                    values_to = "Probability")

# Plot
ggplot(pred_df_long, aes(x = OWH_Merchantable, y = Probability,
                         color = GSSTKCD)) +
  geom_line(size = 1.2) +
  labs(
    title = "Predicted Probability of Stocking Density by Distance to Mill",
    x = "Distance to Mill (OWH_Merchantable)",
    y = "Predicted Probability",
    color = "Stocking Density"
  ) +
  theme_minimal(base_size = 14)

# Chips

# Make sure stocking density is treated as an ordered factor
ca_grm_travel_wildfire$GSSTKCD <- factor(
  ca_grm_travel_wildfire$GSSTKCD,
  levels = c(1, 2, 3, 4, 5),
  ordered = TRUE
)

# Fit ordinal logistic regression
stockingC <- polr(GSSTKCD ~ OWH_Chips,
                  data = ca_grm_travel_wildfire,
                  Hess = TRUE)

# Summary
summary(stockingC)

# Get p-values for coefficients
(ctable <- coef(summary(stockingC)))
p_values <- pnorm(abs(ctable[, "t value"]), lower.tail = FALSE) * 2
p_values

# McFadden's pseudo-R² for model fit
null_logLik <- logLik(polr(GSSTKCD ~ 1, 
                           data = ca_grm_travel_wildfire, Hess = TRUE))
model_logLik <- logLik(stockingC)

pseudoR2 <- 1 - (as.numeric(model_logLik) / as.numeric(null_logLik))
pseudoR2

# plot predicted probabilities

# Generate new data across the observed range of OWH_Merchantable
new_data <- data.frame(
  OWH_Merchantable = seq(
    min(ca_grm_travel_wildfire$OWH_Chips, na.rm = TRUE),
    max(ca_grm_travel_wildfire$OWH_Chips, na.rm = TRUE),
    length.out = 100
  )
)

# Predict probabilities for each stocking density category
pred_probs <- predict(stockingC, newdata = new_data, type = "probs")

# Convert to long format for ggplot
pred_df <- cbind(new_data, pred_probs)
pred_df_long <- tidyr::pivot_longer(pred_df,
                                    cols = starts_with("1"):starts_with("5"),
                                    names_to = "GSSTKCD",
                                    values_to = "Probability")

# Plot
ggplot(pred_df_long, aes(x = OWH_Chips, y = Probability,
                         color = GSSTKCD)) +
  geom_line(size = 1.2) +
  labs(
    title = "Predicted Probability of Stocking Density by Distance to Mill",
    x = "Distance to Mill (OWH_Chips)",
    y = "Predicted Probability",
    color = "Stocking Density"
  ) +
  theme_minimal(base_size = 14)
```
# now analyze with bins
```{r}
library(sf)
library(dplyr)
library(ggplot2)

# preview data: look at distance distribution
hist(ca_grm_travel_wildfire$OWH_Chips, breaks = 50)
hist(ca_grm_travel_wildfire$OWH_Merchantable, breaks = 50)

summary(ca_grm_travel_wildfire$OWH_Chips)
summary(ca_grm_travel_wildfire$OWH_Merchantable)

## CHIPS
# Step 1: Create distance bins 
ca_grm_travel_wildfire_chips_bins <- ca_grm_travel_wildfire %>%
  filter(!is.na(OWH_Chips)) %>%
  mutate(
    distance_chips_bin = cut(
      OWH_Chips,
      breaks = c(0, 1, 2, 3, 4, 5),
      include.lowest = TRUE,
      right = FALSE,  # e.g., [0,1), [1,2), etc.
      labels = c("0–1 hr", "1–2 hr", "2–3 hr", "3–4 hr", "4–5 hr")
    )
  )
# Step 2: Spatial join to determine fire presence
fia_with_fire <- fia_sf %>%
  mutate(plot_id = row_number()) %>%
  st_join(wildfire_clean %>% select(geometry), join = st_intersects, left = TRUE) %>%
  mutate(fire_present = !is.na(st_geometry(.)))

# Step 3: Summarize fire probability by distance bin
fire_prob_by_chips_bin <- ca_grm_travel_wildfire_chips_bins %>%
  st_drop_geometry() %>%
  group_by(distance_chips_bin) %>%
  summarise(
    n_total = n(),
    n_fire = sum(as.numeric(wildfire_intersect), na.rm = TRUE),
    fire_probability = n_fire / n_total
  )

# Step 4: Plot the results
ggplot(fire_prob_by_chips_bin, aes(x = distance_chips_bin, y = fire_probability)) +
  geom_col(fill = "firebrick") +
  labs(
    title = "Probability of Fire by Distance to Chips Mill",
    x = "Distance to Mill (binned)",
    y = "Probability of Fire"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Step 5: Logistic Regression
  # Ensure distance_bin is a factor
  ca_grm_travel_wildfire_chips_bins$distance_chips_bin <- factor(ca_grm_travel_wildfire_chips_bins$distance_chips_bin, ordered = TRUE)
  
  # Fit logistic regression with distance bins
  logit_bin_model <- glm(wildfire_intersect ~ distance_chips_bin, data = ca_grm_travel_wildfire_chips_bins, family = binomial())
  
  # Model summary
  summary(logit_bin_model)

## MERCH
# Step 1: Create distance bins 
ca_grm_travel_wildfire_merch_bins <- ca_grm_travel_wildfire %>%
  filter(!is.na(OWH_Merchantable)) %>%
  mutate(
    distance_merch_bin = cut(
      OWH_Merchantable,
      breaks = c(0, 1, 2, 3, 4, 5),
      include.lowest = TRUE,
      right = FALSE,  # e.g., [0,1), [1,2), etc.
      labels = c("0–1 hr", "1–2 hr", "2–3 hr", "3–4 hr", "4–5 hr")
    )
  )

# Step 3: Summarize fire probability by distance bin
fire_prob_by_merch_bin <- ca_grm_travel_wildfire_merch_bins %>%
  st_drop_geometry() %>%
  group_by(distance_merch_bin) %>%
  summarise(
    n_total = n(),
    n_fire = sum(as.numeric(wildfire_intersect), na.rm = TRUE),
    fire_probability = n_fire / n_total
  )

# Step 4: Plot the results
ggplot(fire_prob_by_merch_bin, aes(x = distance_merch_bin, y = fire_probability)) +
  geom_col(fill = "firebrick") +
  labs(
    title = "Probability of Fire by Distance to Merchantable Mill",
    x = "Distance to Mill (binned)",
    y = "Probability of Fire"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Step 5: Logistic Regression
  # Ensure distance_bin is a factor
  ca_grm_travel_wildfire_merch_bins$distance_merch_bin <- factor(ca_grm_travel_wildfire_merch_bins$distance_merch_bin, ordered = TRUE)
  
  # Fit logistic regression with distance bins
  logit_bin_model <- glm(wildfire_intersect ~ distance_merch_bin, data = ca_grm_travel_wildfire_merch_bins, family = binomial())
  
  # Model summary
  summary(logit_bin_model)
  
# Step 6: non linear model
  # GAM without bins chips
  model_gam_chips <- gam(wildfire_intersect ~ s(OWH_Chips), data = ca_grm_travel_wildfire, family = binomial)
  
  plot(model_gam_chips)
  
    # switch to probability
    # Keep only complete rows for the distance column
    valid_rows <- complete.cases(ca_grm_travel_wildfire$OWH_Chips)
    
    # Filter both x and y to the same valid rows
    x_vals <- ca_grm_travel_wildfire$OWH_Chips[valid_rows]
    predicted_prob_chips <- predict(model_gam_chips, newdata = ca_grm_travel_wildfire[valid_rows, ], type = "response")
    
    plot(x_vals, predicted_prob_chips,
         xlab = "Travel Time to Mill (hrs)",
         ylab = "Predicted Wildfire Probability",
         main = "Wildfire Risk vs. Distance to Mill")
    
    # GAM without bins merch
  model_gam_merch <- gam(wildfire_intersect ~ s(OWH_Merchantable), data = ca_grm_travel_wildfire, family = binomial)
  
  plot(model_gam_merch)
  
    # switch to probability
    # Keep only complete rows for the distance column
    valid_rows <- complete.cases(ca_grm_travel_wildfire$OWH_Merchantable)
    
    # Filter both x and y to the same valid rows
    x_vals <- ca_grm_travel_wildfire$OWH_Merchantable[valid_rows]
    predicted_prob_merch <- predict(model_gam_merch, newdata = ca_grm_travel_wildfire[valid_rows, ], type = "response")
    
    plot(x_vals, predicted_prob_merch,
         xlab = "Travel Time to Mill (hrs)",
         ylab = "Predicted Wildfire Probability",
         main = "Wildfire Risk vs. Distance to Mill")
    
# Split by fuel load
library(mgcv)
library(dplyr)

# Get unique fuel types (remove NA and blanks)
fuel_types <- ca_grm_travel_wildfire %>%
  filter(!is.na(DWM_FUELBED_TYPCD), DWM_FUELBED_TYPCD != "") %>%
  distinct(DWM_FUELBED_TYPCD) %>%
  pull()

# Create empty summary table
gam_summaries <- data.frame(
  fuel_type = character(),
  p_value = numeric(),
  deviance_explained = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each fuel type
for (fuel in fuel_types) {
  df_sub <- ca_grm_travel_wildfire %>%
    filter(DWM_FUELBED_TYPCD == fuel)

  if (nrow(df_sub) >= 30 && length(unique(df_sub$OWH_Chips)) >= 5) {
    model <- tryCatch(
      gam(wildfire_intersect ~ s(OWH_Chips, k = 10), data = df_sub, 
          family = binomial, method = "REML"),
      error = function(e) NULL
    )

    if (!is.null(model)) {
      s <- summary(model)
      # Check if smooth term exists
      if (!is.null(s$s.table)) {
        gam_summaries <- rbind(gam_summaries, data.frame(
          fuel_type = fuel,
          p_value = s$s.table[1, "p-value"],
          deviance_explained = s$dev.expl
        ))
      }
    }
  }
}

# View result
print(gam_summaries)

### working code  
  # Create empty list to collect data
prediction_list <- list()

  # Get all unique fuel types (excluding NA and empty string)
  fuel_types <- ca_grm_travel_wildfire %>%
    filter(!is.na(DWM_FUELBED_TYPCD), DWM_FUELBED_TYPCD != "") %>%
    distinct(DWM_FUELBED_TYPCD) %>%
    pull()

  # Loop over fuel types
  for (fuel in fuel_types) {
    df_sub <- ca_grm_travel_wildfire %>% filter(DWM_FUELBED_TYPCD == fuel)

  # Only run if we have enough data
  if (nrow(df_sub) >= 30 && length(unique(df_sub$OWH_Chips)) >= 5) {
    # Fit GAM with smoothing spline
    model <- gam(wildfire_intersect ~ s(OWH_Chips, k = 5), data = df_sub, family = binomial)

    # Generate new travel time values
    new_df <- data.frame(OWH_Chips = seq(min(df_sub$OWH_Chips, na.rm = TRUE),
                                         max(df_sub$OWH_Chips, na.rm = TRUE), length.out = 100))
    
    # Predict probability
    new_df$probability <- predict(model, newdata = new_df, type = "response")
    new_df$fuel_type <- fuel

    # Add to list
    prediction_list[[fuel]] <- new_df
  }
}

    # Combine all predictions
    prediction_df <- bind_rows(prediction_list)
    
    # plot all
    ggplot(prediction_df, aes(x = OWH_Chips, y = probability)) +
  geom_line(color = "#D95F02", size = 1) +
  facet_wrap(~ fuel_type, scales = "free_y") +
  labs(
    title = "Predicted Wildfire Probability by Travel Time and Fuel Type",
    x = "Travel Time to Mill (hours)",
    y = "Predicted Probability of Wildfire"
  ) +
  theme_minimal(base_size = 12)

  # extract p-value and edf
  
    # Create a dataframe to store summary stats
    gam_summaries <- data.frame(
      fuel_type = character(),
      edf = numeric(),
      F_value = numeric(),
      p_value = numeric(),
      deviance_explained = numeric(),
      stringsAsFactors = FALSE
    )
    
    # Loop over each fuel type and fit a GAM
    for (fuel in fuel_types) {
      df_sub <- ca_grm_travel_wildfire %>%
        filter(DWM_FUELBED_TYPCD == fuel)
    
      if (nrow(df_sub) > 10 && length(unique(df_sub$OWH_Chips)) > 3) {
        model <- gam(wildfire_intersect ~ s(OWH_Chips), data = df_sub, family = binomial)
        model_summary <- summary(model)
    
      # protect from small sample size
      safe_summary <- function(model) {
      s <- tryCatch(summary(model), error = function(e) NULL)
      if (!is.null(s) && !is.null(s$s.table)) {
        return(s$s.table)
      } else {
        return("No smooth terms or model failed.")
      }
    }
    
    # Example for a list of models
    lapply(model, safe_summary)

        
        gam_summaries <- rbind(gam_summaries, data.frame(
          fuel_type = fuel,
          edf = model_summary$s.table[1, "edf"],
          F_value = model_summary$s.table[1, "F"],
          p_value = model_summary$s.table[1, "p-value"],
          deviance_explained = model_summary$dev.expl
        ))
      }
    }
  
  
    
    
  
  
        
``` 




# now analyze # of fires per pixel 
```{r}
# Step 1: Ensure wildfire has unique fire ID
wildfire_vect <- vect(wildfire_clean)

# Step 2: Template raster
template <- rast(ext(wildfire_vect), resolution = 1000, crs = crs(wildfire_vect))

# Step 3: Create output raster with 0s
fire_count <- rast(template)
values(fire_count) <- 0

# Step 4: Loop over unique fire IDs
unique_ids <- unique(wildfire$FIRE_NAME)
pb <- txtProgressBar(min = 0, max = length(unique_ids), style = 3)

for (i in seq_along(unique_ids)) {
  fid <- unique_ids[i]
  
  # Subset to one fire (could be multipolygon)
  fire_i <- wildfire_vect[wildfire_vect$FIRE_NAME == fid, ]

  # Rasterize: presence = 1 where this fire burned
  r_fire <- tryCatch({
    rasterize(fire_i, template, field = 1, background = 0)
  }, error = function(e) NULL)
  
  if (!is.null(r_fire)) {
    # Add fire presence to cumulative count
    fire_count <- fire_count + (r_fire > 0)
  }

  setTxtProgressBar(pb, i)
}
close(pb)

# Save result
writeRaster(fire_count, "unique_wildfires_per_pixel.tif", overwrite = TRUE)

# Plot
plot(fire_count, main = "Unique Fires per Pixel")

# Check if FIRE_NAME exists and is unique per fire
length(unique(wildfire_vect$FIRE_NAME))   # Should be ~22,000
nrow(wildfire_vect)                     # Might be >> 22,000 if multiple polygons per fire


# 1. Validate and fix geometries
# Create a copy to work with
wildfire_valid <- wildfire_vect

# Identify invalid geometries
invalid_idx <- which(!is.valid(wildfire_valid))

# Fix only invalid ones
wildfire_valid[invalid_idx] <- makeValid(wildfire_valid[invalid_idx])

# Optional: confirm all are valid now
all(is.valid(wildfire_valid))  # should return TRUE


# Count total geometries
total <- nrow(wildfire_vect)

# Identify invalid ones
invalid_idx <- which(!is.valid(wildfire_vect))

# Count invalid geometries
num_invalid <- length(invalid_idx)

# Report
cat("Total features:", total, "\n")
cat("Invalid geometries:", num_invalid, "\n")
cat("Percent invalid:", round(100 * num_invalid / total, 2), "%\n")

# Drop invalid geometries
wildfire_valid <- wildfire_valid[is.valid(wildfire_valid), ]

# Confirm
all(is.valid(wildfire_valid))  # should now return TRUE

# Group by FIRE_ID: each fire becomes one geometry
wildfire_dissolved <- aggregate(wildfire_valid, by = "FIRE_NAME", fun = "first")  # or use dplyr::first if needed

# Confirm correct count
n_fires <- nrow(wildfire_dissolved)         # Should match length of unique fire IDs
length(unique(wildfire_dissolved$FIRE_NAME))  # Should be the same

# Create template raster
template <- rast(ext(wildfire_dissolved), resolution = 1000, crs = crs(wildfire_dissolved))

# Initialize output raster
fire_count <- rast(template)
values(fire_count) <- 0

# Loop through each fire
pb <- txtProgressBar(min = 0, max = n_fires, style = 3)

for (i in seq_len(n_fires)) {
  fire_geom <- wildfire_dissolved[i, ]

  r_fire <- tryCatch({
    rasterize(fire_geom, template, field = 1, background = 0)
  }, error = function(e) NULL)

  if (!is.null(r_fire)) {
    fire_count <- fire_count + (r_fire > 0)
  }

  setTxtProgressBar(pb, i)
}
close(pb)

#inspect values
hist(values(fire_count), breaks = 20, main = "Histogram: Unique Fires per Pixel")
summary(values(fire_count))
plot(fire_count)

# take 3
library(terra)

# Step 1: Read fire shapefile
fire <- vect(wildfire_clean)

# Step 2: Create a unique fire ID (optional: adjust if needed)
fire$fire_id <- paste0(fire$FIRE_NAME)

  # Drop invalid geometries
  fire_valid <- fire[is.valid(fire), ]

# Step 3: Dissolve fire polygons by unique fire ID
fire_dissolved <- aggregate(fire_valid, by = "fire_id")

  # Remove invalids
  fire_dissolved <- fire_dissolved[is.valid(fire_dissolved), ]
  
  # Also remove empty ones
  fire_dissolved <- fire_dissolved[!is.empty(fire_dissolved), ]

# Step 4: Create an empty raster (same extent/res as your analysis raster or base map)
template <- rast(res = 30, extent = ext(fire), crs = crs(fire))  # example with 30m resolution

# Step 5: Rasterize each fire individually
# We'll loop through fires and mark presence (1) on each raster

raster_list <- list()

for (i in 1:nrow(fire_dissolved)) {
  this_fire <- fire_dissolved[i, ]
  r <- rasterize(this_fire, template, field = 1)
  raster_list[[i]] <- r
}

# Step 6: Stack and sum (clipping values >1 to 1, then summing)
# This prevents multiple polygons from the same fire being counted more than once per pixel

fire_stack <- rast(raster_list)
fire_binary <- clamp(fire_stack, lower = 0, upper = 1)  # convert all to 0/1
fire_count <- app(fire_binary, fun = sum, na.rm = TRUE)

# Step 7: Check values
hist(values(fire_count), breaks = 20, main = "Histogram: Unique Fires per Pixel")
summary(values(fire_count))



# bug fixes
    # Inspect one fire polygon
plot(fire_dissolved[1, ], col = "red", main = "Single Fire")

# Plot raster grid on top
plot(template, add = TRUE, border = "gray")

# check geom
geom_type <- geomtype(fire_dissolved[1, ])
print(geom_type)  # Should say "polygons" or "multipolygons"

# Area of geometry
expanse(fire_dissolved[1, ])  # Should be > 0

# buffer small polygons
fire_dissolved_buffered <- buffer(fire_dissolved, width = 15)  # buffer by half a cell

# Test again
r_test <- rasterize(fire_dissolved[1, ], template, field = 1)
plot(r_test)
unique(values(r_test))  # Should include 1s now

# Check geometry is not empty and has area
fire_geom <- fire_dissolved[1, ]
print(geomtype(fire_geom))         # Should be "polygons" or "multipolygons"
print(expanse(fire_geom))          # Should be > 0
print(st_is_empty(st_geometry(fire_geom)))  # Should be FALSE

# recreate minimal raster
  # Build a very tight raster around that fire polygon
  tight_raster <- rast(ext(fire_geom), res = 30, crs = crs(fire_geom))
  
  # Now try rasterizing
  r_test <- rasterize(fire_geom, tight_raster, field = 1)
  plot(r_test)
  unique(values(r_test))


```
# cursor
```{r}
# fix invalid geometries
wildfire_valid <- st_make_valid(wildfire_clean)

# Dissolve all polygons belonging to the same fire
fires_dissolved <- wildfire_valid %>%
  group_by_at(FIRE_NAME) %>%
  summarise(geometry = st_union(geometry), .groups = "drop")

```

#######################
#######################



## carbon
Carbon
```{r}
ca_cond <- ca_data$COND #cond table
ca_plot <- ca_data$PLOT #plot table
ca_tree <- ca_data$TREE #tree table

# get carbon by plot
carbon_plot<- carbon(ca_data, byPlot = TRUE)

# join
ca_table <- carbon_plot %>%
  # Join tables
  left_join(ca_cond, by = 'PLT_CN') # The foreign key for the COND table is PLT_CN. There is always a match of the PLT_CN value to the CN value in the PLOT table.

# make variables factors
ca_table$OWNGRPCD <- as.factor(ca_table$OWNGRPCD)
ca_table$FORTYPCD <- as.factor(ca_table$FORTYPCD)
ca_table$TRTCD1 <- as.factor(ca_table$TRTCD1)
ca_table$SITECLCD <- as.factor(ca_table$SITECLCD)
ca_table$PHYSCLCD <- as.factor(ca_table$PHYSCLCD)
ca_table$RESERVCD <- as.factor(ca_table$RESERVCD)


# carbon per acre estimates

lm1 <- lm(CARB_ACRE ~ OWNGRPCD, data = ca_table)
summary(lm1) # Adjusted R-squared:  0.0181 

lm2 <- lm(CARB_ACRE ~ OWNGRPCD + FORTYPCD, data = ca_table)
summary(lm2) # Adjusted R-squared:  0.1025 

lm3 <- lm(CARB_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1, data = ca_table)
summary(lm3) # Adjusted R-squared:  0.1028

lm4 <- lm(CARB_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD, data = ca_table)
summary(lm4) # Adjusted R-squared:  0.1152

lm5 <- lm(CARB_ACRE ~ OWNGRPCD + FORTYPCD + TRTCD1 + SITECLCD + PHYSCLCD, data = ca_table)
summary(lm5) # Adjusted R-squared:  0.1221

```


analyze linear regressions
```{r}
result <- AIC(lm1, lm2, lm3, lm4, lm5)

models <- list(lm1, lm2, lm3, lm4, lm5)
result$BIC <- sapply(models, BIC) 
model_summary <- lapply(models, summary)

for(i in 1:length(models)){ #this creates a variable i that starts with the value i=1
  result$rsq[i] <- model_summary[[i]]$r.squared #we assign the rsq value from model i to the i'th row of the column 'rsq' in the table 'results'
  result$adj_rsq[i] <- model_summary[[i]]$adj.r.squared #same for adjusted rsq
} #now we go back to the beginning of the for-loop, add 1 to the value of i, and do everything again

kable(result, digits = 2, align = "c")


```

exploration
```{r}
# what are fuel loads like?
ca_cond$DWM_FUELBED_TYPCD <- as.factor(ca_cond$DWM_FUELBED_TYPCD) # factorize
summary(ca_cond$DWM_FUELBED_TYPCD) # summarize fuel load by type

# check if fuel loads affect mortality
lm_mort_fuel <- lm(MORT_CARB_AG_ACRE ~ DWM_FUELBED_TYPCD, data = ca_grm_table) 
summary(lm_mort_fuel) # Adjusted R-squared: 0.04507

lm_mort_fuel_1 <- lm(MORT_CARB_AG_ACRE ~ OWNGRPCD * DSTRBCD1 + OWNGRPCD * FORTYPCD + DSTRBCD2 + DSTRBCD3, data = ca_grm_table) 
summary(lmi_mort1_2) # Adjusted R-squared: 0.3326

# check if fuel loads affect fire. We need to use a glm because fire is not a continuous variable (i think)
  # first factorize
  ca_grm_travel_fire$DWM_FUELBED_TYPCD <- as.factor(ca_grm_travel_fire$DWM_FUELBED_TYPCD)

  # now run glm
  glm_fire_fuel <- glm(fire_intersect ~ DWM_FUELBED_TYPCD, data = ca_grm_travel_fire, family = "binomial")
summary(glm_fire_fuel)

# check if treatment affects fuel loads
  # factorize
  ca_grm_travel_trtcd_grp$DWM_FUELBED_TYPCD <- as.factor(ca_grm_travel_trtcd_grp$DWM_FUELBED_TYPCD)

  #run glm  
  glm_fuel_trt <- glm(DWM_FUELBED_TYPCD ~  TRTCD_CUT_PREP, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(glm_fuel_trt)
  
  #let's try a boxplot to see if different loads have different treatment averages
  boxplot(TRTCD_CUT_PREP ~ DWM_FUELBED_TYPCD, data = ca_grm_travel_trtcd_grp)
    # not helpful, maybe they were treated because of the high load
  
  # get means
  fuel_trtcd_means <- ca_grm_travel_trtcd_grp %>%
    group_by(DWM_FUELBED_TYPCD) %>%
    summarise(Mean_Value = mean(TRTCD_CUT_PREP, na.rm = TRUE))
  
  fuel_trtcd_means
  
  # check if mill distance affects fuel load
  
    # Merchantable
    ca_grm_travel_wildfire$DWM_FUELBED_TYPCD <- as.factor(ca_grm_travel_wildfire$DWM_FUELBED_TYPCD)
    glm_mill_fuel <- multinom(DWM_FUELBED_TYPCD ~ OWH_Merchantable, data = ca_grm_travel_wildfire)
  summary(glm_mill_fuel)
```