---
title: "Wildfire_Mills_Clean"
author: "Alyssa N Jain"
format: html
editor: visual
---

```{r setup, include=FALSE}
#| echo: false
library(rFIA)
library(ggplot2)
library(dplyr)
library(stats)
library(knitr)
library(tidyverse)
library(tidyr)
library(stringr)
library(readr)
library(sf)
library(multcomp)
library(mediation)
library(lme4)
library(broom.mixed)
library(ggeffects)
library(lfe)
library(broom)
library(purrr)
library(nnet)
library(terra)
library(colorspace)
```

# How does distance to mill affect wildfire and management?

**Questions:** We aim to answer the following questions:

*Q1) Does distance from the nearest mill affect the probability of fire?*

*Q2) Does distance increase the probability of fire because there’s less management happening farther from mills?*

**Data Setup**

FIA Data and GRM

```{r}
#| echo: false
# read in FIA data
ca_data <- readFIA("/Volumes/Alyssa_Backup/CA_CSV") 

# identify condition, plot, and tree tables
ca_cond <- ca_data$COND #cond table
ca_plot <- ca_data$PLOT #plot table
ca_tree <- ca_data$TREE #tree table


# get grm by plot
grm_plot <- growMort(ca_data, byPlot = TRUE, stateVar = "CARB_AG")

# join
ca_grm_table <- grm_plot %>%
  # Join tables
  left_join(ca_cond, by = 'PLT_CN') # The foreign key for the COND table is PLT_CN. There is always a match of the PLT_CN value to the CN value in the PLOT table.

# make variables factors
ca_grm_table$OWNGRPCD <- as.factor(ca_grm_table$OWNGRPCD)
ca_grm_table$FORTYPCD <- as.factor(ca_grm_table$FORTYPCD)
ca_grm_table$TRTCD1 <- as.factor(ca_grm_table$TRTCD1)
ca_grm_table$SITECLCD <- as.factor(ca_grm_table$SITECLCD)
ca_grm_table$PHYSCLCD <- as.factor(ca_grm_table$PHYSCLCD)
ca_grm_table$RESERVCD <- as.factor(ca_grm_table$RESERVCD)
ca_grm_table$YEAR <- as.factor(ca_grm_table$YEAR)
ca_grm_table$DSTRBCD1 <- as.factor(ca_grm_table$DSTRBCD1)
ca_grm_table$DSTRBCD2 <- as.factor(ca_grm_table$DSTRBCD2)
ca_grm_table$DSTRBCD3 <- as.factor(ca_grm_table$DSTRBCD3)
```

Mill Travel Time (from Biosum)

```{r}
#| echo: false
# read in mill travel time data
min_travel_time <- read_csv("min_travel_time.csv")

# filter by CA
ca_min_travel_time <- min_travel_time[min_travel_time$STATE == 'CA',] #this leaves some plots with FIPS code 41 (OR), even though STATE = CA
# filter for CA FIPS code (6)
ca_min_travel_time <- min_travel_time[min_travel_time$STATECD == '6',]

# the last 5 digits of pltID in ca_grm_table are PLOT IDs from ca_min_travel time

# make new column in ca_grm_with last 5 digits of pltID
ca_grm_table1 <- ca_grm_table

ca_grm_table1$PLOT_CD <- substring(ca_grm_table$pltID, nchar(ca_grm_table$pltID) - 4)

# now pivot so that time to chips and merchantable facilities are separate columns
ca_min_travel_time_clean <- ca_min_travel_time %>%
  dplyr::select(PLOT, BIOCD_DEF, ONE_WAY_HOURS) %>%
  group_by(PLOT, BIOCD_DEF) %>%
  # summarise(ONE_WAY_HOURS = min(ONE_WAY_HOURS), .groups = "drop") %>%
  pivot_wider(
    names_from = BIOCD_DEF,
    values_from = ONE_WAY_HOURS,
    names_prefix = "OWH_"
  )

# rename PLOT to match with ca_grm PLOT_CD
names(ca_min_travel_time_clean)[names(ca_min_travel_time_clean) == "PLOT"] <- "PLOT_CD"

# make PLOT_CD a character
ca_min_travel_time_clean$PLOT_CD <- as.character(ca_min_travel_time_clean$PLOT_CD)

# join with ca_grm
ca_grm_travel <- ca_grm_table1 %>%
  full_join(ca_min_travel_time_clean, by = 'PLOT_CD')

```

Fire Spatial Data (MTBS)

1.  fire presence

```{r}
# read in spatial fire perimeters
fire_path <- "/Users/alyssajain/Downloads/mtbs_perimeter_data/mtbs_perims_DD.shp"
burns <- st_read(fire_path)

#check crs
st_crs(burns)

# filter for wildfire
wildfire_all <- burns[burns$Incid_Type == "Wildfire", ]

# wildfire_tbl <- as_tibble(wildfire_all)

wildfire_tbl <- wildfire_all %>%
  as_tibble() %>%
  slice_head(n = 10)



# read in State Shapefiles to clip boundaries
states <- st_read("/Users/alyssajain/Downloads/cb_2018_us_state_5m/cb_2018_us_state_5m.shp")
v

#check crs
st_crs(states) # matches

# filter states to CA
ca <- states[states$STUSPS %in% c( "CA"), ]

# Replace invalid geometries with valid ones
wildfire_all <- st_make_valid(wildfire_all)
ca <- st_make_valid(ca)

# clip burns to CA
CA_wildfire <- st_intersection(wildfire_all, ca)

wildfire <- CA_wildfire


# Convert FIA plot data to an sf object 
  # first remove NAs
  ca_plot_clean <- ca_plot %>%
  filter(!is.na(LON) & !is.na(LAT))
  
  #then use coords to convert to sf
  fia_sf <- st_as_sf(ca_plot_clean, coords = c("LON", "LAT"), crs = 4326)
  
# read in CA resource regions
resource_reg <- read_sf("CAResourceAreas/CAResourceAreas.shp")
  head(resource_reg)
    # info: 6 resource areas (ResourceAr)
  # match crs to fia
  resource_reg <- st_transform(resource_reg, st_crs(fia_sf))

  # Dissolve multipolygons into single polygons by region name
  resource_reg_clean <- resource_reg %>%
    group_by(ResourceAr) %>%
    summarise(geometry = st_union(geometry), .groups = "drop") %>%
    st_as_sf()

  # join resource reg to fia data
  fia_with_region <- st_join(fia_sf, resource_reg_clean, left = TRUE)

  
# Reproject wildfire data to match FIA CRS
wildfire <- st_transform(wildfire, st_crs(fia_sf))

# filter year > 2002
wildfire_2002 <- wildfire %>%
  filter(Ig_Date >= as.Date("2002-01-01"))

# This returns a list: one element per plot, containing indices of intersecting fire polygons
wildfire_intersections <- st_intersects(fia_with_region, wildfire_2002, sparse = TRUE)

# Check if each plot has any intersections (i.e., non-empty list element)
fia_with_region$wildfire_intersect <- lengths(wildfire_intersections) > 0

# Convert to 1/0 instead of TRUE/FALSE 
fia_with_region$wildfire_intersect <- as.integer(fia_with_region$wildfire_intersect)

# now we will join to ca_grm_travel
  # First create a plot ID column to match ca_grm_travel$pltID. This is UNITCD_STATECD_COUNTYCD_PLOT.
    fia_with_region$pltID <- paste(fia_with_region$UNITCD, fia_with_region$STATECD, fia_with_region$COUNTYCD, fia_with_region$PLOT, sep = "_")

  # Then extract just the plot ID, resource area, and fire_intersect column from the sf object
  wildfire_flag_df <- fia_with_region %>%
  sf::st_drop_geometry() %>%
  dplyr::select(pltID, wildfire_intersect, ResourceAr) 

  # make pltID a character
  wildfire_flag_df$pltID <- as.character(wildfire_flag_df$pltID)
  
  # make sure each plot appears only once, in case a plot experiences multiple fires
  wildfire_flag_unique <- wildfire_flag_df %>%
  group_by(pltID) %>%
  summarise(
    wildfire_intersect = max(wildfire_intersect, na.rm = TRUE),
    ResourceAr = first(ResourceAr),   # keep the resource region
    .groups = "drop"
  )
  
  
  # now join to ca_grm
  ca_grm_travel_wildfire <- ca_grm_travel %>%
  dplyr::left_join(wildfire_flag_unique, by = "pltID")
  
  # make fire_intersect a factor
  ca_grm_travel_wildfire$wildfire_intersect <- as.factor(ca_grm_travel_wildfire$wildfire_intersect)
  
  # make resource area a factor
  ca_grm_travel_wildfire$ResourceAr <- as.factor(ca_grm_travel_wildfire$ResourceAr)  
```

2.  Fire Severity

```{r}

# -----------------------------
# 1. Load FIA data (many rows per plot)
# -----------------------------
fia_big <- fia_sf

# Creat pltID
  # First create a plot ID column to match ca_grm_travel$pltID. This is UNITCD_STATECD_COUNTYCD_PLOT.
    fia_big$pltID <- paste(fia_big$UNITCD, fia_big$STATECD, fia_big$COUNTYCD, fia_big$PLOT, sep = "_")

  # make pltID a character
  fia_big$pltID <- as.character(fia_big$pltID)

# Identify the unique plot geometry (one centroid per pltID)
fia_plots <- fia_big %>%
  group_by(pltID) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  st_as_sf()

# -----------------------------
# 2. Reproject FIA geometries to MTBS CRS
# -----------------------------
# read 1 MTBS file to get CRS
first_mtbs <- rast(list.files("/Users/alyssajain/FIA/CA_mtbs_unzipped",
                              full.names = TRUE,
                              pattern = "\\.tif$")[1])

fia_plots <- st_transform(fia_plots, crs(first_mtbs))

# -----------------------------
# 3. Create severity flag table per plot
# -----------------------------
fia_sev <- fia_plots %>%
  st_drop_geometry() %>%
  mutate(sev1 = 0, sev2 = 0, sev3 = 0, sev4 = 0, sev5 = 0)

# -----------------------------
# 4. Loop over all MTBS rasters
# -----------------------------
sev_files <- list.files("/Users/alyssajain/FIA/CA_mtbs_unzipped",
                        full.names = TRUE, pattern = "\\.tif$")

for (f in sev_files) {
  
  r <- rast(f)
  
  # Extract severity for each UNIQUE plot
  vals <- terra::extract(r, vect(fia_plots))[,2]   # col2 = severity
  vals[is.na(vals)] <- 0 # plots that are not in the mosaic should be 0, not NA
  
  # Update flags (ignore 0 and 6)
  fia_sev$sev1 <- fia_sev$sev1 | (vals == 1)
  fia_sev$sev2 <- fia_sev$sev2 | (vals == 2)
  fia_sev$sev3 <- fia_sev$sev3 | (vals == 3)
  fia_sev$sev4 <- fia_sev$sev4 | (vals == 4)
  fia_sev$sev5 <- fia_sev$sev5 | (vals == 5)
}

# Convert logical → integer
fia_sev <- fia_sev %>% mutate(across(starts_with("sev"), as.integer))

# -----------------------------
# 5. Join back into full FIA dataset
# -----------------------------
fia_with_severity <- fia_big %>%
  left_join(fia_sev, by = "pltID")

# join with ca_grm
# now join to ca_grm
ca_grm_travel_wildfire_sev <- ca_grm_travel_wildfire %>%
  dplyr::left_join(fia_sev, by = "pltID")

# change this later, just so rest of code runs with ca_grm_travel_wildfire
ca_grm_travel_wildfire <- ca_grm_travel_wildfire_sev

```

Resource Regions Color Palette

```{r}
# resource regions

# set colors
region_cols <- c(
  "North Coast California" = "#440154",
  "Central Coast California" = "#414487",
  "Northern Interior California" = "#2A788E",
  "Sacramento California" = "#22A884",
  "San Joaquin California" = "#7AD151",
  "Southern California California" = "#FDE725"
)
# plot resource regions
plot(resource_reg, col = region_cols[resource_reg$ResourceAr])
```

## **Findings:**

### *Q1)* We ran a logistic regression to determine how distance from the nearest mill affects the probability of fire, using CA resource region, forest type, and year as a random effect.

```{r}
  # Merchantable
   
  # fixed effects for resource area, year and fortypcd
    q1M1 <- glm(wildfire_intersect ~ OWH_Merchantable + ResourceAr + FORTYPCD + YEAR,
                      data = ca_grm_travel_wildfire, family = binomial)
    summary(q1M1)
    
    
    # Extract coefficients (log-odds)
    fixef_vals_q1M1 <- fixef(q1M1)
    
    # Tidy summary with SE, z, p-values
    fixed_summary_q1M1 <- broom.mixed::tidy(q1M1, effects = "fixed")
    
    # Add odds ratios for interpretation
    fixed_summary_q1M1 <- fixed_summary_q1M1 %>%
      mutate(
        odds_ratio = exp(estimate),
        prob_change = plogis(estimate)
      )
    
    print(fixed_summary_q1M1)
    view(fixed_summary_q1M1)
    
    

        
  # Chips
    
    # fixed effects for resource area, year and fortypcd
    q1C1 <- glm(wildfire_intersect ~ OWH_Chips + ResourceAr + FORTYPCD + YEAR,
                      data = ca_grm_travel_wildfire, family = binomial)
    summary(q1C1)
    
    # Extract coefficients (log-odds)
    fixef_vals_q1C1 <- fixef(q1C1)
    
    # Tidy summary with SE, z, p-values
    fixed_summary_q1C1 <- broom.mixed::tidy(q1C1, effects = "fixed")
    
    # Add odds ratios for interpretation
    fixed_summary_q1C1 <- fixed_summary_q1C1 %>%
      mutate(
        odds_ratio = exp(estimate),
        prob_change = plogis(estimate)
      )
    
    print(fixed_summary_q1C1)
    view(fixed_summary_q1C1)
    
    
```

**Interpretation:** Using the MTBS data to determine fire occurrence, we find that for every additional hour in distance to a merchantable mill, the log odds of fire increase by 0.49 (p \< 2e-16). Similarly, for every additional hour in distance to a chips mill, the log odds of fire increase by 0.53 (p \< 2e-16). These results indicate that plots farther away from mills are significantly more likely to experience fire.

### *Q2)* To see if decreased management is the mechanism increasing probability of fire farther from mills, we ran a logistic regression to see how distance from mills affects probability of treatment (which mostly refers to cutting, as well as a little preparation for planting).

Prep Data

```{r}
#| echo: false
# clean data for TRTCD = 10 (cutting) or 20 (Site preparation - Clearing, slash burning, chopping, disking, bedding, or other practices clearly intended to prepare a site for either natural or artificial regeneration.)
  ca_grm_travel_trtcd_grp <- ca_grm_travel_wildfire %>%
  mutate(
    TRTCD_CUT_PREP = if_else(
      as.numeric(as.character(TRTCD1)) %in% 10:20 |
      as.numeric(as.character(TRTCD2)) %in% 10:20 |
      as.numeric(as.character(TRTCD3)) %in% 10:20,
      1, 0
    )
  )
```

Run logistic regression

```{r}
# Merchantable
  q2aM <- glm(TRTCD_CUT_PREP ~ OWH_Merchantable + ResourceAr + FORTYPCD + 
    YEAR, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2aM)
  
# check differences between regions
    # Slopes (trends) by region
slopes <- emtrends(
  mod_region,
  specs = "ResourceAr",
  var = "OWH_Merchantable"
)

# Pairwise comparisons vs Central Coast
contr <- contrast(
  slopes,
  method = "trt.vs.ctrl",
  ref = "Central Coast California",
  adjust = "tukey"
)

summary(contr)
  
  # Chips
 q2aC <- glm(TRTCD_CUT_PREP ~ OWH_Chips + ResourceAr + FORTYPCD + 
    YEAR, data = ca_grm_travel_trtcd_grp, family = "binomial")
  summary(q2aC)
```

**Interpretation:** The log odds of treatment decrease by 0.59 and 0.61 for merchantable and chip mills, respectively (p \<2e-16). 

Figures (just for chip mill since they're almost the same)

```{r}

# Ensure numeric/factor types
ca_grm_travel_trtcd_grp <- ca_grm_travel_trtcd_grp %>%
  mutate(
    OWH_Merchantable = as.numeric(as.character(OWH_Chips)),
    YEAR = factor(YEAR),
    FORTYPCD = factor(FORTYPCD),
    ResourceAr = factor(ResourceAr)
  )

#  Overall pooled model (no interactions)
mod_overall <- glm(
  TRTCD_CUT_PREP ~ OWH_Chips + ResourceAr + YEAR + FORTYPCD,
  data = ca_grm_travel_trtcd_grp,
  family = binomial
)

# Extract overall slope & CI
overall_slope <- broom::tidy(mod_overall, conf.int = TRUE) %>%
  filter(term == "OWH_Chips") %>%
  mutate(
    region = "Overall",
    OR = exp(estimate),
    OR_low = exp(conf.low),
    OR_high = exp(conf.high)
  ) %>%
  dplyr::select(region, estimate, std.error, OR, OR_low, OR_high)

# Region-specific slopes from interaction model
mod_region <- glm(
  TRTCD_CUT_PREP ~ OWH_Chips * ResourceAr + YEAR + FORTYPCD,
  data = ca_grm_travel_trtcd_grp,
  family = binomial
)

coefs <- coef(mod_region)
vcov_mat <- vcov(mod_region)
regions <- levels(ca_grm_travel_trtcd_grp$ResourceAr)

region_slopes <- lapply(regions, function(r) {
  int_name <- paste0("OWH_Chips:ResourceAr", r)
  
  if (int_name %in% names(coefs)) {
    slope <- coefs["OWH_Chips"] + coefs[int_name]
    se <- sqrt(
      vcov_mat["OWH_Chips","OWH_Chips"] +
      vcov_mat[int_name,int_name] +
      2*vcov_mat["OWH_Chips", int_name]
    )
  } else {
    slope <- coefs["OWH_Chips"]  # reference region
    se <- sqrt(vcov_mat["OWH_Chips","OWH_Chips"])
  }
  
  data.frame(
    region = r,
    estimate = slope,
    std.error = se,
    OR = exp(slope),
    OR_low = exp(slope - 1.96*se),
    OR_high = exp(slope + 1.96*se)
  )
}) %>% bind_rows()

# Put Overall first
all_slopes <- bind_rows(overall_slope, region_slopes)

# order x-axis
all_slopes$region <- factor(
  all_slopes$region,
  levels = c("Overall", regions)  # Overall first, then regions
)

# Palette
region_cols <- c(
  "Overall" = "#000000",  # black
  "North Coast California" = "#440154",
  "Central Coast California" = "#414487",
  "Northern Interior California" = "#2A788E",
  "Sacramento California" = "#22A884",
  "San Joaquin California" = "#7AD151",
  "Southern California California" = "#FDE725"
)

# plot
ggplot(all_slopes, aes(x = region, y = OR, ymin = OR_low, ymax = OR_high, color = region)) +
  geom_pointrange(size = 2.5, fatten = 2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = region_cols) +
  scale_y_log10() +
  labs(
    x = NULL,
    y = "Odds ratio (log scale)",
    title = "Effect of travel time to chip mills on thinning",
    subtitle = "Overall effect and region-specific slopes (adjusted for YEAR + FORTYPCD)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    axis.ticks = element_line(linewidth = 1.2),
    axis.ticks.length = unit(0.3, "cm"),
    axis.line = element_line(linewidth = 1, color = "black"),

    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5), # center
    plot.subtitle = element_text(hjust = 0.5) # center
  )
```

Make orient plots for presentation

```{r}
# first compute y limits and tick breaks so they're the same for all plots
y_limits <- range(
  c(all_slopes$OR_low, all_slopes$OR_high),
  na.rm = TRUE
)

y_limits <- y_limits[y_limits > 0] # log scale so must be greater than 0


# blank
base_plot <- ggplot(all_slopes, aes(x = region, y = OR)) +
  geom_blank() +                 # forces axes, draws nothing
  scale_y_log10(
    limits = y_limits, # add limits
    breaks = c(0.3, 0.5, 1)) + # add breaks
  labs(
    x = NULL,
    y = "Odds ratio (log scale)",
    title = "Effect of travel time to chip mills on management",
    subtitle = "Overall effect and region-specific slopes"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16),
    axis.ticks = element_line(linewidth = 1.2),
    axis.ticks.length = unit(0.3, "cm"),
    axis.line = element_line(linewidth = 1.2, color = "black"),

    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

# with reference line at 1
orient_plot <- base_plot +
  geom_hline(
    yintercept = 1,
    linetype = "dashed",
    color = "grey40"
  )

# with data
final_plot <- orient_plot +
  geom_pointrange(
    aes(
      ymin = OR_low,
      ymax = OR_high,
      color = region
    ),
    size = 2.5,
    fatten = 2
  ) +
  scale_color_manual(values = region_cols)

# plot all
base_plot
orient_plot
final_plot
```

### Nonlinear effect of distance on fire?

**Fixed effects linear model with bins**

##### Faceted for both mill types

```{r}


# ------------------------------------------------
# Function to run binned FE model and extract coefs
# ------------------------------------------------

run_binned_model <- function(data, travel_var, facility_name){

  # 1 Copy dataset so we can modify safely
  df <- data

  # 2 Create bins from the selected travel-time variable
  df$bin <- cut(
    df[[travel_var]],
    breaks = seq(0,5,0.5),
    include.lowest = TRUE,
    labels = FALSE,
    right = FALSE
  )

  # 3 Run fixed effects regression
  model <- felm(
    wildfire_intersect ~ as.factor(bin) | ResourceAr + YEAR + FORTYPCD,
    data = df
  )
  
  # 4 print model output so you see the coefficients
  cat("\n\n============================\n")
  cat("Model for:", facility_name, "\n")
  cat("============================\n")
  print(summary(model))

  # 5 Extract coefficients
  coefs <- tidy(model, conf.int = TRUE) %>%
    filter(grepl("as.factor\\(bin\\)", term)) %>%
    mutate(
      bin = as.numeric(gsub("as.factor\\(bin\\)", "", term)),
      hours_mid = (bin - 1)*0.5 + 0.25,
      facility = facility_name
    )

  return(coefs)
}

# ------------------------------------------------
# Run models
# ------------------------------------------------
# copy data
ca_grm_travel_trtcd_grp_bins <- ca_grm_travel_trtcd_grp

merchant_coefs <- run_binned_model(
  ca_grm_travel_trtcd_grp,
  "OWH_Merchantable",
  "Merchantable mills"
)

chips_coefs <- run_binned_model(
  ca_grm_travel_trtcd_grp,
  "OWH_Chips",
  "Chip facilities"
)


plot_data <- bind_rows(merchant_coefs, chips_coefs)

# make facility a factor
plot_data$facility <- factor(
  plot_data$facility,
  levels = c("Merchantable mills", "Chip facilities")
)

# ------------------------------------------------
# Plot
# ------------------------------------------------

ggplot(plot_data, aes(hours_mid, estimate)) +

  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    fill = "#74add1",
    alpha = 0.25
  ) +

  geom_smooth(
    method = "loess",
    se = FALSE,
    span = 0.6,
    color = "#0570b0",
    linewidth = 1.3
  ) +

  facet_wrap(~facility, ncol = 2) +

  labs(
    x = "Travel time to facility (hours)",
    y = "Change in wildfire probability",
    title = "Travel Time to Processing Facilities and Wildfire Probability"
  ) +

  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )



```

###### Facet By Region

```{r}
# get rid of regions we don't want (low sample size)
ca_clean <- ca_grm_travel_trtcd_grp %>%
  filter(!ResourceAr %in% c(
    "Southern California California",
    "Central Coast California"
  ))

# get bins for chips and merch
ca_clean <- ca_clean %>%
  mutate(
    chips_bin = floor(OWH_Chips / 0.5),
    merch_bin = floor(OWH_Merchantable / 0.5)
  )

# convert to long format
travel_long <- ca_clean %>%
   dplyr::select(
    wildfire_intersect,
    YEAR,
    FORTYPCD,
    ResourceAr,
    chips_bin,
    merch_bin
  ) %>%
  tidyr::pivot_longer(
    cols = c(chips_bin, merch_bin),
    names_to = "mill_type",
    values_to = "bin"
  ) %>%
  mutate(
    mill_type = recode(
      mill_type,
      chips_bin = "Chips",
      merch_bin = "Merchantable"
    )
  )

# exclude groups with too few observations to avoid errors
travel_long_clean <- travel_long %>%
  filter(!is.na(bin), !is.na(wildfire_intersect)) %>%
  group_by(ResourceAr, mill_type) %>%
  filter(n() > 10) %>%   # minimum observations safeguard
  ungroup()

# fit models for both bin types and all regions
models <- travel_long_clean %>%
  group_split(ResourceAr, mill_type) %>%
  map(~ felm(
    wildfire_intersect ~ as.factor(bin) | YEAR + FORTYPCD,
    data = .x
  ))

model_keys <- travel_long_clean %>%
  distinct(ResourceAr, mill_type)

names(models) <- paste(model_keys$ResourceAr, model_keys$mill_type)

# extract coefficients 
coef_df <- map_dfr(names(models), function(name) {

  m <- models[[name]]

  parts <- strsplit(name, " ")[[1]]
  region <- paste(parts[-length(parts)], collapse = " ")
  mill <- parts[length(parts)]

  coefs <- tidy(m, conf.int = TRUE) %>%
    filter(grepl("as.factor\\(bin\\)", term)) %>%
    mutate(
      bin = as.numeric(gsub("as\\.factor\\(bin\\)", "", term)),
      hours_mid = bin * 0.5 + 0.25
    )

  # reference bin (intercept)
  intercept <- tidy(m) %>%
    filter(term == "(Intercept)") %>%
    pull(estimate)

  coefs %>%
    add_row(
      estimate = intercept,
      conf.low = intercept,
      conf.high = intercept,
      bin = 0,
      hours_mid = 0.25
    ) %>%
    mutate(
      ResourceAr = region,
      mill_type = mill
    )
})

# make colors lighter for chips
region_cols_chips <- colorspace::lighten(region_cols, 0.4)

coef_df <- coef_df %>%
  mutate(
    plot_col = ifelse(
      mill_type == "Chips",
      region_cols_chips[ResourceAr],
      region_cols[ResourceAr]
    )
  )

# plot
ggplot(
  coef_df,
  aes(
    x = hours_mid,
    y = estimate,
    color = plot_col,
    fill = plot_col,
    group = interaction(ResourceAr, mill_type),
    linetype = mill_type,
    alpha = mill_type
  )
) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    color = NA
  ) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    span = 0.6,
    linewidth = 1.2
  ) +
  facet_wrap(~ ResourceAr) +
  scale_color_identity() +
  scale_fill_identity() +
  scale_alpha_manual(values = c(
    Chips = 0.15,
    Merchantable = 0.25
  )) +
  labs(
    x = "Hours to Facility",
    y = "Wildfire Probability Coefficient",
    title = "Regional Effect of Travel Time on Wildfire Probability",
    linetype = "Facility Type",
    alpha = "Facility Type"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(3, "lines")
  )
```

make a table with coefficients

```{r}
coef_table <- purrr::imap_dfr(models, function(model, name) {

  broom::tidy(model) %>%
    mutate(model = name)

})
coef_table
```

##### Merchantable

separate plots into bins

```{r}

# get the range of travel times
summary(ca_grm_travel_trtcd_grp$OWH_Merchantable) # max: 4.9980

tmp <- ca_grm_travel_trtcd_grp %>%
  mutate(test = ifelse(
    OWH_Merchantable >= 0 & OWH_Merchantable < 0.5, 0, 
    ifelse(OWH_Merchantable >= 0.5 & OWH_Merchantable < 1, 1,
         ifelse(OWH_Merchantable >= 1 & OWH_Merchantable < 1.5, 2,
              ifelse(OWH_Merchantable >= 1.5 & OWH_Merchantable < 2, 3,
                     ifelse(OWH_Merchantable >= 2 & OWH_Merchantable < 2.5, 4,
                            ifelse(OWH_Merchantable >= 2.5 & OWH_Merchantable < 3, 5,
                                   ifelse(OWH_Merchantable >= 3 & OWH_Merchantable < 3.5, 6,
                                          ifelse(OWH_Merchantable >= 3.5 & OWH_Merchantable < 4, 7,
                                                 ifelse(OWH_Merchantable >= 4 & OWH_Merchantable < 4.5, 8,
                                                        ifelse(OWH_Merchantable >= 4.5 & OWH_Merchantable < 5, 9, 10)))))))))))
  
# filter by resource area
# model <- felm(formula = sev4 ~ as.factor(test)  |       YEAR + FORTYPCD,     data = filter(tmp, ResourceAr == "Southern California California"))

# unfiltered
model <- felm(formula = wildfire_intersect ~ as.factor(test)  |    YEAR + FORTYPCD,     data = tmp)

summary(model)
                       
```

plot felm relationship per bin

```{r}
# --- 1️⃣ Extract bin-level coefficients from felm model ---
coefs <- broom::tidy(model, conf.int = TRUE) %>%
  filter(grepl("as.factor\\(test\\)", term))

# --- 2️⃣ Clean up term names and assign bin midpoints ---
coefs <- coefs %>%
  mutate(
    bin = as.numeric(gsub("as\\.factor\\(test\\)", "", term)),
    hours_mid = bin * 0.5 + 0.25  # adjust if your bins differ
  )

# --- 3️⃣ Add baseline (reference bin = intercept) ---
intercept <- broom::tidy(model) %>%
  filter(term == "(Intercept)") %>%
  pull(estimate)

coefs <- coefs %>%
  add_row(
    term = "(Intercept)",
    estimate = intercept,
    conf.low = intercept,
    conf.high = intercept,
    bin = -1,
    hours_mid = 0
  )

# --- 4️⃣ Create Brookings-style plot ---
base_plot_nl <- ggplot(coefs, aes(x = hours_mid, y = estimate)) +
  
  # Axis and label formatting
  scale_x_continuous(
    name = "Hours to Merchantable Facility",
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  scale_y_continuous(
    name = "Wildfire Probability Coefficient",
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  
  # Title and subtitle (optional)
  labs(
    title = "Binned Effect of Travel Time on Wildfire Probability",
    subtitle = "fixed effects = resource area, forest type, and year"
  ) +
  
  # Clean, professional theme
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#08306b"),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray85", linetype = "dotted"),
    axis.text = element_text(color = "gray25", size = 12),
    plot.margin = margin(10, 15, 10, 15),
    axis.title = element_text(size = 14),
    axis.line = element_line(linewidth = 1, color = "black"),
    axis.ticks = element_line(linewidth = 1.2),
    axis.ticks.length = unit(0.3, "cm")
  )

final_plot_nl <- base_plot_nl + 
    # Confidence ribbon (subtle, transparent)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              fill = "#74add1", alpha = 0.25) +
  
  # Smooth nonlinear curve (primary visual focus)
  geom_smooth(
    method = "loess", se = FALSE, span = 0.6,
    color = "#0570b0", size = 1.3, linetype = "solid"
  ) 

base_plot_nl
final_plot_nl

```

Look for reasons for decline after 2.5 hours

```{r}
# Get # of plots in each bin
counts <- filter(
  tmp
  #, ResourceAr == "Southern California California"
  ) %>% count(test)
print(counts)

# get median OWH to merchantable
summary(as.numeric(ca_min_travel_time_clean$OWH_Merchantable))
```

Potential covariates of mill distance that could drive wildfire prob?

```{r}
# stocking density
   gsstk_lm <- lm(GSSTKCD ~ OWH_Merchantable , data = ca_grm_travel_wildfire)
    summary(gsstk_lm)
      # increases with distance to mill (goes from overstocked to nonstocked)
    
  # check felm, excluding fixed effects
    gsstk_felm <- felm(formula = GSSTKCD ~ as.factor(test)  |      0,     data = tmp)
    summary(gsstk_felm)
    
  # felm with fixed effects
    gsstk_felm <- felm(formula = GSSTKCD ~ as.factor(test)  |      ResourceAr + YEAR + FORTYPCD,     data = tmp)
    summary(gsstk_felm)
    # coefficients change with fixed effects, so stocking density doesn't independently covary with mill distance

# moisture
    
    # make it numeric (even though it is not ordinal)
    #-------------------------------------------------
        phys_lm <- lm(as.numeric(PHYSCLCD) ~ OWH_Merchantable , data = ca_grm_travel_trtcd_grp)
        summary(phys_lm)
          # increases with distance to mill 
        
      # check felm, excluding fixed effects
        phys_felm <- felm(formula = as.numeric(PHYSCLCD) ~ as.factor(test)  |      0,     data = tmp)
        summary(phys_felm)
        
      # felm with fixed effects
        phys_felm <- felm(formula = as.numeric(PHYSCLCD) ~ as.factor(test)  |      ResourceAr + YEAR + FORTYPCD,     data = tmp)
        summary(phys_felm)
        
    # check chi sq
    #-------------------------------------------------
      # make table
        phys_table = table(tmp$PHYSCLCD, tmp$test)
        print(phys_table)
        
        print(chisq.test(phys_table))
        
    # multinom log reg
      multinom_logreg <- multinom(PHYSCLCD ~ OWH_Merchantable + ResourceAr + YEAR + FORTYPCD,                                    data = ca_grm_travel_trtcd_grp)
      
      # Summary
      summary(multinom_logreg)
      
      # Get p-values (since nnet::multinom doesn't print them by default)
      z <- summary(multinom_logreg)$coefficients / summary(multinom_logreg)$standard.errors
      p <- 2 * (1 - pnorm(abs(z)))
      p
        

# Basal area living trees
    balive_lm <- lm(BALIVE ~ OWH_Merchantable , data = ca_grm_travel_trtcd_grp)
    summary(balive_lm)
     
    
  # check felm, excluding fixed effects
    balive_felm <- felm(formula = BALIVE ~ as.factor(test)  |      0,     data = tmp)
    summary(balive_felm)
    
  # felm with fixed effects
    balive_felm <- felm(formula = BALIVE ~ as.factor(test)  |      ResourceAr + YEAR + FORTYPCD,     data = tmp)
    summary(balive_felm)
```

###### Facet by Region

prep data

```{r}
# get rid of socal
ca_nosocal_grm_travel_trtcd_grp <- ca_grm_travel_trtcd_grp %>%
  filter(ResourceAr != "Southern California California")

# get rid of central coast
# get rid of socal
ca_nosocal_grm_travel_trtcd_grp <- ca_nosocal_grm_travel_trtcd_grp %>%
  filter(ResourceAr != "Central Coast California")

# define regions
regions <- unique(ca_nosocal_grm_travel_trtcd_grp$ResourceAr)

# get bins
ca_grm_travel_trtcd_grp_binned <- ca_nosocal_grm_travel_trtcd_grp %>%
  mutate(
    merchantable_bin = floor(OWH_Merchantable / 0.5)
  )

# fit one binned model per region
models <- map(
  regions,
  ~ felm(
      wildfire_intersect ~ as.factor(merchantable_bin) | YEAR + FORTYPCD,
      data = filter(ca_grm_travel_trtcd_grp_binned, ResourceAr == .x)
    )
)

names(models) <- regions

# extract coefficients
coef_df <- map_dfr(names(models), function(rg) {

  m <- models[[rg]]

  # bin coefficients
  coefs <- tidy(m, conf.int = TRUE) %>%
    filter(grepl("as.factor\\(merchantable_bin\\)", term)) %>%
    mutate(
      bin = as.numeric(gsub("as\\.factor\\(merchantable_bin\\)", "", term)),
      hours_mid = bin * 0.5 + 0.25,
      ResourceAr = rg
    )

  # reference bin (intercept)
  intercept <- tidy(m) %>%
    filter(term == "(Intercept)") %>%
    pull(estimate)

  coefs %>%
    add_row(
      term = "(Intercept)",
      estimate = intercept,
      conf.low = intercept,
      conf.high = intercept,
      bin = 0,
      hours_mid = 0.25,
      ResourceAr = rg
    )
})
```

plot

```{r}
ggplot(
  coef_df,
  aes(
    x = hours_mid, 
    y = estimate,
    color = ResourceAr,
    fill = ResourceAr)
) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    alpha = 0.2,
    color = NA
  ) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    span = 0.6,
    linewidth = 1.2
  ) +
  facet_wrap(~ ResourceAr) +
  scale_color_manual(values = region_cols) +
  scale_fill_manual(values = region_cols) +
  labs(
    x = "Hours to Merchantable Facility",
    y = "Wildfire Probability Coefficient",
    title = "Regional Effect of Travel Time on Wildfire Probability"
  ) +
  theme_minimal() + 
  theme (
    panel.spacing = unit(3, "lines")
  )

```

##### Chips

put plot in bins

```{r}
# get the range of travel times
summary(ca_grm_travel_trtcd_grp$OWH_Chips) # max:  4.959  
summary(ca_grm_travel_trtcd_grp$OWH_Merchantable) # max: 4.9980

tmp <- ca_grm_travel_trtcd_grp %>%
  mutate(test = ifelse(
    OWH_Chips >= 0 & OWH_Chips < 0.5, 0, 
    ifelse(OWH_Chips >= 0.5 & OWH_Chips < 1, 1,
         ifelse(OWH_Chips >= 1 & OWH_Chips < 1.5, 2,
              ifelse(OWH_Chips >= 1.5 & OWH_Chips < 2, 3,
                     ifelse(OWH_Chips >= 2 & OWH_Chips < 2.5, 4,
                            ifelse(OWH_Chips >= 2.5 & OWH_Chips < 3, 5,
                                   ifelse(OWH_Chips >= 3 & OWH_Chips < 3.5, 6,
                                          ifelse(OWH_Chips >= 3.5 & OWH_Chips < 4, 7,
                                                 ifelse(OWH_Chips >= 4 & OWH_Chips < 4.5, 8,
                                                        ifelse(OWH_Chips >= 4.5 & OWH_Chips < 5, 9, 10)))))))))))
  

model <- felm(formula = wildfire_intersect ~ as.factor(test)  |      ResourceAr + YEAR + FORTYPCD,     data = tmp)

summary(model)

# get the range of travel times but wiith smaller bins
summary(ca_grm_travel_trtcd_grp$OWH_Chips) # max:  4.959  
summary(ca_grm_travel_trtcd_grp$OWH_Merchantable) # max: 4.9980

tmp1 <- ca_grm_travel_trtcd_grp %>%
  mutate(test = ifelse(
    OWH_Chips >= 0 & OWH_Chips < 0.25, 0,
    ifelse(OWH_Chips >= 0.25 & OWH_Chips < 0.5, 1,
      ifelse(OWH_Chips >= 0.5 & OWH_Chips < 0.75, 2,
        ifelse(OWH_Chips >= 0.75 & OWH_Chips < 1, 3,
          ifelse(OWH_Chips >= 1 & OWH_Chips < 1.25, 4,
            ifelse(OWH_Chips >= 1.25 & OWH_Chips < 1.5, 5,
              ifelse(OWH_Chips >= 1.5 & OWH_Chips < 1.75, 6,
                ifelse(OWH_Chips >= 1.75 & OWH_Chips < 2, 7,
                  ifelse(OWH_Chips >= 2 & OWH_Chips < 2.25, 8,
                    ifelse(OWH_Chips >= 2.25 & OWH_Chips < 2.5, 9,
                      ifelse(OWH_Chips >= 2.5 & OWH_Chips < 2.75, 10,
                        ifelse(OWH_Chips >= 2.75 & OWH_Chips < 3, 11,
                          ifelse(OWH_Chips >= 3 & OWH_Chips < 3.25, 12,
                            ifelse(OWH_Chips >= 3.25 & OWH_Chips < 3.5, 13,
                              ifelse(OWH_Chips >= 3.5 & OWH_Chips < 3.75, 14,
                                ifelse(OWH_Chips >= 3.75 & OWH_Chips < 4, 15,
                                  ifelse(OWH_Chips >= 4 & OWH_Chips < 4.25, 16,
                                    ifelse(OWH_Chips >= 4.25 & OWH_Chips < 4.5, 17,
                                      ifelse(OWH_Chips >= 4.5 & OWH_Chips < 4.75, 18,
                                        ifelse(OWH_Chips >= 4.75 & OWH_Chips < 5, 19, 20)
                                      )
                                    )
                                  )
                                )
                              )
                            )
                          )
                        )
                      )
                    )
                  )
                )
              )
            )
          )
        )
      )
    )
  ))

  

# model <- felm(formula = wildfire_intersect ~ as.factor(test)  |      ResourceAr + YEAR + FORTYPCD,     data = tmp1)

summary(model)
                       
```

plot felm relationship per bin

```{r}
# --- 1️⃣ Extract bin-level coefficients from felm model ---
coefs <- broom::tidy(model, conf.int = TRUE) %>%
  filter(grepl("as.factor\\(test\\)", term))

# --- 2️⃣ Clean up term names and assign bin midpoints ---
coefs <- coefs %>%
  mutate(
    bin = as.numeric(gsub("as\\.factor\\(test\\)", "", term)),
    hours_mid = bin * 0.5 + 0.25  # adjust if your bins differ
  )

# --- 3️⃣ Add baseline (reference bin = intercept) ---
intercept <- broom::tidy(model) %>%
  filter(term == "(Intercept)") %>%
  pull(estimate)

coefs <- coefs %>%
  add_row(
    term = "(Intercept)",
    estimate = intercept,
    conf.low = intercept,
    conf.high = intercept,
    bin = -1,
    hours_mid = 0
  )

# --- 4️⃣ Create Brookings-style plot ---
ggplot(coefs, aes(x = hours_mid, y = estimate)) +
  
  # Confidence ribbon (subtle, transparent)
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              fill = "#74add1", alpha = 0.25) +
  
  # Smooth nonlinear curve (primary visual focus)
  geom_smooth(
    method = "loess", se = FALSE, span = 0.6,
    color = "#0570b0", size = 1.3, linetype = "solid"
  ) +
  
  # Axis and label formatting
  scale_x_continuous(
    name = "Hours to Chip Facility",
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  scale_y_continuous(
    name = "Wildfire Probability Coefficient",
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  
  # Title and subtitle (optional)
  labs(
    title = "Binned Effect of Travel Time Wildfire Probability",
    subtitle = "fixed effects = resource area, forest type, and year"
  ) +
  
  # Clean, professional theme
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5, color = "#08306b"),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30"),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray85", linetype = "dotted"),
    axis.text = element_text(color = "gray25", size = 12),
    plot.margin = margin(10, 15, 10, 15),
    axis.title = element_text(size = 14),
    axis.line = element_line(linewidth = 1, color = "black"),
    axis.ticks = element_line(linewidth = 1.2),
    axis.ticks.length = unit(0.3, "cm")
  )

base_plot_nl
```

###### Facet by Region

prep data

```{r}
# get rid of socal
ca_nosocal_grm_travel_trtcd_grp <- ca_grm_travel_trtcd_grp %>%
  filter(ResourceAr != "Southern California California")

# get rid of central coast
# get rid of socal
ca_nosocal_grm_travel_trtcd_grp <- ca_nosocal_grm_travel_trtcd_grp %>%
  filter(ResourceAr != "Central Coast California")

# define regions
regions <- unique(ca_nosocal_grm_travel_trtcd_grp$ResourceAr)

# get bins
ca_grm_travel_trtcd_grp_binned <- ca_nosocal_grm_travel_trtcd_grp %>%
  mutate(
    chips_bin = floor(OWH_Chips / 0.5)
  )

# fit one binned model per region
models <- map(
  regions,
  ~ felm(
      wildfire_intersect ~ as.factor(chips_bin) | YEAR + FORTYPCD,
      data = filter(ca_grm_travel_trtcd_grp_binned, ResourceAr == .x)
    )
)

names(models) <- regions

# extract coefficients
coef_df <- map_dfr(names(models), function(rg) {

  m <- models[[rg]]

  # bin coefficients
  coefs <- tidy(m, conf.int = TRUE) %>%
    filter(grepl("as.factor\\(chips_bin\\)", term)) %>%
    mutate(
      bin = as.numeric(gsub("as\\.factor\\(chips_bin\\)", "", term)),
      hours_mid = bin * 0.5 + 0.25,
      ResourceAr = rg
    )

  # reference bin (intercept)
  intercept <- tidy(m) %>%
    filter(term == "(Intercept)") %>%
    pull(estimate)

  coefs %>%
    add_row(
      term = "(Intercept)",
      estimate = intercept,
      conf.low = intercept,
      conf.high = intercept,
      bin = 0,
      hours_mid = 0.25,
      ResourceAr = rg
    )
})

```

plot

```{r}
ggplot(
  coef_df,
  aes(x = hours_mid, y = estimate)
) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    fill = "steelblue",
    alpha = 0.2
  ) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    span = 0.6,
    color = "steelblue",
    linewidth = 1.2
  ) +
  facet_wrap(~ ResourceAr) +
  labs(
    x = "Hours to Chip Mill",
    y = "Wildfire Probability Coefficient",
    title = "Nonlinear Relationship Between Chip Mill Access and Wildfire "
  ) +
  theme_minimal()

```

## Severity

**Nonlinear Model for each severity class**

reorganize bins for merchantable mills

```{r}
# get the range of travel times
summary(ca_grm_travel_trtcd_grp$OWH_Merchantable) # max: 4.9980

tmp <- ca_grm_travel_trtcd_grp %>%
  mutate(test = ifelse(
    OWH_Merchantable >= 0 & OWH_Merchantable < 0.5, 0, 
    ifelse(OWH_Merchantable >= 0.5 & OWH_Merchantable < 1, 1,
         ifelse(OWH_Merchantable >= 1 & OWH_Merchantable < 1.5, 2,
              ifelse(OWH_Merchantable >= 1.5 & OWH_Merchantable < 2, 3,
                     ifelse(OWH_Merchantable >= 2 & OWH_Merchantable < 2.5, 4,
                            ifelse(OWH_Merchantable >= 2.5 & OWH_Merchantable < 3, 5,
                                   ifelse(OWH_Merchantable >= 3 & OWH_Merchantable < 3.5, 6,
                                          ifelse(OWH_Merchantable >= 3.5 & OWH_Merchantable < 4, 7,
                                                 ifelse(OWH_Merchantable >= 4 & OWH_Merchantable < 4.5, 8,
                                                        ifelse(OWH_Merchantable >= 4.5 & OWH_Merchantable < 5, 9, 10)))))))))))

# tmp <- ca_grm_travel_trtcd_grp %>%
#   mutate(test = ifelse(
#     OWH_Chips >= 0 & OWH_Chips < 0.5, 0, 
#     ifelse(OWH_Chips >= 0.5 & OWH_Chips < 1, 1,
#          ifelse(OWH_Chips >= 1 & OWH_Chips < 1.5, 2,
#               ifelse(OWH_Chips >= 1.5 & OWH_Chips < 2, 3,
#                      ifelse(OWH_Chips >= 2 & OWH_Chips < 2.5, 4,
#                             ifelse(OWH_Chips >= 2.5 & OWH_Chips < 3, 5,
#                                    ifelse(OWH_Chips >= 3 & OWH_Chips < 3.5, 6,
#                                           ifelse(OWH_Chips >= 3.5 & OWH_Chips < 4, 7,
#                                                  ifelse(OWH_Chips >= 4 & OWH_Chips < 4.5, 8,
#                                                         ifelse(OWH_Chips >= 4.5 & OWH_Chips < 5, 9, 10)))))))))))
#   

# filter by resource area
# model <- felm(formula = sev4 ~ as.factor(test)  |       YEAR + FORTYPCD,     data = filter(tmp, ResourceAr == "Southern California California"))

# unfiltered
model <- felm(formula = wildfire_intersect ~ as.factor(test)  |    YEAR + FORTYPCD,     data = tmp)

summary(model)
```

prep for plot

```{r}
# model all sev classes
model_sev1 <- felm(formula = sev1 ~ as.factor(test)  |      ResourceAr + YEAR + FORTYPCD,     data = tmp)
summary(model_sev1)

model_sev2 <- felm(formula = sev2 ~ as.factor(test)  |      ResourceAr + YEAR + FORTYPCD,     data = tmp)
summary(model_sev2)

model_sev3 <- felm(formula = sev3 ~ as.factor(test)  |      ResourceAr + YEAR + FORTYPCD,     data = tmp)
summary(model_sev3)

model_sev4 <- felm(formula = sev4 ~ as.factor(test)  |      ResourceAr + YEAR + FORTYPCD,     data = tmp)
summary(model_sev4)

# combine
models <- list(
  sev1 = model_sev1,
  sev2 = model_sev2,
  sev3 = model_sev3,
  sev4 = model_sev4
)

```

plot

```{r}

# extract and clean coefficients
coef_df <- map_dfr(names(models), function(sev) {

  m <- models[[sev]]

  # bin-level coefficients
  coefs <- tidy(m, conf.int = TRUE) %>%
    filter(grepl("as.factor\\(test\\)", term)) %>%
    mutate(
      bin = as.numeric(gsub("as\\.factor\\(test\\)", "", term)),
      hours_mid = bin * 0.5 + 0.25,   # adjust if needed
      severity = sev
    )

  # intercept = reference bin
  intercept <- tidy(m) %>%
    filter(term == "(Intercept)") %>%
    pull(estimate)

  coefs %>%
    add_row(
      term = "(Intercept)",
      estimate = intercept,
      conf.low = intercept,
      conf.high = intercept,
      bin = -1,
      hours_mid = 0,
      severity = sev
    )
})

# filter by 2 and 4 
coef_df <- coef_df %>%
  filter(severity %in% c("sev2", "sev4"))

# plot
ggplot(
  coef_df,
  aes(
    x = hours_mid,
    y = estimate,
    color = severity,
    fill = severity,
    group = severity
  )
) +

  # confidence ribbons (subtle)
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    alpha = 0.18,
    color = NA
  ) +

  # nonlinear smoothed curves
  geom_smooth(
    method = "loess",
    se = FALSE,
    span = 0.6,
    linewidth = 1.2
  ) +

  scale_x_continuous(
    name = "Hours to Chip Mill",
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  scale_y_continuous(
    name = "Wildfire Probability Coefficient",
    expand = expansion(mult = c(0.05, 0.05))
  ) +
  
  scale_color_discrete(
    labels = c("sev2" = "Low Severity",
               "sev4" = "High Severity")
  ) +
  scale_fill_discrete(
    labels = c("sev2" = "Low Severity",
               "sev4" = "High Severity")
  ) +

  labs(
    title = "Nonlinear Effect of Mill Access on Wildfire Probability",
    subtitle = "Binned fixed-effects estimates by severity class",
    color = "Severity",
    fill = "Severity"
  ) +

  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "gray85", linetype = "dotted"),
    axis.text = element_text(color = "gray25")
  )

```

**Linear Model for each severity class**

Overall Merchantable

```{r}

# Overall
#4
model_4 <- felm(formula = sev4 ~ OWH_Merchantable  |       YEAR + FORTYPCD,     data = filter(ca_grm_travel_trtcd_grp))

summary(model_4)

#3
model_3 <- felm(formula = sev3 ~ OWH_Merchantable  |       YEAR + FORTYPCD,     data = filter(ca_grm_travel_trtcd_grp))

summary(model_3)

# 2
model_2 <- felm(formula = sev2 ~ OWH_Merchantable  |       YEAR + FORTYPCD,     data = filter(ca_grm_travel_trtcd_grp))

summary(model_2)

#1
model_1 <- felm(formula = sev1 ~ OWH_Merchantable  |       YEAR + FORTYPCD,     data = filter(ca_grm_travel_trtcd_grp))

summary(model_1)

models <- list(
  "sev1" = model_1,
  "sev2" = model_2,
  "sev3" = model_3,
  "sev4" = model_4
)

####### linear plot #######
coef_df <- bind_rows(
  lapply(names(models), function(nm) {
    tidy(models[[nm]], conf.int = TRUE) %>%
      filter(term == "OWH_Merchantable") %>%
      mutate(severity = nm)
  })
)

ggplot(coef_df, aes(x = severity, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1) +
  scale_x_discrete(
    labels = c(
      sev1 = "Unburned-Low Severity",
      sev2 = "Low Severity",
      sev3 = "Moderate Severity",
      sev4 = "High Severity"
    )
  ) +
  labs(
    x = "Fire Severity Class",
    y = "Δ Probability (per 1-hour increase in travel time)",
    title = "Association Between Travel Time to Merchantable Mill and Wildfire"
  ) +
  theme_minimal()  +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 14,
      face = "bold"
    )
  )

####### predicted plot #######

# Construct prediction grid
newdat <- tibble(
  OWH_Merchantable = seq(
    min(ca_grm_travel_trtcd_grp$OWH_Merchantable, na.rm = TRUE),
    max(ca_grm_travel_trtcd_grp$OWH_Merchantable, na.rm = TRUE),
    length.out = 200
  )
)

# Prediction function for felm models without intercept
predict_felm_partial <- function(model, newdat, name) {
  b1 <- coef(model)[["OWH_Merchantable"]]

  tibble(
    OWH_Merchantable = newdat$OWH_Merchantable,
    pred = b1 * newdat$OWH_Merchantable,
    severity = name
  )
}

# Build combined prediction dataframe
pred_df <- bind_rows(
  predict_felm_partial(model_1, newdat, "sev1"),
  predict_felm_partial(model_2, newdat, "sev2"),
  predict_felm_partial(model_3, newdat, "sev3"),
  predict_felm_partial(model_4, newdat, "sev4")
)

# Plot partial predictions
ggplot(pred_df, aes(x = OWH_Merchantable, y = pred, color = severity)) +
  geom_line(linewidth = 1) +
  labs(
    x = "OWH_Merchantable",
    y = "Partial Predicted Severity (within FE)",
    title = "Partial Effect of OWH_Merchantable Across Severity Models",
    color = "Model"
  ) +
  theme_minimal()
```

make point plots

```{r}
# step by step plot for presentation

# get limits and breaks
y_limits <- range(
  c(coef_df$conf.low, coef_df$conf.high),
  na.rm = TRUE
)

y_breaks <- pretty(y_limits, n = 5)

# base plot
base_plot_sev <- ggplot(coef_df, aes(x = severity, y = estimate)) +
  geom_blank() +

  # LOCKED Y SCALE
  scale_y_continuous(
    limits = y_limits,
    breaks = y_breaks
  ) +

  scale_x_discrete(
    labels = c(
      sev1 = "Unburned–Low Severity",
      sev2 = "Low Severity",
      sev3 = "Moderate Severity",
      sev4 = "High Severity"
    )
  ) +

  labs(
    x = "Fire Severity Class",
    y = "Δ Probability (per 1-hour increase in travel time)",
    title = "Effect of Travel Time on Wildfire Probability, by Severity"
  ) +

  theme_minimal(base_size = 13) +
  theme(
    # AXIS TEXT
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),

    # AXIS LINES (thicker)
    axis.line = element_line(linewidth = 1, color = "black"),

    # TICKS (thicker + longer)
    axis.ticks = element_line(linewidth = 1.2),
    axis.ticks.length = unit(0.3, "cm"),

    # TITLE
    plot.title = element_text(
      hjust = 0.5,
      size = 16,
      face = "bold"),
    
    # MARGINS
    axis.title.x = element_text(
      size = 14,
      margin = margin(t = 12)
    ),
    axis.title.y = element_text(
      size = 14,
      margin = margin(r = 12)
    )
  )

# final plot
final_plot_sev <- base_plot_sev +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.1,
  )


# plot
base_plot_sev
final_plot_sev
```

Overall Chips

```{r}
# Overall
#4
model_4 <- felm(formula = sev4 ~ OWH_Chips  |       YEAR + FORTYPCD,     data = filter(ca_grm_travel_trtcd_grp))

summary(model_4)

#3
model_3 <- felm(formula = sev3 ~ OWH_Chips  |       YEAR + FORTYPCD,     data = filter(ca_grm_travel_trtcd_grp))

summary(model_3)

# 2
model_2 <- felm(formula = sev2 ~ OWH_Chips  |       YEAR + FORTYPCD,     data = filter(ca_grm_travel_trtcd_grp))

summary(model_2)

#1
model_1 <- felm(formula = sev1 ~ OWH_Chips  |       YEAR + FORTYPCD,     data = filter(ca_grm_travel_trtcd_grp))

summary(model_1)

models <- list(
  "sev1" = model_1,
  "sev2" = model_2,
  "sev3" = model_3,
  "sev4" = model_4
)

####### linear plot #######
coef_df <- bind_rows(
  lapply(names(models), function(nm) {
    tidy(models[[nm]], conf.int = TRUE) %>%
      filter(term == "OWH_Chips") %>%
      mutate(severity = nm)
  })
)

ggplot(coef_df, aes(x = severity, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1) +
   scale_x_discrete(
    labels = c(
      sev1 = "Unburned-Low Severity",
      sev2 = "Low Severity",
      sev3 = "Moderate Severity",
      sev4 = "High Severity"
    )
  ) +
  labs(
    x = "Fire Severity Class",
    y = "Δ Probability (per 1-hour increase in travel time)",
    title = "Association Between Travel Time to Merchantable Mill and Wildfire"
  ) +
  theme_minimal()  +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      size = 14,
      face = "bold"
    )
  )

####### predicted plot #######

# Construct prediction grid
newdat <- tibble(
  OWH_Chips = seq(
    min(ca_grm_travel_trtcd_grp$OWH_Chips, na.rm = TRUE),
    max(ca_grm_travel_trtcd_grp$OWH_Chips, na.rm = TRUE),
    length.out = 200
  )
)

# Prediction function for felm models without intercept
predict_felm_partial <- function(model, newdat, name) {
  b1 <- coef(model)[["OWH_Chips"]]

  tibble(
    OWH_Chips = newdat$OWH_Chips,
    pred = b1 * newdat$OWH_Chips,
    severity = name
  )
}

# Build combined prediction dataframe
pred_df <- bind_rows(
  predict_felm_partial(model_1, newdat, "sev1"),
  predict_felm_partial(model_2, newdat, "sev2"),
  predict_felm_partial(model_3, newdat, "sev3"),
  predict_felm_partial(model_4, newdat, "sev4")
)

# Plot partial predictions
ggplot(pred_df, aes(x = OWH_Chips, y = pred, color = severity)) +
  geom_line(linewidth = 1) +
  labs(
    x = "OWH_Chips",
    y = "Partial Predicted Severity (within FE)",
    title = "Partial Effect of OWH_Chips Across Severity Models",
    color = "Model"
  ) +
  theme_minimal()
```
